### 2019-07-09 doc/speedstrength.tex
notes

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}

\title{Speed and strength of epidemic intervention}
\author{Jonathan Dushoff, Sang Woo Park, and?}

\usepackage{tabularx}

\usepackage{amsmath}
\usepackage{natbib}
\usepackage{hyperref}
\bibliographystyle{chicago}
\date{\today}

\usepackage{bm}

\usepackage{afterpage}
\usepackage{pdflscape}

\newcommand{\etal}{\textit{et al.}}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{BMB}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\begin{document}

\maketitle

\section{Introduction}

An epidemic can be characterized by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, 
measures how \emph{fast} an epidemic grows initially. On the other hand, the 
strength of an epidemic, often characterized by the basic reproductive number $\Ro$,
measures how \emph{large} an epidemic will grow. Knowing the speed and strength of 
an epidemic allows us to make short- and long-term predictions.

Infectious disease research often focuses on estimating the basic reproductive
number -- defined as the average number of secondary cases caused by a primary 
case in a fully susceptible population -- because it allows us to make 
prediction about the final size of an epidemic and the difficulty of the 
associated epidemic intervention. In particuar, the required control effort 
(i.e., the proportion of the population that needs to be vaccinated in order to
prevent the disease from spreading) is often expressed as $1 - 1/\Ro$. 
\swp{Maybe a sentence about Ronald Ross?}

Like epidemic strength, epidemic speed measures how disease spreads in the population
and, therefore, can help us understand the effectiveness of a control strategy.

\section{Methods}

\subsection{Epidemic moodel}

Infection kernel, $K(\tau)$, describes the rate at which secondary infections are 
expected to be caused by an infected individual, on \emph{average} (see Park 2019
for distinction between individual- and population-level infection kernel). Infection
kernel can be decomposed into two parts: reproduction number, 
$$
\Ro = \int K(s) ds,
$$ 
and the generation-interval distribution, 
$$
g(\tau) = \frac{K(\tau)}{\Ro},
$$
which represents the expected time distribution of the secondary cases. The 
generation-interval distribution plays an important role in liking speed and
strength of an epidemic.

Renewal equation:
\begin{equation}
i(t) = S(t) \int K(s) i(t-s) ds
\end{equation}

Euler-Lotka:
\begin{equation}
1/\RR = \int g(s) \exp(-rs) ds
\end{equation}

\subsection{Control strategies}

\begin{itemize}
	\item $L(\tau)$ represents intervention over the course of infection
	\item Then, $\hat K(\tau) = K(\tau)/L(\tau)$
	\item Define intervention strength:
	$$
	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)},
	$$
	harmonic mean of $L(\tau)$ weighted by generation-interval distribution; outbreak controlled if $\theta > \mathcal R$ (not affected by $\RR$)
	\item Define intervention speed ($\phi = r - \hat r$):
	$$
	1 = \langle \frac{\exp(\phi \tau) }{L(\tau)} \rangle_{b(\tau)}
	$$
	sort of a mean associated with $L(\tau)$; outbreak controlled if $\phi > r$ (not affected by $r$).
\end{itemize}

\subsection{HIV example}

\begin{itemize}
	\item Can active testing and treatment stop the epidemic?
	\item Explain different scenarios? assumptions?
	\item Early vs late transmission (change proportion of early transmission)
	\item Compare required strength and speed of intervention required to control the disease
\end{itemize}

\section{Results}

\begin{itemize}
	\item Early transmission matters for condom
	\item Early transmission doesn't matter so much for target and treat
	\item Think about speed vs strength axis
\end{itemize}

\section{Discussion}

\begin{itemize}
	\item Talk about generation time and reinterpret Eaton and Hallet?
	\item Short generation time -> teating and treating will help less but lower epidemic strength
	\item High generation time -> high proportion prevented but higher R.
\end{itemize}

\begin{itemize}
	\item r and R have more in common than we think
	\item Why test and treat predictions are robust to assumptions about transmission
\end{itemize}

\end{document}

### 2019-07-09 doc/speedstrength.tex
stuff
update
tweak

@@
 	$$
-	1 = \langle \frac{\exp(\phi \tau) }{L(\tau)} \rangle_{b(\tau)}
+	1 = \left\langle \frac{\exp(\phi \tau) }{L(\tau)} \right\rangle_{b(\tau)}
 	$$
@@
 
-\subsection{HIV example}
+\section{Examples}
+
+\subsection{Test and treatment - HIV}
 
@@
 	\item Compare required strength and speed of intervention required to control the disease
-\end{itemize}
-
-\section{Results}
-
-\begin{itemize}
 	\item Early transmission matters for condom
@@
 \end{itemize}
+
+\subsection{Asymptomatic infection}
+
+\begin{itemize}
+	\item See ``Factors that make an infectious disease outbreak controllable'' by Fraser et al.
+	\item Our framework generalizes their work
+	\item Timing of target intervention (Fraser works on this stuff and did a clinical trial on this; unpublished)
+\end{itemize}
+
+\subsection{Maybe one more example}
 

### 2019-07-09 doc/speedstrength.tex
more

@@
 
-Renewal equation:
+Here, we use the renewal equation to model disease incidence $i(\tau)$:
 \begin{equation}
-i(t) = S(t) \int K(s) i(t-s) ds
+i(\tau) = S(\tau) \int K(s) i(\tau-s) ds,
 \end{equation}
-
-Euler-Lotka:
+where current incidence $i(\tau)$ is expressed as a product of the proportion of the population susceptible $S(\tau)$ and the current infectiousness of all previously infected individuals $\int K(s) i(\tau - s) ds$.
+In the beginning of an epidemic, the proportion of population susceptible remains roughly constant ($S(\tau) \approx 1$) and this model simplifies to the Von Foerster equation [CITE Fraser et al.];
+during this period, we expect the epidemic to grow exponentially.
+Substituting $i(t) = i(0) \exp(rt)$ to (?) allows us to link speed and strength of an epidemic via the generation-interval distribution:
 \begin{equation}
-1/\RR = \int g(s) \exp(-rs) ds
+1/\RR = \int g(s) \exp(-rs) ds.
 \end{equation}
@@
 \subsection{Control strategies}
+
+We assume that the effectiveness of intervention strategies $L(\tau)$ varies over the course of infection.
 

### 2019-07-09 doc/speedstrength.tex
minor tweaks

@@
 \end{equation}
-where current incidence $i(\tau)$ is expressed as a product of the proportion of the population susceptible $S(\tau)$ and the current infectiousness of all previously infected individuals $\int K(s) i(\tau - s) ds$.
-In the beginning of an epidemic, the proportion of population susceptible remains roughly constant ($S(\tau) \approx 1$) and this model simplifies to the Von Foerster equation [CITE Fraser et al.];
-during this period, we expect the epidemic to grow exponentially.
-Substituting $i(t) = i(0) \exp(rt)$ to (?) allows us to link speed and strength of an epidemic via the generation-interval distribution:
+where current incidence $i(\tau)$ is expressed as a product of the proportion 
+of the population susceptible $S(\tau)$ and the current infectiousness of all 
+previously infected individuals $\int K(s) i(\tau - s) ds$. In the beginning 
+of an epidemic, the proportion of population susceptible remains roughly 
+constant ($S(\tau) \approx 1$) and this model simplifies to the Von Foerster 
+equation [CITE Fraser et al.]; during this period, we expect the epidemic 
+to grow exponentially. Substituting $i(t) = i(0) \exp(rt)$ to (?) allows us 
+to link speed and strength of an epidemic via the generation-interval distribution:
 \begin{equation}
@@
 
-\subsection{Control strategies}
+\subsection{Speed and strength of epidemic intervention}
 
-We assume that the effectiveness of intervention strategies $L(\tau)$ varies over the course of infection.
+We assume that the intervention strategies $L(\tau)$ vary over the course 
+of infection such that the effective kernel can be written as 
+$\hat K(\tau) = K(\tau)/L(\tau)$. For example, constant
 

### 2019-07-17 doc/speedstrength.tex
In the middle of a first attempt at rewriting

--- 2019-07-09 doc/speedstrength.tex
@@
 \title{Speed and strength of epidemic intervention}
-\author{Jonathan Dushoff, Sang Woo Park, and?}
+\author{Jonathan Dushoff, Sang Woo Park}
 
@@
 
-\usepackage{amsmath}
+%% \usepackage{amsmath}
 \usepackage{natbib}
@@
 
-\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
+\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
+\newcommand{\figlab}[1]{\label{fig:#1}}
+\newcommand{\eqref}[1]{(\ref{eq:#1})}
+\newcommand{\eqlab}[1]{\label{eq:#1}}
 \begin{document}
@@
 An epidemic can be characterized by its \emph{speed} and \emph{strength}.
-The speed of an epidemic, often characterized by the exponential growth rate $r$, 
-measures how \emph{fast} an epidemic grows initially. On the other hand, the 
-strength of an epidemic, often characterized by the basic reproductive number $\Ro$,
-measures how \emph{large} an epidemic will grow. Knowing the speed and strength of 
-an epidemic allows us to make short- and long-term predictions.
+The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
+The strength of an epidemic, often characterized by the reproductive number $\RR$, measures how many new cases are caused by a typical \emph{individual} case.
+Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.
 
-Infectious disease research often focuses on estimating the basic reproductive
-number -- defined as the average number of secondary cases caused by a primary 
-case in a fully susceptible population -- because it allows us to make 
-prediction about the final size of an epidemic and the difficulty of the 
-associated epidemic intervention. In particuar, the required control effort 
-(i.e., the proportion of the population that needs to be vaccinated in order to
-prevent the disease from spreading) is often expressed as $1 - 1/\Ro$. 
-\swp{Maybe a sentence about Ronald Ross?}
+Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
+The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \cite{Ross}, and the idea of averaging by defining a `typical' case was formalized in the 19x0s \cite{Heesterbeek}.
+$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic.
 
-Like epidemic strength, epidemic speed measures how disease spreads in the population
-and, therefore, can help us understand the effectiveness of a control strategy.
+Like epidemic strength, epidemic speed measures how disease spreads in the population.
+Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's strength on the same scale as the reproductive number.
+We then show that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed.
+We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
 
@@
 
-\subsection{Epidemic moodel}
+\subsection{Epidemic model}
+
+We model disease incidence using the renewal-equation framework \cite{SeeChamp?}. 
+This is a simple, flexible framework that can cover a wide range model structures.
+In our model, disease incidence at time $t$ is given by:
+\begin{equation}
+i(t) = \int K(\tau, t) i(t-\tau) d\tau.
+\end{equation}
+Here, $K$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
+In general, $K$ will depend on population characteristics that may change through time -- notably, the proportion of the population susceptible.
+Since we are interested in invasion and control, we will generally neglect changes in $K$ through time, thus we will assume $K\equiv K(\tau)$. 
+This means we are neglecting changes in susceptible proportion through time.
+
+\subsection{Strength-based decomposition}
+
+Assuming that $K$ doesn't change with time, we write:
+\begin{equation}
+	K(\tau) = \RR g(\tau),
+	\eqlab{speed}
+\end{equation}
+where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \cite{Champ}.
+Since $g$ is a distribution, it integrates to 1, and $\RR$ is thus the integral of $K$.
+
+Imagine a control measure that proportionally reduces $K$, so that the post-intervention kernel $\hat K = K/\theta$.
+The factorization \eqref{speed} show that the reduction needed to prevent invasion (or to eliminate disease)  is exactly \RR.
+We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention is larger than the strength of spread.
+
+We generalize this idea by allowing an intervention strategy to reduce $K$ by different proportions over the course of an individual infection. We write the post-intervention kernel:
+\begin{equation}
+	\hat K(\tau) = K(\tau)/L(\tau), 
+\end{equation}
+where $L(\tau)$ is the average proportional reduction for an individual infected time $\tau$ ago.
+
+We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$.
+It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution.
+
+\begin{equation}
+	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)},
+\end{equation}
+
+\subsection{Speed-based decomposition}
+
+By analogy with the strength-based factorization \eqref{speed}, we can rewrite the Euler-Lotka equation \eqref{euler} as a speed-based factorization:
+
+\begin{equation}
+K(\tau) = b(\tau)\exp(r\tau)
+\end{equation}
+
+Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times when the disease spreads exponentially.
 
@@
 $$
-g(\tau) = \frac{K(\tau)}{\Ro},
 $$
 which represents the expected time distribution of the secondary cases. The 
-generation-interval distribution plays an important role in liking speed and
+generation-interval distribution plays an important role in linking speed and
 strength of an epidemic.
 
-Renewal equation:
-\begin{equation}
-i(t) = S(t) \int K(s) i(t-s) ds
-\end{equation}
-
-Euler-Lotka:
-\begin{equation}
-1/\RR = \int g(s) \exp(-rs) ds
-\end{equation}
-
-\subsection{Control strategies}
-
-\begin{itemize}
-	\item $L(\tau)$ represents intervention over the course of infection
-	\item Then, $\hat K(\tau) = K(\tau)/L(\tau)$
-	\item Define intervention strength:
-	$$
-	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)},
-	$$
-	harmonic mean of $L(\tau)$ weighted by generation-interval distribution; outbreak controlled if $\theta > \mathcal R$ (not affected by $\RR$)
-	\item Define intervention speed ($\phi = r - \hat r$):
+	Define intervention speed ($\phi = r - \hat r$):
 	$$
@@
 	sort of a mean associated with $L(\tau)$; outbreak controlled if $\phi > r$ (not affected by $r$).
-\end{itemize}
 

### 2019-07-19 doc/speedstrength.tex
Trying to build the pretty part of the narrative

@@
 
-Imagine a control measure that proportionally reduces $K$, so that the post-intervention kernel $\hat K = K/\theta$.
-The factorization \eqref{speed} show that the reduction needed to prevent invasion (or to eliminate disease)  is exactly \RR.
+Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles. We then have:
+\begin{equation}
+	\hat K(\tau) = (\RR/\theta) g(\tau).
+\end{equation}
+Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$.
 We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention is larger than the strength of spread.
@@
 We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$.
-It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution.
-
+It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
 \begin{equation}
-	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)},
+	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
 \end{equation}
+In this more general case, we have again that the disease cannot spread when $\theta \geq \RR$.
 
@@
 
-Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times when the disease spreads exponentially.
+Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times when the disease spreads exponentially \cite{Champ}.
 
-Infection kernel, $K(\tau)$, describes the rate at which secondary infections are 
-expected to be caused by an infected individual, on \emph{average} (see Park 2019
-for distinction between individual- and population-level infection kernel). Infection
-kernel can be decomposed into two parts: reproduction number, 
-$$
-\Ro = \int K(s) ds,
-$$ 
-and the generation-interval distribution, 
-$$
-$$
-which represents the expected time distribution of the secondary cases. The 
-generation-interval distribution plays an important role in linking speed and
-strength of an epidemic.
+Now imagine an intervention that reduces transmission at a constant hazard rate $h$ across the disease generation, for example by identifying and isolating infectious individuals.
+We then have:
+\begin{equation}
+	\hat K(\tau) = K(\tau)\exp(-h\tau) = b(\tau)\exp((r-h)\tau)
+\end{equation}
+Since $b$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $h=r$.
+We call $h$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is larger than the speed of spread.
 
-	Define intervention speed ($\phi = r - \hat r$):
-	$$
+We define the strength of a general intervention $L(\tau)$ \eqref{Ldef} to be $\phi = r - \hat r$. We can then show that $\phi$ is a (sort of) mean satisfying:
+\begin{equation}
 	1 = \left\langle \frac{\exp(\phi \tau) }{L(\tau)} \right\rangle_{b(\tau)}
-	$$
-	sort of a mean associated with $L(\tau)$; outbreak controlled if $\phi > r$ (not affected by $r$).
+\end{equation}
 

### 2019-07-19 doc/speedstrength.tex
Finished this cut at the decomposition section

@@
 
-Now imagine an intervention that reduces transmission at a constant hazard rate $h$ across the disease generation, for example by identifying and isolating infectious individuals.
+Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation, for example by identifying and isolating infectious individuals.
 We then have:
 \begin{equation}
-	\hat K(\tau) = K(\tau)\exp(-h\tau) = b(\tau)\exp((r-h)\tau)
+	\hat K(\tau) = K(\tau)\exp(-\phi\tau) = b(\tau)\exp((r-\phi)\tau)
 \end{equation}
-Since $b$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $h=r$.
-We call $h$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is larger than the speed of spread.
+Since $b$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\phi=r$.
+We call $\phi$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is faster than the speed of spread.
 
-We define the strength of a general intervention $L(\tau)$ \eqref{Ldef} to be $\phi = r - \hat r$. We can then show that $\phi$ is a (sort of) mean satisfying:
+We generalize this idea by allowing the hazard rate $h(\tau)$ at which $K$ is reduced to vary through time, thus:
 \begin{equation}
-	1 = \left\langle \frac{\exp(\phi \tau) }{L(\tau)} \right\rangle_{b(\tau)}
+	\hat K(\tau) = K(\tau) \exp(-\int_0^\tau h(\sigma) d\sigma)
 \end{equation}
+
+We define the speed of a general intervention $L(\tau)$ \eqref{Ldef} to be $\phi = r - \hat r$. We can then show that $\phi$ is a (sort of) mean satisfying:
+\begin{equation}
+	1 = \left\langle \frac{\exp(\phi \tau) }{\exp(\int_0^\tau h(\sigma) d\sigma)} \right\rangle_{b(\tau)}
+\end{equation}
+The speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.
 

### 2019-07-23 doc/speedstrength.tex
merge conflicts

--- 2019-07-17 doc/speedstrength.tex
@@
 \title{Speed and strength of epidemic intervention}
-\author{Jonathan Dushoff, Sang Woo Park}
+\author{Jonathan Dushoff, Sang Woo Park, and?}
 
@@
 
-%% \usepackage{amsmath}
+\usepackage{amsmath}
 \usepackage{natbib}
@@
 
-\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
-\newcommand{\figlab}[1]{\label{fig:#1}}
-\newcommand{\eqref}[1]{(\ref{eq:#1})}
-\newcommand{\eqlab}[1]{\label{eq:#1}}
+\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
 \begin{document}
@@
 An epidemic can be characterized by its \emph{speed} and \emph{strength}.
-The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
-The strength of an epidemic, often characterized by the reproductive number $\RR$, measures how many new cases are caused by a typical \emph{individual} case.
-Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.
+The speed of an epidemic, often characterized by the exponential growth rate $r$, 
+measures how \emph{fast} an epidemic grows initially. On the other hand, the 
+strength of an epidemic, often characterized by the basic reproductive number $\Ro$,
+measures how \emph{large} an epidemic will grow. Knowing the speed and strength of 
+an epidemic allows us to make short- and long-term predictions.
 
-Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
-The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \cite{Ross}, and the idea of averaging by defining a `typical' case was formalized in the 19x0s \cite{Heesterbeek}.
-$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic.
+Infectious disease research often focuses on estimating the basic reproductive
+number -- defined as the average number of secondary cases caused by a primary 
+case in a fully susceptible population -- because it allows us to make 
+prediction about the final size of an epidemic and the difficulty of the 
+associated epidemic intervention. In particuar, the required control effort 
+(i.e., the proportion of the population that needs to be vaccinated in order to
+prevent the disease from spreading) is often expressed as $1 - 1/\Ro$. 
+\swp{Maybe a sentence about Ronald Ross?}
 
-Like epidemic strength, epidemic speed measures how disease spreads in the population.
-Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's strength on the same scale as the reproductive number.
-We then show that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed.
-We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
+Like epidemic strength, epidemic speed measures how disease spreads in the population
+and, therefore, can help us understand the effectiveness of a control strategy.
 
@@
 
-\subsection{Epidemic model}
-
-We model disease incidence using the renewal-equation framework \cite{SeeChamp?}. 
-This is a simple, flexible framework that can cover a wide range model structures.
-In our model, disease incidence at time $t$ is given by:
-\begin{equation}
-i(t) = \int K(\tau, t) i(t-\tau) d\tau.
-\end{equation}
-Here, $K$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
-In general, $K$ will depend on population characteristics that may change through time -- notably, the proportion of the population susceptible.
-Since we are interested in invasion and control, we will generally neglect changes in $K$ through time, thus we will assume $K\equiv K(\tau)$. 
-This means we are neglecting changes in susceptible proportion through time.
-
-\subsection{Strength-based decomposition}
-
-Assuming that $K$ doesn't change with time, we write:
-\begin{equation}
-	K(\tau) = \RR g(\tau),
-	\eqlab{speed}
-\end{equation}
-where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \cite{Champ}.
-Since $g$ is a distribution, it integrates to 1, and $\RR$ is thus the integral of $K$.
-
-Imagine a control measure that proportionally reduces $K$, so that the post-intervention kernel $\hat K = K/\theta$.
-The factorization \eqref{speed} show that the reduction needed to prevent invasion (or to eliminate disease)  is exactly \RR.
-We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention is larger than the strength of spread.
-
-We generalize this idea by allowing an intervention strategy to reduce $K$ by different proportions over the course of an individual infection. We write the post-intervention kernel:
-\begin{equation}
-	\hat K(\tau) = K(\tau)/L(\tau), 
-\end{equation}
-where $L(\tau)$ is the average proportional reduction for an individual infected time $\tau$ ago.
-
-We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$.
-It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution.
-
-\begin{equation}
-	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)},
-\end{equation}
-
-\subsection{Speed-based decomposition}
-
-By analogy with the strength-based factorization \eqref{speed}, we can rewrite the Euler-Lotka equation \eqref{euler} as a speed-based factorization:
-
-\begin{equation}
-K(\tau) = b(\tau)\exp(r\tau)
-\end{equation}
-
-Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times when the disease spreads exponentially.
+\subsection{Epidemic moodel}
 
@@
 $$
+g(\tau) = \frac{K(\tau)}{\Ro},
 $$
 which represents the expected time distribution of the secondary cases. The 
-generation-interval distribution plays an important role in linking speed and
+generation-interval distribution plays an important role in liking speed and
 strength of an epidemic.
 
-	Define intervention speed ($\phi = r - \hat r$):
+Here, we use the renewal equation to model disease incidence $i(\tau)$:
+\begin{equation}
+i(\tau) = S(\tau) \int K(s) i(\tau-s) ds,
+\end{equation}
+where current incidence $i(\tau)$ is expressed as a product of the proportion 
+of the population susceptible $S(\tau)$ and the current infectiousness of all 
+previously infected individuals $\int K(s) i(\tau - s) ds$. In the beginning 
+of an epidemic, the proportion of population susceptible remains roughly 
+constant ($S(\tau) \approx 1$) and this model simplifies to the Von Foerster 
+equation [CITE Fraser et al.]; during this period, we expect the epidemic 
+to grow exponentially. Substituting $i(t) = i(0) \exp(rt)$ to (?) allows us 
+to link speed and strength of an epidemic via the generation-interval distribution:
+\begin{equation}
+1/\RR = \int g(s) \exp(-rs) ds.
+\end{equation}
+
+\subsection{Speed and strength of epidemic intervention}
+
+We assume that the intervention strategies $L(\tau)$ vary over the course 
+of infection such that the effective kernel can be written as 
+$\hat K(\tau) = K(\tau)/L(\tau)$. For example, constant $L(\tau)$
+
+\begin{itemize}
+	\item $L(\tau)$ represents intervention over the course of infection
+	\item Then, $\hat K(\tau) = K(\tau)/L(\tau)$
+	\item Define intervention strength:
+	$$
+	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)},
+	$$
+	harmonic mean of $L(\tau)$ weighted by generation-interval distribution; outbreak controlled if $\theta > \mathcal R$ (not affected by $\RR$)
+	\item Define intervention speed ($\phi = r - \hat r$):
 	$$
@@
 	sort of a mean associated with $L(\tau)$; outbreak controlled if $\phi > r$ (not affected by $r$).
+\end{itemize}
 

### 2019-07-23 doc/speedstrength.tex
no more merge conflicts?
tweaks

--- 2019-07-19 doc/speedstrength.tex
@@
 Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
-The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \cite{Ross}, and the idea of averaging by defining a `typical' case was formalized in the 19x0s \cite{Heesterbeek}.
-$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic.
+The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
+the idea of averaging by defining a `typical' case was formalized in the 1990s \citep{diekmann1990definition}.
+$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.
 
@@
 
-We model disease incidence using the renewal-equation framework \cite{SeeChamp?}. 
-This is a simple, flexible framework that can cover a wide range model structures.
+\swp{edited to make the notation slightly clearer}
+We model disease incidence using the renewal-equation framework \citep{champredon2015intrinsic}. 
+This is a simple, flexible framework that can cover a wide range model structures \citep{Champredon2018equivalence}.
 In our model, disease incidence at time $t$ is given by:
@@
 \end{equation}
-Here, $K$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
-In general, $K$ will depend on population characteristics that may change through time -- notably, the proportion of the population susceptible.
-Since we are interested in invasion and control, we will generally neglect changes in $K$ through time, thus we will assume $K\equiv K(\tau)$. 
-This means we are neglecting changes in susceptible proportion through time.
+Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
+In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible.
+Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$. 
+This means we are neglecting changes in susceptible proportion through time;
+under this assumption, the renewal equation is equivalent to the Von Foerster equations, outlined in \cite{fraser2004factors}.
 
@@
 
-Assuming that $K$ doesn't change with time, we write:
+Assuming that the infection kernel $K$ doesn't change with time, we write:
 \begin{equation}
 	K(\tau) = \RR g(\tau),
-	\eqlab{speed}
+	\eqlab{strength}
 \end{equation}
-where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \cite{Champ}.
-Since $g$ is a distribution, it integrates to 1, and $\RR$ is thus the integral of $K$.
+where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \citep{champredon2015intrinsic}.
+Since $g$ is a distribution, it integrates to 1, and the reproduction number $\RR$ is thus the integral of $K$.
 
@@
 where $L(\tau)$ is the average proportional reduction for an individual infected time $\tau$ ago.
+Then, we define the post-intervention reproductive number with the following integral:
+\begin{equation}
+	\Rhat = \int K(\tau) d\tau
+\end{equation}
+This framework generalizes the work of \cite{fraser2004factors} who made parameteric assumptions about the shape of $1/L(\tau)$.
 
@@
 
-By analogy with the strength-based factorization \eqref{speed}, we can rewrite the Euler-Lotka equation \eqref{euler} as a speed-based factorization:
-
+The Euler-Lokta equation allows us to calculate the initial exponential growth rate $r$ of an epidemic given an infection kernel $K$:
+\begin{equation}
+	1 = \int K(\tau) \exp(-r\tau) d\tau
+	\eqlab{euler}
+\end{equation}
+By analogy with the strength-based factorization \eqref{strength}, we can rewrite the Euler-Lotka equation \eqref{euler} as a speed-based factorization:
 \begin{equation}
@@
 \end{equation}
-
-Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times when the disease spreads exponentially \cite{Champ}.
+Like $g$, $b$ is a distribution: in this case the initial backward generation interval, which gives the distribution of realized generation times (measured from the infectee's point of view) when the disease spreads exponentially \citep{champredon2015intrinsic, britton2019estimation}.
 
@@
 \end{equation}
-Since $b$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\phi=r$.
+Since $b$ is a distribution (which integrates to 1), the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\phi=r$. 
 We call $\phi$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is faster than the speed of spread.
@@
 \begin{equation}
-	\hat K(\tau) = K(\tau) \exp(-\int_0^\tau h(\sigma) d\sigma)
+	\hat K(\tau) = K(\tau) \exp\left(-\int_0^\tau h(\sigma) d\sigma\right)
 \end{equation}
-
-We define the speed of a general intervention $L(\tau)$ \eqref{Ldef} to be $\phi = r - \hat r$. We can then show that $\phi$ is a (sort of) mean satisfying:
+The associated post-intervention epidemic speed $\hat r$ is given by:
 \begin{equation}
-	1 = \left\langle \frac{\exp(\phi \tau) }{\exp(\int_0^\tau h(\sigma) d\sigma)} \right\rangle_{b(\tau)}
+	1 = \int \hat K(\tau) \exp(-\hat r\tau) d\tau.	
+\end{equation}
+We define the speed of a general intervention to be $\phi = r - \hat r$. 
+We can then show that $\phi$ is a (sort of) mean satisfying:
+\begin{equation}
+	1 = \left\langle \frac{\exp(\phi \tau) }{\exp\left(\int_0^\tau h(\sigma) d\sigma\right)} \right\rangle_{b(\tau)}
 \end{equation}
@@
 
-\section{Examples}
+\section{Results}
+
+In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for HIV.
+In particular, we compare how the amount of early vs.
+late transmission affect each intervention strategies.
+
+In order to
 
@@
 
+\bibliography{speedstrength}
+
 \end{document}

### 2019-07-24 doc/speedstrength.tex
working

@@
 
-\section{Results}
+\section{Examples}
 
-In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for HIV.
-In particular, we compare how the amount of early vs.
-late transmission affect each intervention strategies.
+In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the Human immunodeficiency virus (HIV). 
+These examples are not meant to provide precise or accurate estimates of the effectiveness of intervention strategies; 
+instead, we use these examples to demonstrate how strength- and speed-based decompositions allow us to better evaluate each control strategy under different scenarios.
+In particular, we study how the amount of early HIV transmission can affect our decisions.
 
-In order to
+We model the infection kernel of the HIV as a sum of two gamma distributions:
+\begin{equation}
+K(\tau) = \mathcal R \left(\tsub{p}{early} \tsub{f}{early}(\tau) + (1-\tsub{p}{early}) \tsub{f}{late}(\tau) \right).
+\end{equation}
+The first gamma distribution, $\tsub{f}{early}(\tau)$, models early HIV transmission during the acute infection stage.
+We assume that $\tsub{f}{early}(\tau)$ has a mean of 3 months \citep{hollingsworth2008hiv} and a shape parameter of 3.
+The second gamma distribution, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
+We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \citep{babiker2000time} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
+Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.
+
+As our baseline scenario, we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
+Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$) from a HIV prevalence data in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
+Increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{../figure/figure1.pdf}
+\caption{This is not a figure.}
+\end{figure}
+
+\subsection{Univarsal testing and treatment}
 

### 2019-07-24 doc/speedstrength.tex
minor tweak

@@
 
-As our baseline scenario, we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
-Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$) from a HIV prevalence data in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
-Increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number.
+As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
+Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from a HIV prevalence data in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
+Further increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number (\figref{example}C).
 
@@
 \caption{This is not a figure.}
+\figlab{example}
 \end{figure}

### 2019-07-25 doc/speedstrength.tex
more work

@@
 \documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
 
@@
 
-\section{Examples}
+\section{Intervention strategies for HIV}
 
@@
 
-As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
-Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from a HIV prevalence data in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
-Further increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number (\figref{example}C).
-
-\begin{figure}
+\begin{figure}[!t]
 \includegraphics[width=\textwidth]{../figure/figure1.pdf}
@@
 
-\subsection{Univarsal testing and treatment}
+As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
+Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from a HIV prevalence data in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
+Further increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number (\figref{example}C);
+therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is more early transmission.
 
-\subsection{Test and treatment - HIV}
+\swp{Probably don't need a figure for this?}
+First, consider a condom intervention, which is approximately 80\% effective in reducing HIV seroconversion \citep{weller2002condom}.
+Assuming that ?\% of the sexually active population uses condom consistently,
+we can model this intervention as a constant function: $\tsub{L}{condom} = ???$.
+In this case, the strength of condom intervention is always constant regardless of the amount of early proportion.
+Therefore, we expect condom intervention to be able to control the spread of the HIV when more than ???\% of transmission is caused by early transmission (i.e., when the intervention strength is greater than the epidemic strength).
 
-\begin{itemize}
-	\item Can active testing and treatment stop the epidemic?
-	\item Explain different scenarios? assumptions?
-	\item Early vs late transmission (change proportion of early transmission)
-	\item Compare required strength and speed of intervention required to control the disease
-	\item Early transmission matters for condom
-	\item Early transmission doesn't matter so much for target and treat
-	\item Think about speed vs strength axis
-\end{itemize}
+\begin{figure}[!t]
+\includegraphics[width=\textwidth]{../figure/figure2.pdf}
+\caption{This is not a figure.}
+\figlab{test}
+\end{figure}
 
-\subsection{Asymptomatic infection}
+On the other hand, consider a test-and-treat strategy in which infected individuals are identified and put to clinical care (e.g., antiretroviral therapy).
+We can model this intervention $\tsub{L}{test}$ as a saturating function that begins at 1 and plateaus at a sufficiently high value (such that identified individuals are effectively no longer infectious);
+in particular, we assume that all individuals will be identified and put to treatment within the first 5 years since their infection (\figref{test}A).
+Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength; 
+however, it is unclear why this should be the case.
 
-\begin{itemize}
-	\item See ``Factors that make an infectious disease outbreak controllable'' by Fraser et al.
-	\item Our framework generalizes their work
-	\item Timing of target intervention (Fraser works on this stuff and did a clinical trial on this; unpublished)
-\end{itemize}
-
-\subsection{Maybe one more example}
+Using the speed-based decomposition allows us to understand the problem much more clearly.
+The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because individuals are actively getting tested and treated);
+the hazard rate goes down soon after the peak and reaches 0 after about 5 years because infected individuals who have access to medical care would have been identified within the first 5 years or so.
+The speed of epidemic stays constant regardless of the proportion of the early transmission because epidemic speed corrresponds to the observed growth rate.
+The speed of intervention stays roughly constant but slightly increases as proportion of early transmission increases \swp{because?}.
+This framework clearly demonstrates that test-and-treat intervention will be effective regardless of the proportion of early transmission.
 
 \section{Discussion}
-
-\begin{itemize}
-	\item Talk about generation time and reinterpret Eaton and Hallet?
-	\item Short generation time -> teating and treating will help less but lower epidemic strength
-	\item High generation time -> high proportion prevented but higher R.
-\end{itemize}
-
-\begin{itemize}
-	\item r and R have more in common than we think
-	\item Why test and treat predictions are robust to assumptions about transmission
-\end{itemize}
 

### 2019-07-25 figure/figure1.R
figures

@@ -1,0 +1,73 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+source("../R/data.R")
+source("../R/HIVfuns.R")
+source("../R/kernel.R")
+
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+g1 <- ggplot(h_base) +
+	geom_line(aes(time/12, density)) +
+	scale_x_continuous("Generation time (years)") +
+	scale_y_continuous("Density", expand=c(0, 0)) +
+	ggtitle("A") +
+	theme(
+		legend.position=c(0.9, 0.9),
+		legend.title=element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+coef(lfit)
+
+pp <- predict(lfit, se=TRUE)
+
+g2 <- ggplot(hiv_ts) +
+	geom_point(aes(year, prevalence)) +
+	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
+	scale_x_continuous("Year") +
+	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
+	ggtitle("B") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+r2R(h_base, coef(lfit)[2]/12)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <-  sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+g3 <- ggplot(earlydata) +
+	geom_line(aes(early, R0)) +
+	geom_point(aes(0.23, 3.17)) +
+	geom_segment(aes(x=0.23, y=2, xend=0.23, yend=3.17), lty=2) +
+	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
+	scale_x_continuous("Proportion of early transmission") +
+	scale_y_continuous("Reproduction number", expand=c(0, 0)) +
+	ggtitle("C") +
+	theme(
[…]

### 2019-07-25 figure/figure2.R

@@ -1,0 +1,107 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+source("../R/data.R")
+source("../R/HIVfuns.R")
+source("../R/kernel.R")
+
+maxRate <- 1/6
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+b_base <- b0fun(h_base, coef(lfit)[2]/12)
+
+tt <- testingFun(maxRate)
+
+g1 <- ggplot(tt) +
+	geom_line(aes(time/12, strength), col=2) +
+	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
+	scale_x_continuous("Time (years)") +
+	scale_y_continuous(expression(L[test])) +
+	ggtitle("A") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+interstrength <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	theFun(hh, tt)
+})
+
+strengthdata <- data.frame(
+	early=early,
+	strength=interstrength
+)
+
+g2 <- ggplot(earlydata) +
+	geom_line(aes(early, R0, col="Epidemic")) +
+	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
+	scale_color_manual(values=c("black", "red")) +
+	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
+	scale_y_continuous("Strength") +
+	ggtitle("B") +
+	theme(
+		legend.position=c(0.85, 0.9),
+		legend.title = element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+g3 <- ggplot(tt) +
+	geom_line(aes(time/12, hazRate), col=2) +
[…]

### 2019-07-25 R/HIVfuns.R
R files

@@ -1,0 +1,43 @@
+# Units are months
+# Units of g are 1/month, so Sum{step*g) should be 1
+
+testingFun <- function(maxRate
+					   , hMean=12, hShape=2, hheight=4
+					   , step = 0.1, window=200
+){
+	time <- seq(step, window, by=step)
+	tHaz0 <- dgamma(time, shape=hShape, scale=hMean/hShape)
+	tHaz <- hheight*tHaz0/max(tHaz0)
+	hazRate <- maxRate*(1-exp(-tHaz))
+	Haz <- step*cumsum(hazRate)
+	return(data.frame(time
+					  , hazRate
+					  , strength = exp(Haz)
+	))
+}
+
+flatFun <- function(L
+					, step = 0.1, window=200
+){
+	time <- seq(step, window, by=step)
+	hazRate <- 0*time
+	return(data.frame(time
+					  , hazRate
+					  , strength = hazRate+L
+	))
+}
+
+HIVgen <- function(earlyProp 
+				   , earlyMean=3, earlyShape=3
+				   , lateMean=120, lateShape=2
+				   , step = 0.1, window=200
+){
+	time <- seq(step, window, by=step)
+	d0 <- (
+		earlyProp*dgamma(time, shape=earlyShape, scale=earlyMean/earlyShape)
+		+ (1-earlyProp)*dgamma(time, shape=lateShape, scale=lateMean/lateShape)
+	)
+	return(data.frame(time
+					  , density=d0/sum(step*d0)
+	))
+}

### 2019-07-25 R/data.R

@@ -1,0 +1,8 @@
+## https://www.who.int/bulletin/volumes/91/1/12-106807/en/
+
+hiv_ts <- data.frame(
+	year=1990:2010,
+	prevalence=c(0.8, 1.4, 2.4, 4.3, 7.6, 10.4, 14.2, 17, 22.8,
+				 22.4, 24.5, 24.8, 26.5, 27.9, 29.5, 30.2, 29.1, 29.4,
+				 29.3, 29.4, 30.2)/100
+)

### 2019-07-25 R/kernel.R

@@ -1,0 +1,121 @@
+# g, b are data frames with time and density
+# L is a data frame with time, strength and something about hazard
+
+## Calculate the initial backward distribution
+b0fun <- function(g, r){
+	step <- with(g, time[[2]]-time[[1]])
+	d0 <- with(g, density*exp(-r*time))
+	return(with(g, data.frame(time
+							  , density=d0/sum(step*d0)
+	)))
+}
+
+# How much does a given discount (r) _reduce_ R for distribution g?
+r2R <- function(g, r){
+	step <- with(g, time[[2]]-time[[1]])
+	return(with(g, 1/(step*sum(density*exp(-r*time)))))
+}
+
+R2r <- function(g, R){
+	step <- with(g, time[[2]]-time[[1]])
+	return(uniroot(
+		function(r, g){r2R(g, r)-R}
+		, interval=c(0, 1/step)
+		, g=g
+	)$root)
+}
+
+# Calculate the arithmetic mean inside the implicit equation for phi
+phiBar <- function(b, L, phi){
+	step <- with(b, time[[2]]-time[[1]])
+	return(with(cbind(b,L), {
+		step*sum(density*exp(phi*time)/strength)
+	}))
+}
+
+phiFun <- function(b, L){
+	step <- with(b, time[[2]]-time[[1]])
+	return(uniroot(
+		function(phi, b, L){phiBar(b, L, phi)-1}
+		, interval=c(0, 1/step)
+		, b=b, L=L
+	)$root)
+}
+
+theFun <- function(g, L){
+	step <- with(g, time[[2]]-time[[1]])
+	return(with(cbind(g, L), {
+		1/(step*sum(density/strength))
+	}))
+}
+
+densPlot <- function(time, k, scen, color="black", ymax=NULL, title=NULL){
+	if(is.null(ymax)) ymax <- max(k)
+	with(c(scen), {
+		if(is.null(title)) title <- disease
+		if(is.numeric(title)) title <-
+				paste("Reproductive number", sprintf("%5.2f", title))
+		# if(is.numeric(title)) title <- expression(R=title)
+		plot(time, k
+			 , main = title
+			 , xlab = paste0("time (", unit, ")")
+			 , ylab = paste0("Density (1/", unit, ")")
+			 , type = "l", col=color
+			 , ylim = c(0, ymax)
+		)
+	})
+}
+
[…]

### 2019-07-26 doc/speedstrength.tex
figure captions

@@
 In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the Human immunodeficiency virus (HIV). 
-These examples are not meant to provide precise or accurate estimates of the effectiveness of intervention strategies; 
-instead, we use these examples to demonstrate how strength- and speed-based decompositions allow us to better evaluate each control strategy under different scenarios.
+These examples are meant to provide neither precise nor accurate estimates of the effectiveness of intervention strategies; 
+instead, we use these examples to demonstrate how strength- and speed-based decompositions can help us better evaluate each control strategy under different scenarios.
 In particular, we study how the amount of early HIV transmission can affect our decisions.
@@
 \includegraphics[width=\textwidth]{../figure/figure1.pdf}
-\caption{This is not a figure.}
+\caption{
+\textbf{The infection kernel of the HIV.}
+A. The infection kernel of the HIV is approximated using a sum of two gamma distributions. We assume that the baseline proportion of early transmission is 23\% \citep{hayes2006amplified}.
+B.
+Time series of HIV prevalence in pregnant women in South Africa, 1990 - 2010 \citep{barron2013eliminating}.
+The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
+C. Increasing the proportion of early transmission reduces the estimate of the reproduction number.
+}
 \figlab{example}
@@
 As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
-Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from a HIV prevalence data in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
+Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from a HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
 Further increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number (\figref{example}C);
@@
 
-\swp{Probably don't need a figure for this?}
+\begin{figure}[!t]
+\includegraphics[width=\textwidth]{../figure/figure2.pdf}
+\caption{
+\textbf{Understanding condom intervention using strength-based decomposition.}
+A. Condom intervention $\tsub{L}{condom}$ can be represeted as a constant function.
+B. The amount of early transmission affects the strength of an epidemic but not the strength of condom intervention. 
+}
+\figlab{condom}
+\end{figure}
+
 First, consider a condom intervention, which is approximately 80\% effective in reducing HIV seroconversion \citep{weller2002condom}.
 Assuming that ?\% of the sexually active population uses condom consistently,
-we can model this intervention as a constant function: $\tsub{L}{condom} = ???$.
-In this case, the strength of condom intervention is always constant regardless of the amount of early proportion.
+we can model this intervention as a constant function: $\tsub{L}{condom} = ???$ (\figref{condom}A).
+In this case, the strength of condom intervention is always constant regardless of the amount of early transmission, whereas the strength of epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
 Therefore, we expect condom intervention to be able to control the spread of the HIV when more than ???\% of transmission is caused by early transmission (i.e., when the intervention strength is greater than the epidemic strength).
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/figure2.pdf}
-\caption{This is not a figure.}
+\includegraphics[width=\textwidth]{../figure/figure3.pdf}
+\caption{
+\textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
+A. Test-and-treat intervention $\tsub{L}{test}$ starts at 1 and saturates after about 5 years.
+B. Increasing the amount of early transmission decreases the strength of an epidemic as well as the strength of test-and-treat intervention.
+C. Hazard rate $\tsub{h}{test}$ of test-and-treat intervention $\tsub{L}{test}$ peaks and plateaus very quickly and reaches zero after about 5 years.
+D. The amount of early transmission has no effect on the speed of an epidemic and small effect on the speed of intervention.
+We assume $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau h(\sigma) d\sigma \right)$ and $h(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $12$ months and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 1/6$.
+}
 \figlab{test}

### 2019-07-26 figure/figure2.R

@@ -6,20 +6,15 @@ source("../R/data.R")
 source("../R/HIVfuns.R")
 source("../R/kernel.R")
 
-maxRate <- 1/6
 h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+yintercept <- 4 ## arbitrary value
 
 lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
 
-b_base <- b0fun(h_base, coef(lfit)[2]/12)
-
-tt <- testingFun(maxRate)
-
-g1 <- ggplot(tt) +
-	geom_line(aes(time/12, strength), col=2) +
-	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
+g1 <- ggplot() +
+	geom_hline(yintercept=yintercept, col=2) +
 	scale_x_continuous("Time (years)") +
-	scale_y_continuous(expression(L[test])) +
+	scale_y_continuous(expression(L[condom])) +
 	ggtitle("A") +
 	theme(
 		panel.grid = element_blank(),
@@ -39,19 +34,9 @@ earlydata <- data.frame(
 	R0=R0
 )
 
-interstrength <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	theFun(hh, tt)
-})
-
-strengthdata <- data.frame(
-	early=early,
-	strength=interstrength
-)
-
 g2 <- ggplot(earlydata) +
 	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
+	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
 	scale_color_manual(values=c("black", "red")) +
 	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
 	scale_y_continuous("Strength") +
@@ -64,44 +49,6 @@ g2 <- ggplot(earlydata) +
 		axis.line = element_line()
 	)
 
-g3 <- ggplot(tt) +
-	geom_line(aes(time/12, hazRate), col=2) +
-	geom_hline(aes(yintercept=phiFun(b_base, tt)), lty=2, col=2) +
-	scale_x_continuous("Time (years)") +
-	scale_y_continuous(expression(h[test])) +
-	ggtitle("C") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
[…]

### 2019-07-26 figure/figure3.R

@@ -1,0 +1,107 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+source("../R/data.R")
+source("../R/HIVfuns.R")
+source("../R/kernel.R")
+
+maxRate <- 1/6
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+b_base <- b0fun(h_base, coef(lfit)[2]/12)
+
+tt <- testingFun(maxRate)
+
+g1 <- ggplot(tt) +
+	geom_line(aes(time/12, strength), col=2) +
+	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
+	scale_x_continuous("Time (years)") +
+	scale_y_continuous(expression(L[test])) +
+	ggtitle("A") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+interstrength <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	theFun(hh, tt)
+})
+
+strengthdata <- data.frame(
+	early=early,
+	strength=interstrength
+)
+
+g2 <- ggplot(earlydata) +
+	geom_line(aes(early, R0, col="Epidemic")) +
+	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
+	scale_color_manual(values=c("black", "red")) +
+	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
+	scale_y_continuous("Strength") +
+	ggtitle("B") +
+	theme(
+		legend.position=c(0.85, 0.9),
+		legend.title = element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+g3 <- ggplot(tt) +
+	geom_line(aes(time/12, hazRate), col=2) +
[…]

### 2019-07-26 doc/speedstrength.tex
tweaks

@@
 Like epidemic strength, epidemic speed measures how disease spreads in the population. 
-Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's strength on the same scale as the reproductive number. 
+Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's strength on the same scale as the reproductive number \swp{(and by measuring an intervention's speed on the same scale as the initial exponential growth rate)}.
 We then show that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed. 
@@
 instead, we use these examples to demonstrate how strength- and speed-based decompositions can help us better evaluate each control strategy under different scenarios.
-In particular, we study how the amount of early HIV transmission can affect our decisions.
+In particular, we study how the amount of early HIV transmission affects our decision.
 
@@
 \end{equation}
-The first gamma distribution, $\tsub{f}{early}(\tau)$, models early HIV transmission during the acute infection stage.
+The first component, $\tsub{f}{early}(\tau)$, models early HIV transmission during the acute infection stage.
 We assume that $\tsub{f}{early}(\tau)$ has a mean of 3 months \citep{hollingsworth2008hiv} and a shape parameter of 3.
-The second gamma distribution, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
-We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \citep{babiker2000time} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
+The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
+We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \swp{need to cite something and/or change this value} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
 Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.
@@
 As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
-Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from a HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
+Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
 Further increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number (\figref{example}C);
@@
 
-First, consider a condom intervention, which is approximately 80\% effective in reducing HIV seroconversion \citep{weller2002condom}.
+\swp{Need values}
+First, consider a condom intervention that is approximately 80\% effective in reducing HIV seroconversion \citep{weller2002condom}.
 Assuming that ?\% of the sexually active population uses condom consistently,

### 2019-09-11 doc/speedstrength.tex
Meeting with Daniel
Notes during meeting

@@
-\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
 \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
@@
 \newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
-\newcommand{\jd}[1]{\comment{cyan}{BMB}{#1}}
+\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
 \newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
@@
 \maketitle
+
+\section{To do}
+
+JD to edit once-through by 2019 Sep 18. I don't know if we can meet.
+
+Add a bigger perspective from Eaton and Hallett (SWP can start on that if time). Look also (less important) and Bellan and Dushoff's follow-up to Hollingsworth.
+
+Talk about speed-like and strength-like estimates of an epidemic (e.g., HIV growth rate vs.~measles mean susceptibility) and of an intervention.
+
+JD to do a condom calculation framing.
 

### 2019-09-25 doc/speedstrength.tex
Some edits; I owe you a lot more.

@@
 
-JD to do a condom calculation framing.
-
 \section{Introduction}
@@
 The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
-the idea of averaging by defining a `typical' case was formalized in the 1990s \citep{diekmann1990definition}.
+the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
 $\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.
@@
 Like epidemic strength, epidemic speed measures how disease spreads in the population. 
-Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's strength on the same scale as the reproductive number \swp{(and by measuring an intervention's speed on the same scale as the initial exponential growth rate)}.
-We then show that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed. 
+Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
+We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed.
+\jd{Tried to clarify in your direction.}
 We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
@@
 
-\swp{edited to make the notation slightly clearer}
-We model disease incidence using the renewal-equation framework \citep{champredon2015intrinsic}. 
-This is a simple, flexible framework that can cover a wide range model structures \citep{Champredon2018equivalence}.
+We model disease incidence using the renewal-equation framework, a simple, flexible framework that can cover a wide range of model structures \citep{Champredon2018equivalence}.
 In our model, disease incidence at time $t$ is given by:
@@
 Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$.
-We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention is larger than the strength of spread.
+We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread \RR.
 
@@
 where $L(\tau)$ is the average proportional reduction for an individual infected time $\tau$ ago.
-Then, we define the post-intervention reproductive number with the following integral:
+The post-intervention reproductive number is thus:
 \begin{equation}
-	\Rhat = \int K(\tau) d\tau
+	\Rhat = \int \hat K(\tau) d\tau
 \end{equation}
-This framework generalizes the work of \cite{fraser2004factors} who made parameteric assumptions about the shape of $1/L(\tau)$.
+This framework generalizes the work of \cite{fraser2004factors} who made parametric assumptions about the shape of $1/L(\tau)$.
 
@@
 \textbf{Understanding condom intervention using strength-based decomposition.}
-A. Condom intervention $\tsub{L}{condom}$ can be represeted as a constant function.
-B. The amount of early transmission affects the strength of an epidemic but not the strength of condom intervention. 
+A.
+Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $L$ due to condom use is constant across the course of infection.
+B. The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
 }
@@
 
-\swp{Need values}
-First, consider a condom intervention that is approximately 80\% effective in reducing HIV seroconversion \citep{weller2002condom}.
-Assuming that ?\% of the sexually active population uses condom consistently,
-we can model this intervention as a constant function: $\tsub{L}{condom} = ???$ (\figref{condom}A).
-In this case, the strength of condom intervention is always constant regardless of the amount of early transmission, whereas the strength of epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
-Therefore, we expect condom intervention to be able to control the spread of the HIV when more than ???\% of transmission is caused by early transmission (i.e., when the intervention strength is greater than the epidemic strength).
+We consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
+\jd{I'm not sure we need to be worried about weller2002condom, nor about population vs.~individual level effects.
+Many estimates of efficacy are higher than 80\%, and it's certainly possible to imagine an intervention that involves education about condom use.}
+Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
+The estimated strength of such an intervention is simply the average of $L$: $\theta=4$, whereas the estimated strength of the epidemic increases as the proportion of early transmission decreases (\figref{condom}B).
+Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
 

### 2019-09-26 doc/speedstrength.tex
Comments for Daniel

@@
 
-\section{Intervention strategies for HIV}
+\section{Example: HIV}
+
+\jd{I'd like to be even less detailed about HIV, and maybe do an explicit comparison with measles or Ebola.
+The idea I'd love to get across is that the two ways of thinking can both be helpful to intuition, and that which one fits better will depend on the disease and on the available data.
+I will work on making that part explicit.}
 
@@
 The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
-We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \swp{need to cite something and/or change this value} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
+We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \swp{need to cite something and/or change this value} \jd{Happy to use any value you can cite, or to help search} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
 Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.
@@
 On the other hand, consider a test-and-treat strategy in which infected individuals are identified and put to clinical care (e.g., antiretroviral therapy).
-We can model this intervention $\tsub{L}{test}$ as a saturating function that begins at 1 and plateaus at a sufficiently high value (such that identified individuals are effectively no longer infectious);
-in particular, we assume that all individuals will be identified and put to treatment within the first 5 years since their infection (\figref{test}A).
-Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength; 
-however, it is unclear why this should be the case.
+We assume that the rate of identification and removal is initially constant.
+Thus, the proportion continuing to transmit declines exponentially, and the effect of intervention $\tsub{L}{test}$ increases exponentially.
+\jd{Can you remind me the functional form that we used?}
+However, we assume removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
+\jd{I'm not sure the advantage of the tsub; I kind of feel like we can communicate in context without subscripts.} (\figref{test}A).
+Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength; however, it is unclear why this should be the case.
+\jd{Ref Eaton}
 
@@
 The speed of epidemic stays constant regardless of the proportion of the early transmission because epidemic speed corrresponds to the observed growth rate.
-The speed of intervention stays roughly constant but slightly increases as proportion of early transmission increases \swp{because?}.
-This framework clearly demonstrates that test-and-treat intervention will be effective regardless of the proportion of early transmission.
+The speed of intervention stays roughly constant but slightly increases as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+
+This framework provides an intuitive underpinning for the originally surprising result of Eaton: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
 

### 2019-11-06 doc/speedstrength.tex
Autocommit (doc)
Not meeting ☹

@@
 
-JD to edit once-through by 2019 Sep 18. I don't know if we can meet.
-
 Add a bigger perspective from Eaton and Hallett (SWP can start on that if time). Look also (less important) and Bellan and Dushoff's follow-up to Hollingsworth.
@@
 We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed.
-\jd{Tried to clarify in your direction.}
 We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
@@
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$. 
-This means we are neglecting changes in susceptible proportion through time;
+Importantly, this means we are neglecting changes in susceptible proportion through time;
 under this assumption, the renewal equation is equivalent to the Von Foerster equations, outlined in \cite{fraser2004factors}.
@@
 
+\jd{I'm wondering if we should be diagrammatic about these decompositions.
+I guess you've seen some of the pictures I show during talks about constant-speed and constant-strength interventions and the relationship to generation intervals.}
+
 \subsection{Speed-based decomposition}
 
-The Euler-Lokta equation allows us to calculate the initial exponential growth rate $r$ of an epidemic given an infection kernel $K$:
+The Euler-Lotka equation allows us to calculate the initial exponential growth rate $r$ of an epidemic given an infection kernel $K$:
 \begin{equation}
@@
 \end{equation}
-By analogy with the strength-based factorization \eqref{strength}, we can rewrite the Euler-Lotka equation \eqref{euler} as a speed-based factorization:
+By analogy with the strength-based factorization \eqref{strength}, we can rewrite \eqref{euler} as a speed-based factorization:
 \begin{equation}
@@
 \end{equation}
-The speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.
+Specifically, the speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.
 

### 2019-11-13 doc/speedstrength.tex
trying to work

@@
 
-An epidemic can be characterized by its \emph{speed} and \emph{strength}.
+An epidemic can be described by its \emph{speed} and \emph{strength}.
 The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
-The strength of an epidemic, often characterized by the reproductive number $\RR$, measures how many new cases are caused by a typical \emph{individual} case.
+The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
 Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.
 
-Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
+Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
 The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
 the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
-$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.
+$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.
+
+The study of measles dynamics provides a key example for which the characterization of an epidemic via the reproduction number has been successful.
+One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
+This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.
 
 Like epidemic strength, epidemic speed measures how disease spreads in the population. 
+\swp{Ebola}
+
+\swp{Eaton and Hallett}
+
 Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 

### 2019-11-13 doc/speedstrength.tex
writing is hard

@@
 
-The study of measles dynamics provides a key example for which the characterization of an epidemic via the reproduction number has been successful.
+The study of measles dynamics provides a key example for which the characterization of an epidemic via the basic reproduction number has been successfully applied in controlling the disease.
 One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
 This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.
+Since then, measles control policies have revolved around targetting this critical vaccination threshold \citep{venczel2003measles, wolfson2007has, takahashi2015reduced}.
+\swp{More citations later}
 
-Like epidemic strength, epidemic speed measures how disease spreads in the population. 
-\swp{Ebola}
+Like epidemic strength, epidemic speed measures how disease spreads in a population.
+Despite the prevailing focus on the basic reproduction number in literature \swp{CITE}, 
+relying on epidemic speed can be sometimes more effective in characterizing the course of an epidemic and evoluating intervention strategies.
+In particular, the exponential growth rate can be directly measured from incidence data even early in an epidemic \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}, whereas the basic reproduction number can be difficult to estimate \citep{weitz2015modeling};
+in these scenarios, intervention strategies should seek to reduce the epidemic speed.
+The West African Ebola outbreak in 2014 provides a key example for this perspective: 
+many researchers tried to estimate the basic reproduction number but the major intervention strategies involved slowing down the epidemic by identifying infected individuals and isolating them \citep{pandey2014strategies}.
 

### 2019-11-14 doc/speedstrength.tex
more work

--- 2019-11-13 doc/speedstrength.tex
@@
 
+\swp{r strategy vs K strategy somewhere}
+
 \section{Introduction}
@@
 
-The study of measles dynamics provides a key example for which the characterization of an epidemic via the reproduction number has been successful.
+The study of measles dynamics provides a key example for which the characterization of an epidemic via the basic reproduction number has been successfully applied in controlling the disease.
 One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
 This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.
+Since then, measles control policies have revolved around targetting this critical vaccination threshold \citep{venczel2003measles, wolfson2007has, takahashi2015reduced}.
+\swp{More citations later}
+\swp{I think I need to end the paragraph more braodly so that it feels like strength vs speed comparison...}
 
-Like epidemic strength, epidemic speed measures how disease spreads in the population. 
-\swp{Ebola}
+Like epidemic strength, epidemic speed measures how disease spreads in a population.
+Despite the prevailing focus on the basic reproduction number in literature \swp{CITE},
+using epidemic speed can be sometimes more effective in characterizing the course of an epidemic and evoluating intervention strategies.
+Early in an outbreak, the exponential growth rate can be directly measured from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}, even though the basic reproduction number can be difficult to estimate \citep{weitz2015modeling}.
+In these scenarios, we expect control efforts that seek to reduce the epidemic speed to be more effective.
+The West African Ebola outbreak in 2014 provides an example for this perspective: 
+many researchers tried to estimate the basic reproduction number but the major strategies involved slowing down the epidemic by identifying infected individuals and isolating them \citep{pandey2014strategies}.
+Strategies that relied on reducing the epidemic strength (e.g., Ebola vaccine trials) were not implemented until much later and had relatively small effect on controlling the outbreak.
+\swp{More citations later}
 
-\swp{Eaton and Hallett}
+\swp{Eaton and Hallett?}
 
@@
 
+\begin{figure}[!t]
+\includegraphics[width=\textwidth]{../figure/constant_intervention.pdf}
+\caption{
+\textbf{Effects of constant strength and speed intervention on infection kernels.}
+Ebola-like gamma infection kernel $K(\tau)$ (mean: 16.2 days, CV: 0.58, and $\RR_0$: 1.5) is shown in black \citep{park2019practical}.
+The infection kernel after applying each intervention strategy $\hat K(\tau)$ is shown in red.
+(A) A constant strength intervention with $\theta = 1.5$ is applied to Ebola-like infection kernel.
+A constant strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; the resulting distribution is equivalent to the intrinsic generation-intervla distribution ($\hat K(\tau) = g(\tau)$) when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$).
+(B) A constant speed intervention with $\phi \approx 0.0267\,\mathrm{days}^{-1}$ is applied to Ebola-like infection kernel.
+A constant speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$) when the speed of intervention matches the speed of epidemic ($\phi = r$).
+}
+\figlab{constant}
+\end{figure}
+
 Assuming that the infection kernel $K$ doesn't change with time, we write:
@@
 \end{equation}
-where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \citep{champredon2015intrinsic}.
+where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of an average individual as a function of time since infection) \citep{champredon2015intrinsic}.
 Since $g$ is a distribution, it integrates to 1, and the reproduction number $\RR$ is thus the integral of $K$.
 
-Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles. We then have:
+Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles (\figref{constant}A). We then have:
 \begin{equation}
@@
 I guess you've seen some of the pictures I show during talks about constant-speed and constant-strength interventions and the relationship to generation intervals.}
+\swp{I tried}
 
@@
 
-Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation, for example by identifying and isolating infectious individuals.
+Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation (\figref{constant}B), for example by identifying and isolating infectious individuals.
 We then have:
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/figure1.pdf}
+\includegraphics[width=\textwidth]{../figure/HIV.pdf}
 \caption{
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/figure2.pdf}
+\includegraphics[width=\textwidth]{../figure/condom.pdf}
 \caption{
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/figure3.pdf}
+\includegraphics[width=\textwidth]{../figure/test_and_treat.pdf}
 \caption{

### 2019-11-14 figure/HIV.R

@@ -70,4 +70,4 @@ g3 <- ggplot(earlydata) +
 
 gtot <- arrangeGrob(g1, g2, g3, nrow=1)
 
-ggsave("figure1.pdf", gtot, width=8, height=3)
+ggsave("HIV.pdf", gtot, width=8, height=3)

### 2019-11-14 figure/condom.R

@@ -51,4 +51,4 @@ g2 <- ggplot(earlydata) +
 
 gtot <- arrangeGrob(g1, g2, nrow=1)
 
-ggsave("figure2.pdf", gtot, width=8, height=3)
+ggsave("condom.pdf", gtot, width=8, height=3)

### 2019-11-14 figure/constant_intervention.R

@@ -1,0 +1,60 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+source("../R/kernel.R")
+
+gbar <- 16.2
+gshape <- 1/0.58^2
+
+tvec <- seq(0, 50, by=0.01)
+
+g1 <- data.frame(
+  time=tvec,
+  density=1.5*dgamma(tvec, shape=gshape, rate=gshape/gbar)
+)
+
+g2 <- data.frame(
+  time=tvec,
+  density=dgamma(tvec, shape=gshape, rate=gshape/gbar)
+)
+
+p1 <- ggplot(data=g1) +
+  geom_line(aes(time, density, col="Baseline")) +
+  geom_line(data=g2, aes(time, density, col="Controlled")) +
+  geom_segment(aes(x=10.75, xend=10.75, y=0.07, yend=0.053),
+               arrow = arrow(length = unit(0.3, "cm"))) +
+  ggtitle("A. Constant strength intervention") +
+  scale_x_continuous("Time (days)") +
+  scale_y_continuous("Density") +
+  scale_color_manual(values=c(1, 2)) +
+  theme(
+    panel.grid = element_blank(),
+    panel.border = element_blank(),
+    axis.line=element_line(),
+    legend.position = c(0.85, 0.9),
+    legend.title = element_blank()
+  )
+
+b0 <- b0fun(g2, R2r(g2, 1.5))
+
+p2 <- ggplot(data=g1) +
+  geom_line(aes(time, density, col="Baseline")) +
+  geom_line(data=b0, aes(time, density, col="Controlled")) +
+  geom_segment(aes(x=10.75, xend=9.39, y=0.07, yend=0.06),
+               arrow = arrow(length = unit(0.3, "cm"))) +
+  ggtitle("B. Constant speed intervention") +
+  scale_x_continuous("Time (days)") +
+  scale_y_continuous("Density") +
+  scale_color_manual(values=c(1, 2)) +
+  theme(
+    panel.grid = element_blank(),
+    panel.border = element_blank(),
+    axis.line=element_line(),
+    legend.position = c(0.85, 0.9),
+    legend.title = element_blank()
+  )
+
+ptot <- arrangeGrob(p1, p2, nrow=1)
+
+ggsave("constant_intervention.pdf", ptot, width=8, height=3)

### 2019-11-14 figure/test_and_treat.R

@@ -104,4 +104,4 @@ g4 <- ggplot(speeddata) +
 
 gtot <- arrangeGrob(g1, g2, g3, g4, nrow=2)
 
-ggsave("figure3.pdf", gtot, width=8, height=6)
+ggsave("test_and_treat.pdf", gtot, width=8, height=6)

### 2019-11-15 doc/speedstrength.tex
bigger perspective?

@@
 Like epidemic strength, epidemic speed measures how disease spreads in a population.
-Despite the prevailing focus on the basic reproduction number in literature \swp{CITE},
-using epidemic speed can be sometimes more effective in characterizing the course of an epidemic and evoluating intervention strategies.
+Despite the prevailing focus on the basic reproduction number in literature,
+using the growth rate can be sometimes more effective in characterizing the course of an epidemic and evoluating intervention strategies.
 Early in an outbreak, the exponential growth rate can be directly measured from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}, even though the basic reproduction number can be difficult to estimate \citep{weitz2015modeling}.
@@
 many researchers tried to estimate the basic reproduction number but the major strategies involved slowing down the epidemic by identifying infected individuals and isolating them \citep{pandey2014strategies}.
-Strategies that relied on reducing the epidemic strength (e.g., Ebola vaccine trials) were not implemented until much later and had relatively small effect on controlling the outbreak.
+Strategies that relies on reducing the epidemic strength (e.g., Ebola vaccine trials) were not implemented until much later and had relatively small effect on controlling the outbreak.
 \swp{More citations later}
 
-\swp{Eaton and Hallett?}
+Whether strength- or speed-based perspective is more useful depends on the disease as well as the focal intervention strategy.
+For example, main strategies for controlling the Human immunodeficiency virus (HIV) epidemic are condom, pre-exposure prophylaxis (PrEP), and antiretroviral therapy (ART).
+Like vaccine, condom and PrEP are reduce the mean susceptibility of the population, which in turn reduce the average number of secondary cases. Therefore, they are strength-like.
+On the other hand, ART relies on identifying infected individuals and treating them. This is speed-like because its effectiveness depends on how fast infected individuals can be put on treatment.
+\cite{eaton2014proportion} used a detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART and gave a strength-based interpretation: increasing the amount of early transmission decreases the basic reproduction number, which negatively correlates with the outcome of the ART intervention.
+However, there's a simple speed-based intuition behind this phenomenon: we expect to be able to control the outbreak if we can identify infected individuals and put them on ART faster than the \emph{observed} rate at which new incidence is generated -- this observed rate does not depend on the amount of early transmission.
 
@@
 \jd{I'm not sure the advantage of the tsub; I kind of feel like we can communicate in context without subscripts.} (\figref{test}A).
-Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength; however, it is unclear why this should be the case.
-\jd{Ref Eaton}
+Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength \citep{eaton2014proportion}.
 

### 2020-01-05 doc/speedstrength.tex
Autocommit (doc)
tweaks

--- 2019-11-14 doc/speedstrength.tex
@@
-\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[12pt]{article}
+\usepackage[]{graphicx}
+\usepackage[]{color}
 \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
@@
 
-\section{To do}
+\section*{To do}
 
@@
 An epidemic can be described by its \emph{speed} and \emph{strength}.
-The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
+The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
 The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
@@
 
-Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
+Much research has prioritized estimates of the basic reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
 The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
 the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
-$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.
-
-The study of measles dynamics provides a key example for which the characterization of an epidemic via the basic reproduction number has been successfully applied in controlling the disease.
-One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
-This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.
-Since then, measles control policies have revolved around targetting this critical vaccination threshold \citep{venczel2003measles, wolfson2007has, takahashi2015reduced}.
-\swp{More citations later}
-\swp{I think I need to end the paragraph more braodly so that it feels like strength vs speed comparison...}
-
-Like epidemic strength, epidemic speed measures how disease spreads in a population.
-Despite the prevailing focus on the basic reproduction number in literature \swp{CITE},
-using epidemic speed can be sometimes more effective in characterizing the course of an epidemic and evoluating intervention strategies.
-Early in an outbreak, the exponential growth rate can be directly measured from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}, even though the basic reproduction number can be difficult to estimate \citep{weitz2015modeling}.
-In these scenarios, we expect control efforts that seek to reduce the epidemic speed to be more effective.
-The West African Ebola outbreak in 2014 provides an example for this perspective: 
-many researchers tried to estimate the basic reproduction number but the major strategies involved slowing down the epidemic by identifying infected individuals and isolating them \citep{pandey2014strategies}.
-Strategies that relied on reducing the epidemic strength (e.g., Ebola vaccine trials) were not implemented until much later and had relatively small effect on controlling the outbreak.
-\swp{More citations later}
-
-\swp{Eaton and Hallett?}
+$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.
 
 Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
-We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed.
-We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
+We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR_0=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed.
+We argue that the historical primacy of $\RR_0$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
 
@@
 
-We model disease incidence using the renewal-equation framework, a simple, flexible framework that can cover a wide range of model structures \citep{Champredon2018equivalence}.
+We model disease incidence using the renewal equation, a simple, flexible framework that can cover a wide range of model structures \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
 In our model, disease incidence at time $t$ is given by:
@@
 Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
-In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible.
+In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible, $S(t)$.
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
-Importantly, this means we are neglecting changes in susceptible proportion through time;
+Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx 1$);
 under this assumption, the renewal equation is equivalent to the Von Foerster equations, outlined in \cite{fraser2004factors}.
@@
 \begin{equation}
-	K(\tau) = \RR g(\tau),
+	K(\tau) = \RR_0 g(\tau),
 	\eqlab{strength}
 \end{equation}
-where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of an average individual as a function of time since infection) \citep{champredon2015intrinsic}.
-Since $g$ is a distribution, it integrates to 1, and the reproduction number $\RR$ is thus the integral of $K$.
+where $g(\tau)$ is the ``intrinsic'' generation-interval distribution.
+The generation intervel is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
+therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
+Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\RR_0$ is thus the integral of $K$.
 
-Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles (\figref{constant}A). We then have:
+Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
 \begin{equation}
-	\hat K(\tau) = (\RR/\theta) g(\tau).
+	\hat K(\tau) = (\RR_0/\theta) g(\tau).
 \end{equation}
-Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$.
-We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread \RR.
+Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR_0$.
+We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread $\RR_0$.
 
@@
 
-We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$.
+We define the strength of the intervention $L$ to be $\theta = \RR_0/\Rhat$.
 It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
@@
 \end{equation}
-In this more general case, we have again that the disease cannot spread when $\theta \geq \RR$.
-
-\jd{I'm wondering if we should be diagrammatic about these decompositions.
-I guess you've seen some of the pictures I show during talks about constant-speed and constant-strength interventions and the relationship to generation intervals.}
-\swp{I tried}
+In this more general case, we have again that the disease cannot spread when $\theta \geq \RR_0$.
 
@@
 
-\section{Example: HIV}
+\section{Example: Human immunodeficiency virus (HIV)}
 
@@
 The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
-We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \swp{need to cite something and/or change this value} \jd{Happy to use any value you can cite, or to help search} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
+We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years to reflect its long incubation period \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
 Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.
@@
 The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
-C. Increasing the proportion of early transmission reduces the estimate of the reproduction number.
+C. Increase in the estimate of the amount of early transmission reduces the estimate of the reproduction number.
 }
@@
 As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
-Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
-Further increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number (\figref{example}C);
-therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is more early transmission.
+Based on the growth rate estimate ($0.452\,\mathrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
+Increase in the estimate of the amount of early transmission shortens the mean generation time, which in turn decreases the reproduction number (\figref{example}C).
+Therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is less early transmission.
 
@@
 A.
-Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $L$ due to condom use is constant across the course of infection.
+Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
 B. The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
@@
 
-We consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
-\jd{I'm not sure we need to be worried about weller2002condom, nor about population vs.~individual level effects.
-Many estimates of efficacy are higher than 80\%, and it's certainly possible to imagine an intervention that involves education about condom use.}
+First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
 Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
-The estimated strength of such an intervention is simply the average of $L$: $\theta=4$, whereas the estimated strength of the epidemic increases as the proportion of early transmission decreases (\figref{condom}B).
+The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$: $\theta=4$, whereas the estimated strength of the epidemic increases as the proportion of early transmission decreases (\figref{condom}B).
 Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
@@
 \textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
-A. Test-and-treat intervention $\tsub{L}{test}$ starts at 1 and saturates after about 5 years.
-B. Increasing the amount of early transmission decreases the strength of an epidemic as well as the strength of test-and-treat intervention.
-C. Hazard rate $\tsub{h}{test}$ of test-and-treat intervention $\tsub{L}{test}$ peaks and plateaus very quickly and reaches zero after about 5 years.
-D. The amount of early transmission has no effect on the speed of an epidemic and small effect on the speed of intervention.
-We assume $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau h(\sigma) d\sigma \right)$ and $h(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $12$ months and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 1/6$.
+A. The effectiveness of test-and-treat intervention is thought to increase exponentially until it saturates after about 5 years.
+B. Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
+C. The hazard rate of test-and-treat intervention stays roughly constant for the first few years and gradually decreases until it reaches zero after about 5 years.
+D. The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
+Test-and-treat intervention is modeled phenomenologically: $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau \tsub{h}{test}(\sigma) d\sigma \right)$ and $\tsub{h}{test}(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $1$ year and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 2\,\mathrm{year}^{-1}$.
 }
@@
 
-On the other hand, consider a test-and-treat strategy in which infected individuals are identified and put to clinical care (e.g., antiretroviral therapy).
+On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy \citep{garnett2009treating, nah2017test}.
 We assume that the rate of identification and removal is initially constant.
 Thus, the proportion continuing to transmit declines exponentially, and the effect of intervention $\tsub{L}{test}$ increases exponentially.
-\jd{Can you remind me the functional form that we used?}
 However, we assume removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
-\jd{I'm not sure the advantage of the tsub; I kind of feel like we can communicate in context without subscripts.} (\figref{test}A).
-Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength; however, it is unclear why this should be the case.
-\jd{Ref Eaton}
+(\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
+Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength \citep{eaton2014proportion}.
 
@@
 
-This framework provides an intuitive underpinning for the originally surprising result of Eaton: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
+This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
 

### 2020-01-05 figure/test_and_treat.R

@@ -65,10 +65,10 @@ g2 <- ggplot(earlydata) +
 	)
 
 g3 <- ggplot(tt) +
-	geom_line(aes(time/12, hazRate), col=2) +
-	geom_hline(aes(yintercept=phiFun(b_base, tt)), lty=2, col=2) +
+	geom_line(aes(time/12, hazRate*12), col=2) +
+	geom_hline(aes(yintercept=phiFun(b_base, tt)*12), lty=2, col=2) +
 	scale_x_continuous("Time (years)") +
-	scale_y_continuous(expression(h[test])) +
+	scale_y_continuous(expression(h[test]~(year^{-1}))) +
 	ggtitle("C") +
 	theme(
 		panel.grid = element_blank(),
@@ -92,7 +92,7 @@ g4 <- ggplot(speeddata) +
 	geom_line(aes(early, speed, col="Intervention")) +
 	geom_hline(aes(yintercept=coef(lfit)[[2]], col="Epidemic" )) +
 	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
-	scale_y_continuous(expression(Speed~(year^{-1})) , limits=c(0.3, 0.8)) +
+	scale_y_continuous(expression(Speed~(year^{-1})), limits=c(0, 0.8), expand=c(0, 0)) +
 	scale_color_manual(values=c("black", "red")) +
 	ggtitle("D") +
 	theme(

### 2020-01-06 doc/speedstrength.tex
more edits

@@
 
-\title{Speed and strength of epidemic intervention}
+\title{Speed and strength of an epidemic intervention}
 \author{Jonathan Dushoff, Sang Woo Park}
@@
 
-Add a bigger perspective from Eaton and Hallett (SWP can start on that if time). Look also (less important) and Bellan and Dushoff's follow-up to Hollingsworth.
+Add a bigger perspective from Eaton and Hallett (SWP can start on that if time).
+Look also (less important) and Bellan and Dushoff's follow-up to Hollingsworth.
+\swp{You have so many papers with Bellan...
+I can't figure out which one you're talking about...
+I added some stuff from Eaton and Hallett in the discussion}
 
@@
 
-\swp{r strategy vs K strategy somewhere}
+\swp{Where do we want to add this idea from Fred Adler: The mechanism of density regulation can determine which index of fitness
+is appropriate. For example, r is a valid fitness criterion under
+density regulation if the population is regulated by density-dependent
+effects that affect each individual (or age class) identically (Pasztor
+et al. 1996). R0 is appropriate as an index of fitness if the population
+is regulated by density-dependent juvenile mortality (Mylius and
+Diekmann 1995).}
 
@@
 A. The effectiveness of test-and-treat intervention is thought to increase exponentially until it saturates after about 5 years.
+The dashed line represents the strength of the intervention.
 B. Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
 C. The hazard rate of test-and-treat intervention stays roughly constant for the first few years and gradually decreases until it reaches zero after about 5 years.
+The dashed line represents the speed of the intervention.
 D. The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
@@
 
-On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy \citep{garnett2009treating, nah2017test}.
+On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy (ART) \citep{garnett2009treating, nah2017test}.
 We assume that the rate of identification and removal is initially constant.
@@
 (\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
-Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength \citep{eaton2014proportion}.
+Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength.
 
@@
 The speed of intervention stays roughly constant but slightly increases as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
-
 This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
@@
 
+The effectiveness of an epidemic intervention is often measured by its ability to reduce the basic reproduction below 1.
+In this study, we extend this idea by defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproduction number $\mathcal R_0$) and the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
+We show that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.
+
+We compare two HIV intervention strategies (condom intervention and test-and-treat intervention) to demonstrate that both strength- and speed-based frameworks can be useful in evaluating the effectiveness of an epidemic intervention.
+In particular, we recapitulate the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART.
+While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproduction number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control the outbreak if we can identify infected individuals and put them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
+While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.
+
+The contrast between strength- and speed-based perspectives in characterizing an epidemic and its control can be widely found in studies of the West African Ebola Outbreak in 2014.
+Early in the outbreak, many researchers tried to estimate the basic reproduction number and made policy suggestions based on their ability to reduce the basic reproduction number below 1 \citep{althaus2014estimating, fisman2014early, gomes2014assessing, pandey2014strategies, shaman2014inference, towers2014temporal, who2014ebola}.
+
 \bibliography{speedstrength}

### 2020-01-06 doc/speedstrength.tex
bit more

@@
 An epidemic can be described by its \emph{speed} and \emph{strength}.
-The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
+The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
 The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
@@
 The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
-C. Increase in the estimate of the amount of early transmission reduces the estimate of the reproduction number.
+C. Increase in the estimate of the amount of early transmission reduces the estimate of the reproductive number.
 }
@@
 As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
-Based on the growth rate estimate ($0.452\,\mathrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
-Increase in the estimate of the amount of early transmission shortens the mean generation time, which in turn decreases the reproduction number (\figref{example}C).
+Based on the growth rate estimate ($0.452\,\mathrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproductive number is $3.17$.
+Increase in the estimate of the amount of early transmission shortens the mean generation time, which in turn decreases the reproductive number (\figref{example}C).
 Therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is less early transmission.
@@
 
-The effectiveness of an epidemic intervention is often measured by its ability to reduce the basic reproduction below 1.
-In this study, we extend this idea by defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproduction number $\mathcal R_0$) and the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
+The effectiveness of an epidemic intervention is often measured by its ability to reduce the basic reproductive below 1.
+In this study, we extend this idea by defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$) and the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
 We show that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.
@@
 In particular, we recapitulate the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART.
-While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproduction number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control the outbreak if we can identify infected individuals and put them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
+While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control the outbreak if we can identify infected individuals and put them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
 While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.
 
-The contrast between strength- and speed-based perspectives in characterizing an epidemic and its control can be widely found in studies of the West African Ebola Outbreak in 2014.
-Early in the outbreak, many researchers tried to estimate the basic reproduction number and made policy suggestions based on their ability to reduce the basic reproduction number below 1 \citep{althaus2014estimating, fisman2014early, gomes2014assessing, pandey2014strategies, shaman2014inference, towers2014temporal, who2014ebola}.
+The contrast between strength- and speed-based perspectives in characterizing an epidemic and its control can be also found in studies of the West African Ebola Outbreak in 2014.
+Early in the outbreak, many researchers tried to estimate the basic reproductive number and made policy suggestions based on their estimates \citep{althaus2014estimating, fisman2014early, gomes2014assessing, pandey2014strategies, shaman2014inference, towers2014temporal, who2014ebola}.
+For example, \cite{pandey2014strategies} estimated $\mathcal R_0$ for the Liberian outbreak to be 1.63 and calculated that restricting post-burial transmission can reduce $\mathcal R_0$ below 1 but restricting hospital and community transmission cannot (thus using a strength-based framework).
+However, the strength of an epidemic cannot be reliably estimated early in an outbreak \citep{weitz2015modeling}.
+Its speed can be directly measured from the incidence time series instead \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating} and provides a better perspective for understanding the initial spread.
+In fact, intervention strategies that played major role in controlling the Ebola outbreak were speed-like, such as identifying infected individuals and isolating them \citep{pandey2014strategies}.
 

### 2020-01-06 doc/speedstrength.tex
split

@@
 Diekmann 1995).}
+
+\section*{Abstract}
 
@@
 
-The contrast between strength- and speed-based perspectives in characterizing an epidemic and its control can be also found in studies of the West African Ebola Outbreak in 2014.
+The contrast between strength- and speed-based perspectives in characterizing an epidemic and its control can be widely found in studies of the West African Ebola Outbreak in 2014.
 Early in the outbreak, many researchers tried to estimate the basic reproductive number and made policy suggestions based on their estimates \citep{althaus2014estimating, fisman2014early, gomes2014assessing, pandey2014strategies, shaman2014inference, towers2014temporal, who2014ebola}.
-For example, \cite{pandey2014strategies} estimated $\mathcal R_0$ for the Liberian outbreak to be 1.63 and calculated that restricting post-burial transmission can reduce $\mathcal R_0$ below 1 but restricting hospital and community transmission cannot (thus using a strength-based framework).
-However, the strength of an epidemic cannot be reliably estimated early in an outbreak \citep{weitz2015modeling}.
-Its speed can be directly measured from the incidence time series instead \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating} and provides a better perspective for understanding the initial spread.
-In fact, intervention strategies that played major role in controlling the Ebola outbreak were speed-like, such as identifying infected individuals and isolating them \citep{pandey2014strategies}.
+For example, \cite{pandey2014strategies} estimated $\mathcal R_0$ for the Liberian outbreak to be 1.63 and calculated that restricting post-burial transmission can reduce $\mathcal R_0$ below 1 but restricting hospital and community transmission cannot.
+While these calculations are purely strength-based, these intervention strategies have both strength- and speed-like components: the effectiveness of safe burial practices depends on its ability to reduce the overall transmissibility (strength-like) as well as the amount of time it takes for a deceased individual to be buried (speed-like).
+
+\cite{weitz2015modeling} later demonstrated that the strength of the Ebola outbreak cannot be reliably estimated early in an outbreak.
+Instead, its speed can be directly measured from the incidence time series \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating} and can provide a better framework for understanding the initial spread.
+Many intervention strategies that played major role in controlling the Ebola outbreak were, in fact, speed-like: identifying and isolating infected individuals, contact-tracing, etc.
+\citep{pandey2014strategies}.
+Like the test-and-treat intervention for HIV, the effectiveness of these strategies would not have depended on the uncertainty in the underlying disease generation time (which would have been propagated from the uncertainty in the amount of post-burial transmission).
 

### 2020-01-10 doc/speedstrength.tex
Some light edits

@@
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
+%% jd: Nice command
 
@@
 
+\jd{The whole ``basic'' thing is tricky; we want to linearize for an invading epidemic, but we don't want to assume that we've linearized at the DFE.
+That said, for our main examples, Ebola and HIV, the DFE is the relevant one.}
+
 An epidemic can be described by its \emph{speed} and \emph{strength}.
@@
 Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
-We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR_0=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed.
+We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR_0=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed.
 We argue that the historical primacy of $\RR_0$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
@@
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
-Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx 1$);
-under this assumption, the renewal equation is equivalent to the Von Foerster equations, outlined in \cite{fraser2004factors}.
+Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx 1$ \jd{Not necessarily. This is exactly what I meant by my ``basic'' comment.});
+under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
 
 \subsection{Strength-based decomposition}
+
+\jd{I would like to hyphenate constant-strength and constant-speed when they are used as adjectives.}
 
@@
 The infection kernel after applying each intervention strategy $\hat K(\tau)$ is shown in red.
-(A) A constant strength intervention with $\theta = 1.5$ is applied to Ebola-like infection kernel.
-A constant strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; the resulting distribution is equivalent to the intrinsic generation-intervla distribution ($\hat K(\tau) = g(\tau)$) when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$).
-(B) A constant speed intervention with $\phi \approx 0.0267\,\mathrm{days}^{-1}$ is applied to Ebola-like infection kernel.
-A constant speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$) when the speed of intervention matches the speed of epidemic ($\phi = r$).
+(A) A constant strength intervention with $\theta = 1.5$ is applied to an Ebola-like infection kernel.
+A constant strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$), the resulting distribution is equivalent to the intrinsic generation-interval distribution ($\hat K(\tau) = g(\tau)$).
+(B) A constant speed intervention with $\phi \approx 0.0267/\mathrm{day}$ is applied to the same kernel.
+A constant speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; when the speed of intervention matches the speed of epidemic ($\phi = r$), the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$).
 }
@@
 where $g(\tau)$ is the ``intrinsic'' generation-interval distribution.
-The generation intervel is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
+The generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
 therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
@@
 
-In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the Human immunodeficiency virus (HIV). 
-These examples are meant to provide neither precise nor accurate estimates of the effectiveness of intervention strategies; 
-instead, we use these examples to demonstrate how strength- and speed-based decompositions can help us better evaluate each control strategy under different scenarios.
+In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the human immunodeficiency virus (HIV). 
+These examples are not meant to provide specific estimates of the effectiveness of intervention strategies; 
+instead, we use these examples to demonstrate how strength- and speed-based decompositions can help us better evaluate control strategies under different scenarios.
 In particular, we study how the amount of early HIV transmission affects our decision.
@@
 The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
-We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years to reflect its long incubation period \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
+We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years to reflect its \jd{long incubation period: not sure what this means; is it necessary?} \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
 Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.

### 2020-01-13 doc/speedstrength.tex
Need a break

@@
 	K(\tau) = \RR_0 g(\tau),
-	\eqlab{strength}
+	\eqlab{strengthFactors}
 \end{equation}
@@
 	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
+	\eqlab{strengthMean}
 \end{equation}
@@
 \end{equation}
-By analogy with the strength-based factorization \eqref{strength}, we can rewrite \eqref{euler} as a speed-based factorization:
+By analogy with the strength-based factorization \eqref{strengthFactors}, we can rewrite \eqref{euler} as a speed-based factorization:
 \begin{equation}
@@
 Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
-The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$: $\theta=4$, whereas the estimated strength of the epidemic increases as the proportion of early transmission decreases (\figref{condom}B).
+The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$: $\theta=4$, whereas the estimated strength of the epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
 Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
@@
 \textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
-A. The effectiveness of test-and-treat intervention is thought to increase exponentially until it saturates after about 5 years.
-The dashed line represents the strength of the intervention.
+\jd{We should work on visual implication of quantitative effects: does strength work on a log scale? If not, try at least a scale that goes down to zero.}
+A.
+The effectiveness of test-and-treat intervention is assumed to increase super-exponentially at first, and then saturate.
+The dashed line shows the corresponding effective strength of the intervention (from \eqref{strengthMean}) \jd{For what value of early transmission?}.
 B. Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
-C. The hazard rate of test-and-treat intervention stays roughly constant for the first few years and gradually decreases until it reaches zero after about 5 years.
-The dashed line represents the speed of the intervention.
+C. The hazard rate correesponding to the strength curve in A). 
+The dashed line shows the corresponding effective speed of the intervention (from \eqref{speedMean}) \jd{For what value of early transmission?}.
 D. The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
@@
 (\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
-Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength.
+Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength.
+In this scenario, we predict that the intervention remains effective over the range of considered parameters.
 
-Using the speed-based decomposition allows us to understand the problem much more clearly.
-The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because individuals are actively getting tested and treated);
-the hazard rate goes down soon after the peak and reaches 0 after about 5 years because infected individuals who have access to medical care would have been identified within the first 5 years or so.
+Using the speed-based decomposition allows us to understand the situation much more clearly.
+The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because sexually active individuals are the most likely to seek testing);
+the assumed hazard rate goes down after a few months to reflect the satuaration assumptions discussed above. 
 The speed of epidemic stays constant regardless of the proportion of the early transmission because epidemic speed corrresponds to the observed growth rate.

### 2020-01-14 doc/speedstrength.tex
removing incubation stuff

@@
 The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
-We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years to reflect its \jd{long incubation period: not sure what this means; is it necessary?} \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
+We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
 Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.

### 2020-01-14 doc/speedstrength.tex
light tweaks to ms and edits to figures

@@
 \caption{
-\textbf{Effects of constant strength and speed intervention on infection kernels.}
+\textbf{Effects of constant-strength and constant-speed intervention on infection kernels.}
 Ebola-like gamma infection kernel $K(\tau)$ (mean: 16.2 days, CV: 0.58, and $\RR_0$: 1.5) is shown in black \citep{park2019practical}.
 The infection kernel after applying each intervention strategy $\hat K(\tau)$ is shown in red.
-(A) A constant strength intervention with $\theta = 1.5$ is applied to an Ebola-like infection kernel.
-A constant strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$), the resulting distribution is equivalent to the intrinsic generation-interval distribution ($\hat K(\tau) = g(\tau)$).
-(B) A constant speed intervention with $\phi \approx 0.0267/\mathrm{day}$ is applied to the same kernel.
-A constant speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; when the speed of intervention matches the speed of epidemic ($\phi = r$), the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$).
+(A) A constant-strength intervention with $\theta = 1.5$ is applied to an Ebola-like infection kernel.
+A constant-strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$), the resulting distribution is equivalent to the intrinsic generation-interval distribution ($\hat K(\tau) = g(\tau)$).
+(B) A constant-speed intervention with $\phi \approx 0.0267/\mathrm{day}$ is applied to the same kernel.
+A constant-speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; when the speed of intervention matches the speed of epidemic ($\phi = r$), the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$).
 }
@@
 	1 = \left\langle \frac{\exp(\phi \tau) }{\exp\left(\int_0^\tau h(\sigma) d\sigma\right)} \right\rangle_{b(\tau)}
+	\eqlab{speedMean}
 \end{equation}
@@
 
-\begin{figure}[!t]
+\begin{figure}[!th]
 \includegraphics[width=\textwidth]{../figure/HIV.pdf}
@@
 C. Increase in the estimate of the amount of early transmission reduces the estimate of the reproductive number.
+Black circle indicates the baseline scenario.
 }
@@
 Therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is less early transmission.
+
+First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
+Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
+The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$: $\theta=4$, whereas the estimated strength of the epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
+Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
 
@@
 
-First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
-Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
-The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$: $\theta=4$, whereas the estimated strength of the epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
-Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
-
 \begin{figure}[!t]
@@
 \textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
-\jd{We should work on visual implication of quantitative effects: does strength work on a log scale? If not, try at least a scale that goes down to zero.}
+\swp{We should just say exponentially. I see where you're coming from but the super-exponentiality seems negligible. Plotting on a log scale shows that it's approximately linear.}
 A.
 The effectiveness of test-and-treat intervention is assumed to increase super-exponentially at first, and then saturate.
-The dashed line shows the corresponding effective strength of the intervention (from \eqref{strengthMean}) \jd{For what value of early transmission?}.
+The dashed line shows the corresponding effective strength of the intervention (from \eqref{strengthMean}) assuming 23\% early transmission.
 B. Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
-C. The hazard rate correesponding to the strength curve in A). 
-The dashed line shows the corresponding effective speed of the intervention (from \eqref{speedMean}) \jd{For what value of early transmission?}.
+C. The hazard rate correesponding to the strength curve in (A). 
+The dashed line shows the corresponding effective speed of the intervention (from \eqref{speedMean}) assuming 23\% early transmission.
 D. The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
+Black and red circles indicates the baseline scenarios.
 Test-and-treat intervention is modeled phenomenologically: $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau \tsub{h}{test}(\sigma) d\sigma \right)$ and $\tsub{h}{test}(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $1$ year and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 2\,\mathrm{year}^{-1}$.

### 2020-01-14 figure/HIV.R

@@ -55,11 +55,11 @@ earlydata <- data.frame(
 
 g3 <- ggplot(earlydata) +
 	geom_line(aes(early, R0)) +
-	geom_point(aes(0.23, 3.17)) +
-	geom_segment(aes(x=0.23, y=2, xend=0.23, yend=3.17), lty=2) +
+	geom_point(aes(0.23, 3.17), size=5) +
+	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
 	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
-	scale_x_continuous("Proportion of early transmission") +
-	scale_y_continuous("Reproduction number", expand=c(0, 0)) +
+	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
+	scale_y_continuous("Reproduction number", limits=c(0, 5), expand=c(0, 0)) +
 	ggtitle("C") +
 	theme(
 		legend.position="none",

### 2020-01-14 figure/condom.R

@@ -14,12 +14,12 @@ lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
 g1 <- ggplot() +
 	geom_hline(yintercept=yintercept, col=2) +
 	scale_x_continuous("Time (years)") +
-	scale_y_continuous(expression(L[condom])) +
+	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
 	ggtitle("A") +
 	theme(
 		panel.grid = element_blank(),
 		panel.border = element_blank(),
-		axis.line = element_line()
+		axis.line.y = element_line()
 	)
 
 early <- seq(from=0.1, to=0.4, by=0.01)
@@ -38,8 +38,8 @@ g2 <- ggplot(earlydata) +
 	geom_line(aes(early, R0, col="Epidemic")) +
 	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
 	scale_color_manual(values=c("black", "red")) +
-	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
-	scale_y_continuous("Strength") +
+	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
+	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +
 	ggtitle("B") +
 	theme(
 		legend.position=c(0.85, 0.9),

### 2020-01-14 figure/test_and_treat.R

@@ -18,8 +18,8 @@ tt <- testingFun(maxRate)
 g1 <- ggplot(tt) +
 	geom_line(aes(time/12, strength), col=2) +
 	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
-	scale_x_continuous("Time (years)") +
-	scale_y_continuous(expression(L[test])) +
+	scale_x_continuous("Time (years)", expand=c(0,0), limits=c(0, 16.5)) +
+	scale_y_log10(expression(L[test])) +
 	ggtitle("A") +
 	theme(
 		panel.grid = element_blank(),
@@ -53,8 +53,10 @@ g2 <- ggplot(earlydata) +
 	geom_line(aes(early, R0, col="Epidemic")) +
 	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
 	scale_color_manual(values=c("black", "red")) +
+  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
+  geom_point(data=strengthdata[strengthdata$early==0.23,], aes(early, strength), size=5, col="red") +
 	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
-	scale_y_continuous("Strength") +
+	scale_y_log10("Strength", limits=c(1, 16), expand=c(0, 0), breaks=c(1, 2, 4, 8, 16)) +
 	ggtitle("B") +
 	theme(
 		legend.position=c(0.85, 0.9),
@@ -67,7 +69,7 @@ g2 <- ggplot(earlydata) +
 g3 <- ggplot(tt) +
 	geom_line(aes(time/12, hazRate*12), col=2) +
 	geom_hline(aes(yintercept=phiFun(b_base, tt)*12), lty=2, col=2) +
-	scale_x_continuous("Time (years)") +
+	scale_x_continuous("Time (years)", limits=c(0, 16.5), expand=c(0, 0)) +
 	scale_y_continuous(expression(h[test]~(year^{-1}))) +
 	ggtitle("C") +
 	theme(
@@ -91,6 +93,8 @@ speeddata <- data.frame(
 g4 <- ggplot(speeddata) +
 	geom_line(aes(early, speed, col="Intervention")) +
 	geom_hline(aes(yintercept=coef(lfit)[[2]], col="Epidemic" )) +
[…]

### 2020-01-21 doc/speedstrength.tex
minor tweaks to the figure

@@
 \textbf{The infection kernel of the HIV.}
-A. The infection kernel of the HIV is approximated using a sum of two gamma distributions. We assume that the baseline proportion of early transmission is 23\% \citep{hayes2006amplified}.
-B.
-Time series of HIV prevalence in pregnant women in South Africa, 1990 - 2010 \citep{barron2013eliminating}.
+(A) The infection kernel of the HIV is approximated using a sum of two gamma distributions. We assume that the baseline proportion of early transmission is 23\% \citep{hayes2006amplified}.
+(B) Time series of HIV prevalence in pregnant women in South Africa, 1990 - 2010 \citep{barron2013eliminating}.
 The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
-C. Increase in the estimate of the amount of early transmission reduces the estimate of the reproductive number.
-Black circle indicates the baseline scenario.
+(C) Increase in the estimate of the amount of early transmission reduces the estimate of the reproductive number.
+The black circle indicates the baseline scenario.
 }
@@
 \textbf{Understanding condom intervention using strength-based decomposition.}
-A.
-Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
-B. The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
+(A) Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
+(B) The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
+The black and red circles indicate the baseline scenario.
 }
@@
 \swp{We should just say exponentially. I see where you're coming from but the super-exponentiality seems negligible. Plotting on a log scale shows that it's approximately linear.}
-A.
-The effectiveness of test-and-treat intervention is assumed to increase super-exponentially at first, and then saturate.
+(A) The effectiveness of test-and-treat intervention is assumed to increase super-exponentially at first, and then saturate.
 The dashed line shows the corresponding effective strength of the intervention (from \eqref{strengthMean}) assuming 23\% early transmission.
-B. Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
-C. The hazard rate correesponding to the strength curve in (A). 
+(B) Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
+(C) The hazard rate correesponding to the strength curve in (A). 
 The dashed line shows the corresponding effective speed of the intervention (from \eqref{speedMean}) assuming 23\% early transmission.
-D. The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
-Black and red circles indicates the baseline scenarios.
+(D) The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
+The black and red circles indicate the baseline scenario.
 Test-and-treat intervention is modeled phenomenologically: $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau \tsub{h}{test}(\sigma) d\sigma \right)$ and $\tsub{h}{test}(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $1$ year and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 2\,\mathrm{year}^{-1}$.

### 2020-01-21 figure/condom.R

@@ -37,6 +37,8 @@ earlydata <- data.frame(
 g2 <- ggplot(earlydata) +
 	geom_line(aes(early, R0, col="Epidemic")) +
 	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
+  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
+  geom_point(aes(x=0.23, y=yintercept), size=5, col="red") +
 	scale_color_manual(values=c("black", "red")) +
 	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
 	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +

### 2020-01-22 doc/speedstrength.tex
Working on language and explanations

@@
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
-%% jd: Nice command
+\newcommand{\pEarly}{\ensuremath{\tsub{p}{early}}}
 
@@
 The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
-The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
+Epidemic is characterized by the reproductive number \RR, which measures how many new cases are caused by a typical \emph{individual} case.
 Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.
 
-Much research has prioritized estimates of the basic reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
+ in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}
+Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population, called \Ro, because \RR has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
 The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
 the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
-$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.
+$\RR$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.
 
 Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
-We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR_0=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed.
-We argue that the historical primacy of $\RR_0$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
+We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed.
+We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
 
@@
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
-Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx 1$ \jd{Not necessarily. This is exactly what I meant by my ``basic'' comment.});
+Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx S_0$;
 under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
@@
 \subsection{Strength-based decomposition}
-
-\jd{I would like to hyphenate constant-strength and constant-speed when they are used as adjectives.}
 
@@
 \textbf{Effects of constant-strength and constant-speed intervention on infection kernels.}
-Ebola-like gamma infection kernel $K(\tau)$ (mean: 16.2 days, CV: 0.58, and $\RR_0$: 1.5) is shown in black \citep{park2019practical}.
+Ebola-like gamma infection kernel $K(\tau)$ (mean: 16.2 days, CV: 0.58, and \Ro: 1.5) is shown in black \citep{park2019practical}.
 The infection kernel after applying each intervention strategy $\hat K(\tau)$ is shown in red.
-(A) A constant-strength intervention with $\theta = 1.5$ is applied to an Ebola-like infection kernel.
+(A) The effect of a constant-strength intervention with $\theta = 1.5$.
 A constant-strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$), the resulting distribution is equivalent to the intrinsic generation-interval distribution ($\hat K(\tau) = g(\tau)$).
@@
 \begin{equation}
-	K(\tau) = \RR_0 g(\tau),
+	K(\tau) = \RR g(\tau),
 	\eqlab{strengthFactors}
@@
 therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
-Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\RR_0$ is thus the integral of $K$.
+Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\RR$ is thus the integral of $K$.
 
@@
 \begin{equation}
-	\hat K(\tau) = (\RR_0/\theta) g(\tau).
+	\hat K(\tau) = (\RR/\theta) g(\tau).
 \end{equation}
-Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR_0$.
-We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread $\RR_0$.
+Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$.
+We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread $\RR$.
 
@@
 
-We define the strength of the intervention $L$ to be $\theta = \RR_0/\Rhat$.
+We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$.
 It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
@@
 \end{equation}
-In this more general case, we have again that the disease cannot spread when $\theta \geq \RR_0$.
+In this more general case, we have again that the disease cannot spread when $\theta \geq \RR$.
 
@@
 We define the speed of a general intervention to be $\phi = r - \hat r$. 
-We can then show that $\phi$ is a (sort of) mean satisfying:
+We can then show that $\phi$ is a (sort of) mean satisfying the implicit equation:
 \begin{equation}
@@
 In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the human immunodeficiency virus (HIV). 
-These examples are not meant to provide specific estimates of the effectiveness of intervention strategies; 
-instead, we use these examples to demonstrate how strength- and speed-based decompositions can help us better evaluate control strategies under different scenarios.
-In particular, we study how the amount of early HIV transmission affects our decision.
+In particular, we study how the amount of early HIV transmission affects estimates of intervention effectiveness. 
+These examples are not detailed estimates for specific scenarios; 
+instead, they are meant to demonstrate how strength- and speed-based decompositions can help evaluate control strategies.
 
@@
 
-As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
-Based on the growth rate estimate ($0.452\,\mathrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproductive number is $3.17$.
-Increase in the estimate of the amount of early transmission shortens the mean generation time, which in turn decreases the reproductive number (\figref{example}C).
-Therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is less early transmission.
+The infection kernel is shown in (\figref{example}A) for our baseline value of  
+$\tsub{p}{early} = 0.23$.
+We assume that the initial speed of the epidemic is $r_0=0.452\,\mathrm{year}^{-1}$; \citep{barron2013eliminating} (see \figref{example}B), and ask what value of $\Ro$ would produce this rate of growth.
+When transmission is fast, (i.e., when \pEarly\ is large), individuals don't need to transmit as much to achieve this speed, so the estimated value of \Ro\ decreases (\figref{example}C).
+Therefore, as \pEarly\ gets smaller, we expect stronger intervention to be required in order to control the disease.
 

### 2020-01-24 doc/speedstrength.tex
Working on MS…

@@
 
+We compare two different possible intervention strategies to shed light on the speed and strenght decompositions.
 First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
@@
 
-Using the speed-based decomposition allows us to understand the situation much more clearly.
+Using the speed-based decomposition provides a valuable perspective on this situation.
 The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because sexually active individuals are the most likely to seek testing);
 the assumed hazard rate goes down after a few months to reflect the satuaration assumptions discussed above. 
-The speed of epidemic stays constant regardless of the proportion of the early transmission because epidemic speed corrresponds to the observed growth rate.
-The speed of intervention stays roughly constant but slightly increases as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
+The effective speed of the intervention stays relatively constant, because \jd{WHY?
+Can we understand this part better?
+I was going to say it is driven by the hazard of early detection, but that hazard is higher than I expected compared to the speed.}  It increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
 This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
@@
 
-The effectiveness of an epidemic intervention is often measured by its ability to reduce the basic reproductive below 1.
-In this study, we extend this idea by defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$) and the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
+The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive below 1.
+In this study, we extend this idea by first defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$); and then constructing a parallel interpretation which measures the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
 We show that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.

### 2020-01-24 doc/speedstrength.tex
Time for lab meeting

@@
 \textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
-\swp{We should just say exponentially. I see where you're coming from but the super-exponentiality seems negligible. Plotting on a log scale shows that it's approximately linear.}
-(A) The effectiveness of test-and-treat intervention is assumed to increase super-exponentially at first, and then saturate.
+(A) The strength of the test-and-treat intervention (calculated from the assumed hazard, (C)).
 The dashed line shows the corresponding effective strength of the intervention (from \eqref{strengthMean}) assuming 23\% early transmission.
 (B) Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
-(C) The hazard rate correesponding to the strength curve in (A). 
+(C) The assumed hazard for the test-and-treat intervention. 
 The dashed line shows the corresponding effective speed of the intervention (from \eqref{speedMean}) assuming 23\% early transmission.
-(D) The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
-The black and red circles indicate the baseline scenario.
+(D) The estimated amount of early transmission has little effect on the effective speed of intervention, and none on the speed of the epidemic estimated from incidence data.
+Circles indicate the baseline scenario.
 Test-and-treat intervention is modeled phenomenologically: $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau \tsub{h}{test}(\sigma) d\sigma \right)$ and $\tsub{h}{test}(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $1$ year and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 2\,\mathrm{year}^{-1}$.
@@
 \end{figure}
+
+\jd{I'm horrified that you would suggest saying ``exponentially'' in a non-technical sense in a scientific paper.
+Should we think about showing one or three generation intervals in panels a and c (intrinsic and backward respectively) to help people think about averaging?}
 

### 2020-02-02 doc/speedstrength.tex
comments

@@
 An epidemic can be described by its \emph{speed} and \emph{strength}.
-The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
-Epidemic is characterized by the reproductive number \RR, which measures how many new cases are caused by a typical \emph{individual} case.
+The speed of an epidemic is characterized by the exponential growth rate $r$, which measures how \emph{fast} an epidemic grows at the population level.
+The strength of an epidemic is characterized by the reproductive number \RR, which measures how many new cases are caused by a typical \emph{individual} case.
 Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.
@@
 Should we think about showing one or three generation intervals in panels a and c (intrinsic and backward respectively) to help people think about averaging?}
+\swp{I meant to say approximately exponentially. Based on panel a, it does look like it grows approximately linearly on a log-scale and this rate approximately matches the maximum hazard rate.}
 
@@
 This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
+\swp{I have an explanation.
+The backward GI distribution is incredibly narrow.
+Its mean is 0.97 years but the distribution peaks around 0.2 years.
+Therefore, the inital superexponentiality actually matters quite a lot when we try to average it out, and therefore we get something that's much lower than the peak hazard.
+If we made the initial $h$ constant, rather than starting it from 0, the speed of intervention would match the hazard of early detection much better.}
 
@@
 
-The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive below 1.
+The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number below 1.
 In this study, we extend this idea by first defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$); and then constructing a parallel interpretation which measures the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
@@
 In particular, we recapitulate the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART.
-While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control the outbreak if we can identify infected individuals and put them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
+While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control an outbreak if we can identify infected individuals and treat them with ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
 While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.

### 2020-02-02 doc/speedstrength.tex
tryin to rewrite discussion

@@
 The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number below 1.
-In this study, we extend this idea by first defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$); and then constructing a parallel interpretation which measures the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
-We show that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.
+In this study, we extended this idea by first defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$); and then constructing a parallel interpretation which measures the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
+We showed that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.
 
-We compare two HIV intervention strategies (condom intervention and test-and-treat intervention) to demonstrate that both strength- and speed-based frameworks can be useful in evaluating the effectiveness of an epidemic intervention.
-In particular, we recapitulate the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART.
-While they gave a strength-based argument for the result --  increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion} -- there is a simple speed-based intuition behind this phenomenon: in principle, we can control an outbreak if we can identify infected individuals and treat them with ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
+We compared two HIV intervention strategies (condom intervention and test-and-treat intervention) to demonstrate that both strength- and speed-based frameworks can be useful in evaluating the effectiveness of an epidemic intervention.
+In particular, we provided an alternative explanation for the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART:
+we can control an outbreak if we can identify infected individuals and enroll them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
+On the other hand, the original explanation of the result relied on a strength-based arugment: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.
 While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.

### 2020-02-02 doc/speedstrength.tex
ok

@@
 Talk about speed-like and strength-like estimates of an epidemic (e.g., HIV growth rate vs.~measles mean susceptibility) and of an intervention.
-
-\swp{Where do we want to add this idea from Fred Adler: The mechanism of density regulation can determine which index of fitness
-is appropriate. For example, r is a valid fitness criterion under
-density regulation if the population is regulated by density-dependent
-effects that affect each individual (or age class) identically (Pasztor
-et al. 1996). R0 is appropriate as an index of fitness if the population
-is regulated by density-dependent juvenile mortality (Mylius and
-Diekmann 1995).}
 
@@
 
-The contrast between strength- and speed-based perspectives in characterizing an epidemic and its control can be widely found in studies of the West African Ebola Outbreak in 2014.
-Early in the outbreak, many researchers tried to estimate the basic reproductive number and made policy suggestions based on their estimates \citep{althaus2014estimating, fisman2014early, gomes2014assessing, pandey2014strategies, shaman2014inference, towers2014temporal, who2014ebola}.
-For example, \cite{pandey2014strategies} estimated $\mathcal R_0$ for the Liberian outbreak to be 1.63 and calculated that restricting post-burial transmission can reduce $\mathcal R_0$ below 1 but restricting hospital and community transmission cannot.
-While these calculations are purely strength-based, these intervention strategies have both strength- and speed-like components: the effectiveness of safe burial practices depends on its ability to reduce the overall transmissibility (strength-like) as well as the amount of time it takes for a deceased individual to be buried (speed-like).
-
-\cite{weitz2015modeling} later demonstrated that the strength of the Ebola outbreak cannot be reliably estimated early in an outbreak.
-Instead, its speed can be directly measured from the incidence time series \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating} and can provide a better framework for understanding the initial spread.
-Many intervention strategies that played major role in controlling the Ebola outbreak were, in fact, speed-like: identifying and isolating infected individuals, contact-tracing, etc.
-\citep{pandey2014strategies}.
-Like the test-and-treat intervention for HIV, the effectiveness of these strategies would not have depended on the uncertainty in the underlying disease generation time (which would have been propagated from the uncertainty in the amount of post-burial transmission).
+The contrast between speed- and strength-based perspectives is prevalent in other areas of ecology and epidemiology:
+the instantaneous per capita rate of increase $r$ (speed) and expected life time reproductive success $R_0$ (strength) can both be valid measures of fitness, depending on the mechanism of density dependence \citep{pasztor1996r0};
+whether a species selects for $r$-strategy (speed) or $K$-strategy (strength) relies on ??
+\swp{I've never read MacArthur and Wilson and I can't find the book online.
+Have you read it, JD?
+Can we cite it?};
+and emerging pathogens initially select for high virulence to maximize the growth rate $r$ (speed) but later select for intermediate virulence to maximize the basic reproductive number $\mathcal R_0$ (strength) \citep{bull2008invasion, bolker2010transient}.
+Although this list is not exhaustive, it clearly demonstrates that considering both perspectives can reveal important biological insight for a wide range of applications.
+We strongly suggest that infectious disease modelers
 

### 2020-02-12 doc/speedstrength.tex
11 minutes

@@
 \end{equation}
-This framework generalizes the work of \cite{fraser2004factors} who made parametric assumptions about the shape of $1/L(\tau)$.
+This framework generalizes the work of \cite{fraser2004factors} who made parametric assumptions about the shape of $L(\tau)$.
 
@@
 \section{Example: Human immunodeficiency virus (HIV)}
-
-\jd{I'd like to be even less detailed about HIV, and maybe do an explicit comparison with measles or Ebola.
-The idea I'd love to get across is that the two ways of thinking can both be helpful to intuition, and that which one fits better will depend on the disease and on the available data.
-I will work on making that part explicit.}
 
@@
 
-\jd{I'm horrified that you would suggest saying ``exponentially'' in a non-technical sense in a scientific paper.
-Should we think about showing one or three generation intervals in panels a and c (intrinsic and backward respectively) to help people think about averaging?}
-\swp{I meant to say approximately exponentially. Based on panel a, it does look like it grows approximately linearly on a log-scale and this rate approximately matches the maximum hazard rate.}
-
 On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy (ART) \citep{garnett2009treating, nah2017test}.
@@
 The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because sexually active individuals are the most likely to seek testing);
-the assumed hazard rate goes down after a few months to reflect the satuaration assumptions discussed above. 
+the assumed hazard rate goes down after a few months to reflect the saturation assumptions discussed above. 
 The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
-The effective speed of the intervention stays relatively constant, because \jd{WHY?
-Can we understand this part better?
-I was going to say it is driven by the hazard of early detection, but that hazard is higher than I expected compared to the speed.}  It increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
+\jd{Good?
+I'm tempted to say more, but also want to stay simple}  It increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
 This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
@@
 If we made the initial $h$ constant, rather than starting it from 0, the speed of intervention would match the hazard of early detection much better.}
+\jd{Thanks. I think it would be good to look at three intrinsic and three backward GIs to get an idea of how the averaging works (and of whether we should include such GIs in a figure).
 

### 2020-02-12 doc/speedstrength.tex
Ready for Daniel

@@
 If we made the initial $h$ constant, rather than starting it from 0, the speed of intervention would match the hazard of early detection much better.}
-\jd{Thanks. I think it would be good to look at three intrinsic and three backward GIs to get an idea of how the averaging works (and of whether we should include such GIs in a figure).
+\jd{Thanks. I think it would be good to look at three intrinsic and three backward GIs to get an idea of how the averaging works (and of whether we should include such GIs in a figure).}
 
@@
 
-The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number below 1.
-In this study, we extended this idea by first defining the strength of an intervention on the same scale as the strength of an epidemic (i.e., basic reproductive number $\mathcal R_0$); and then constructing a parallel interpretation which measures the speed of an intervention on the same scale as the speed of an epidemic (i.e., exponential growth rate $r$).
-We showed that an intervention can successfully contain an epidemic if either its strength or speed is greater than their corresponding measures of the epidemic.
+The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \Rx, or outbreak ``strength'' -- below 1.
+The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \Rx\, or even overlooked entirely (ParkPreprint for examples?).
+We argue that \Rx\ and $r$ provide equally valid, complementary perspectives on epidemic control, and that there are situations where each provides a clearer picture than the other.
 
-We compared two HIV intervention strategies (condom intervention and test-and-treat intervention) to demonstrate that both strength- and speed-based frameworks can be useful in evaluating the effectiveness of an epidemic intervention.
+In this study, we:
+first extended the standard paradigm of \Rx\ as critical parameter for control, by defining the strength of an intervention on the same scale as \Rx, the strength of the epidemic; 
+then constructed a parallel interpretation which measures the speed of an intervention on the same scale as $r$, the speed of an epidemic.
+We thus showed that the standard paradigm for \Ro\ and control has a natural parallel interpretation in terms of $r$.
+
+To illustrate this idea, we used simple assumptions to explore the effects of two HIV intervention strategies (condoms and test-and-treat), using both strength- and speed-based frameworks.
 In particular, we provided an alternative explanation for the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART:
@@
 
+\jd{I don't like this as a close; maybe we have to mention some of it \ldots.}
+\jd{We definitely need to cite pasztor1996r0.
+I'm less interested in the Bull and MacArthur evolutionary stuff, which seems tangential to me.
+I wonder what's a good thing to cite for the over-emphasis on big R?
+I kind of feel like the stampede to estimate big R while looking past little R in Ebola and CovID should be part of our story here.}
 The contrast between speed- and strength-based perspectives is prevalent in other areas of ecology and epidemiology:
@@
 Although this list is not exhaustive, it clearly demonstrates that considering both perspectives can reveal important biological insight for a wide range of applications.
-We strongly suggest that infectious disease modelers
+
+We suggest that infectious disease modelers should be aware of the advantages (and complementarity) of these two frameworks when analyzing disease outbreaks.
 

### 2020-02-14 doc/speedstrength.tex
stuff

@@
 
-Add a bigger perspective from Eaton and Hallett (SWP can start on that if time).
-Look also (less important) and Bellan and Dushoff's follow-up to Hollingsworth.
-\swp{You have so many papers with Bellan...
-I can't figure out which one you're talking about...
-I added some stuff from Eaton and Hallett in the discussion}
-
-Talk about speed-like and strength-like estimates of an epidemic (e.g., HIV growth rate vs.~measles mean susceptibility) and of an intervention.
-
 \section*{Abstract}
@@
 \section{Introduction}
-
-\jd{The whole ``basic'' thing is tricky; we want to linearize for an invading epidemic, but we don't want to assume that we've linearized at the DFE.
-That said, for our main examples, Ebola and HIV, the DFE is the relevant one.}
 
@@
 
- in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}
-Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population, called \Ro, because \RR has a threshold value, $\RR=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
+Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population, called \Ro, because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
 The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
@@
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
-Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx S_0$;
-under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
+Importantly, this means we are neglecting changes in susceptible proportion through time: $S(t) \approx S(0)$.
+Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
 
@@
 
-Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
+Imagine a control measure that proportionally reduces $K$, for example, by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
 \begin{equation}
@@
 
-Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation (\figref{constant}B), for example by identifying and isolating infectious individuals.
+Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation (\figref{constant}B), for example, by identifying and isolating infectious individuals.
 We then have:
@@
 The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
-\jd{Good?
-I'm tempted to say more, but also want to stay simple}  It increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
 This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
-\swp{I have an explanation.
-The backward GI distribution is incredibly narrow.
-Its mean is 0.97 years but the distribution peaks around 0.2 years.
-Therefore, the inital superexponentiality actually matters quite a lot when we try to average it out, and therefore we get something that's much lower than the peak hazard.
-If we made the initial $h$ constant, rather than starting it from 0, the speed of intervention would match the hazard of early detection much better.}
-\jd{Thanks. I think it would be good to look at three intrinsic and three backward GIs to get an idea of how the averaging works (and of whether we should include such GIs in a figure).}
 
@@
 The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \Rx, or outbreak ``strength'' -- below 1.
-The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \Rx\, or even overlooked entirely (ParkPreprint for examples?).
+The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \Rx\, or even overlooked entirely \citep{park2020reconciling}.
 We argue that \Rx\ and $r$ provide equally valid, complementary perspectives on epidemic control, and that there are situations where each provides a clearer picture than the other.
@@
 
-\jd{I don't like this as a close; maybe we have to mention some of it \ldots.}
-\jd{We definitely need to cite pasztor1996r0.
-I'm less interested in the Bull and MacArthur evolutionary stuff, which seems tangential to me.
-I wonder what's a good thing to cite for the over-emphasis on big R?
-I kind of feel like the stampede to estimate big R while looking past little R in Ebola and CovID should be part of our story here.}
-The contrast between speed- and strength-based perspectives is prevalent in other areas of ecology and epidemiology:
-the instantaneous per capita rate of increase $r$ (speed) and expected life time reproductive success $R_0$ (strength) can both be valid measures of fitness, depending on the mechanism of density dependence \citep{pasztor1996r0};
-whether a species selects for $r$-strategy (speed) or $K$-strategy (strength) relies on ??
-\swp{I've never read MacArthur and Wilson and I can't find the book online.
-Have you read it, JD?
-Can we cite it?};
-and emerging pathogens initially select for high virulence to maximize the growth rate $r$ (speed) but later select for intermediate virulence to maximize the basic reproductive number $\mathcal R_0$ (strength) \citep{bull2008invasion, bolker2010transient}.
-Although this list is not exhaustive, it clearly demonstrates that considering both perspectives can reveal important biological insight for a wide range of applications.
+The contrast between speed- and strength-based perspectives can be found in other literature as well.
+For example, whether the instantaneous per capita rate of increase $r$ (speed) or expected life time reproductive success $R_0$ (strength) is the correct measures of fitness depends on the mechanism of density dependence \citep{pasztor1996r0}.
+When population is regulated by affecting all individuals identically, $r$ is a valid measure \citep{pasztor1996r0}.
+When population is regulated by affecting density-dependent juvenile mortality, $R_0$ is a valid measure \citep{mylius1995evolutionarily}.
+Even though the contrast between two perspectives was introduced in ecology more than 20 years ago, it is still rare in epidemiology.
 
+Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak clearly demonstrate the lack of emphasis on the speed of an epidemic.
+During the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
+For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
+When an epidemic is growing exponentially, however, the basic reproductive number may not be identifiable \citep{taylor2016stochasticity}.
 We suggest that infectious disease modelers should be aware of the advantages (and complementarity) of these two frameworks when analyzing disease outbreaks.

### 2020-02-14 doc/speedstrength.tex
clean up
ok

@@
 \maketitle
-
-\section*{To do}
 
@@
 $\tsub{p}{early} = 0.23$.
-We assume that the initial speed of the epidemic is $r_0=0.452\,\mathrm{year}^{-1}$; \citep{barron2013eliminating} (see \figref{example}B), and ask what value of $\Ro$ would produce this rate of growth.
+We assume that the initial speed of the epidemic is $r=0.452\,\mathrm{year}^{-1}$; \citep{barron2013eliminating} (see \figref{example}B), and ask what value of $\Ro$ would produce this rate of growth.
 When transmission is fast, (i.e., when \pEarly\ is large), individuals don't need to transmit as much to achieve this speed, so the estimated value of \Ro\ decreases (\figref{example}C).
@@
 Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
-The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$: $\theta=4$, whereas the estimated strength of the epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
+The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$, i.e., $\theta=4$, whereas the estimated strength of the epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
 Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
@@
 In particular, we provided an alternative explanation for the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART:
-we can control an outbreak if we can identify infected individuals and enroll them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the amount of early transmission.
+we can control an outbreak if we can identify infected individuals and enroll them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the estimates of the amount of early transmission.
 On the other hand, the original explanation of the result relied on a strength-based arugment: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.

### 2020-02-24 doc/speedstrength.tex
getting close

@@
 
+While an epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproduction number \RR), the effectiveness of an intervention strategy is often, if not always, determined by its ability to reduce \RR\ below 1.
+Here, we generalize this idea by measuring the speed and strength of intervention on the same scale as the speed and strength of an epidemic, respectively.
+We show that an epidemic can be contained if the speed or strength of intervention is greater than their corresponding counterparts.
+Finally, we compare two intervention strategies for HIV -- namely condom intervention and test-and-treat intervention -- to demonstrate that both speed- and strength-based perspectives can be useful for evaluating intervention strategies under different scenarios.
+
 \section{Introduction}
@@
 
-Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population, called \Ro, because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
-The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
+Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population, called the \emph{basic} reproductive number \Ro, because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
+The insight that a case must on average cause at least one new case under good conditions for a disease to persist goes back $>100$ years \citep{ross1911prevention};
 the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
@@
 In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible, $S(t)$.
-Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
+Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time.
+Thus, we will assume $K(\tau, t) \equiv K(\tau)$.
 Importantly, this means we are neglecting changes in susceptible proportion through time: $S(t) \approx S(0)$.
@@
 The infection kernel is shown in (\figref{example}A) for our baseline value of  
-$\tsub{p}{early} = 0.23$.
-We assume that the initial speed of the epidemic is $r=0.452\,\mathrm{year}^{-1}$; \citep{barron2013eliminating} (see \figref{example}B), and ask what value of $\Ro$ would produce this rate of growth.
+$\tsub{p}{early} = 0.23$.  We assume that the initial speed of the epidemic is $r=0.452\,\mathrm{year}^{-1}$(\figref{example}B), and ask what value of $\Ro$ would produce this rate of growth.
 When transmission is fast, (i.e., when \pEarly\ is large), individuals don't need to transmit as much to achieve this speed, so the estimated value of \Ro\ decreases (\figref{example}C).
@@
 First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
-Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
+Assuming that condoms act as a physical barrier, and that condom use will, on average, remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
 The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$, i.e., $\theta=4$, whereas the estimated strength of the epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
@@
 then constructed a parallel interpretation which measures the speed of an intervention on the same scale as $r$, the speed of an epidemic.
-We thus showed that the standard paradigm for \Ro\ and control has a natural parallel interpretation in terms of $r$.
+We thus showed that the standard paradigm for \Rx\ and control has a natural parallel interpretation in terms of $r$.
 
@@
 On the other hand, the original explanation of the result relied on a strength-based arugment: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.
+
 While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.
+Whether strength- or speed-based perspective is more appropriate depends on the disease as well as the intervention strategies.
+For newly invading pathogens, we expect the speed-based perspective to be almost always better.
+When an epidemic is growing exponentially, the basic reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
 
@@
 When population is regulated by affecting density-dependent juvenile mortality, $R_0$ is a valid measure \citep{mylius1995evolutionarily}.
-Even though the contrast between two perspectives was introduced in ecology more than 20 years ago, it is still rare in epidemiology.
+Even though the contrast between the strength- and speed-based perspectives was formalized in theoretical ecology more than 20 years ago, it is still rare in theoretical epidemiology.
 
@@
 During the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
-For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
-When an epidemic is growing exponentially, however, the basic reproductive number may not be identifiable \citep{taylor2016stochasticity}.
+For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
 We suggest that infectious disease modelers should be aware of the advantages (and complementarity) of these two frameworks when analyzing disease outbreaks.

### 2020-02-25 doc/speedstrength.tex
edits

@@
 
-While an epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproduction number \RR), the effectiveness of an intervention strategy is often, if not always, determined by its ability to reduce \RR\ below 1.
+While an epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproductive number \RR), the effectiveness of an intervention strategy is often, if not always, determined by its ability to reduce \RR\ below 1.
 Here, we generalize this idea by measuring the speed and strength of intervention on the same scale as the speed and strength of an epidemic, respectively.
@@
 Finally, we compare two intervention strategies for HIV -- namely condom intervention and test-and-treat intervention -- to demonstrate that both speed- and strength-based perspectives can be useful for evaluating intervention strategies under different scenarios.
+We suggest that disease modelers should refrain from putting too much emphasis on the reproductive number but rather consider both speed- and strength-based perspectives.
 
@@
 
-The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \Rx, or outbreak ``strength'' -- below 1.
-The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \Rx\, or even overlooked entirely \citep{park2020reconciling}.
-We argue that \Rx\ and $r$ provide equally valid, complementary perspectives on epidemic control, and that there are situations where each provides a clearer picture than the other.
+The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \RR, or outbreak ``strength'' -- below 1.
+The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \RR\, or even overlooked entirely \citep{park2020reconciling}.
+We argue that \RR\ and $r$ provide equally valid, complementary perspectives on epidemic control, and that there are situations where each provides a clearer picture than the other.
 
 In this study, we:
-first extended the standard paradigm of \Rx\ as critical parameter for control, by defining the strength of an intervention on the same scale as \Rx, the strength of the epidemic; 
+first extended the standard paradigm of \RR\ as critical parameter for control, by defining the strength of an intervention on the same scale as \RR, the strength of the epidemic; 
 then constructed a parallel interpretation which measures the speed of an intervention on the same scale as $r$, the speed of an epidemic.
-We thus showed that the standard paradigm for \Rx\ and control has a natural parallel interpretation in terms of $r$.
+We thus showed that the standard paradigm for \RR\ and control has a natural parallel interpretation in terms of $r$.
 
@@
 we can control an outbreak if we can identify infected individuals and enroll them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the estimates of the amount of early transmission.
-On the other hand, the original explanation of the result relied on a strength-based arugment: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.
+The original explanation of the result relied on a strength-based arugment: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.
 
-While both strength- and speed-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other, depending on the situation.
-Whether strength- or speed-based perspective is more appropriate depends on the disease as well as the intervention strategies.
-For newly invading pathogens, we expect the speed-based perspective to be almost always better.
-When an epidemic is growing exponentially, the basic reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
+While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other.
+Which is more appropriate depends on disease dynamics as well as intervention strategies.
+We expect the speed-based framework to be better at characterizing newly invading pathogens and evaluating intervention strategies that seek to slow down transmission at the individual-level:
+when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling},
+whereas the exponential growth rate can be directly measured from incidence curves \citep{ma2014estimating}.
+On the other hand, we expect the strength-based framework to be better for characterizing endemic diseases and evaluating intervention strategies that seek to reduce the overall transmission at the population-level:
+(i) the endemic equilibrium is determined by the reproductive number and (ii) the long-term outcomes of vaccine-like interventions are typically independent of how fast the intervention is implemented.
+There are also cases where both speed- and strength-based perspectives are equally important:
+controlling an outbreak via active-vaccination requires rapid distribution (speed-like) of efficacious vaccine (strength-like).
 

### 2020-02-26 doc/speedstrength.tex
Rewrite Discussion and Abstract

@@
 
-While an epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproductive number \RR), the effectiveness of an intervention strategy is often, if not always, determined by its ability to reduce \RR\ below 1.
-Here, we generalize this idea by measuring the speed and strength of intervention on the same scale as the speed and strength of an epidemic, respectively.
-We show that an epidemic can be contained if the speed or strength of intervention is greater than their corresponding counterparts.
-Finally, we compare two intervention strategies for HIV -- namely condom intervention and test-and-treat intervention -- to demonstrate that both speed- and strength-based perspectives can be useful for evaluating intervention strategies under different scenarios.
-We suggest that disease modelers should refrain from putting too much emphasis on the reproductive number but rather consider both speed- and strength-based perspectives.
+Epidemics can be characterized by speed (i.e., the exponential growth rate $r$) or strength (i.e., the reproductive number \RR).
+Disease modelers have historically placed much more emphasis in strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
+Here we develop a mathematical framework for this classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
+In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
+We argue that, just as the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
+As an example, we show that evaluating the prospects of test-and-treat interventions against HIV can be done more clearly on the strength than speed scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
+We suggest that disease modelers should refrain from putting too much emphasis on the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary approaches.
 
@@
 
-Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
+Here, we show that $r$ can also serve as a threshold, and also provide a useful metric for difficulty of elimination.
+We first generalize the idea that \RR\ measures the difficulty of elmination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number.
 We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed.
@@
 The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
-We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
+We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (to roughly match the wide generation-interval distribution of HIV \citep{fraser2004factors}).
 Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.
@@
 \caption{
-\textbf{Understanding condom intervention using strength-based decomposition.}
+\textbf{Evaluating a condom intervention using strength-based decomposition.}
 (A) Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
@@
 \caption{
-\textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
+\textbf{Evaluating a test-and-treat intervention using strength- and speed-based decomposition.}
 (A) The strength of the test-and-treat intervention (calculated from the assumed hazard, (C)).
@@
 Thus, the proportion continuing to transmit declines exponentially, and the effect of intervention $\tsub{L}{test}$ increases exponentially.
-However, we assume removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
+However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
 (\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
@@
 
-While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, one can provide a clearer interpretation of the outcome than the other.
-Which is more appropriate depends on disease dynamics as well as intervention strategies.
-We expect the speed-based framework to be better at characterizing newly invading pathogens and evaluating intervention strategies that seek to slow down transmission at the individual-level:
-when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling},
-whereas the exponential growth rate can be directly measured from incidence curves \citep{ma2014estimating}.
-On the other hand, we expect the strength-based framework to be better for characterizing endemic diseases and evaluating intervention strategies that seek to reduce the overall transmission at the population-level:
-(i) the endemic equilibrium is determined by the reproductive number and (ii) the long-term outcomes of vaccine-like interventions are typically independent of how fast the intervention is implemented.
-There are also cases where both speed- and strength-based perspectives are equally important:
-controlling an outbreak via active-vaccination requires rapid distribution (speed-like) of efficacious vaccine (strength-like).
+While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of the outcome than the other.
+For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling},
+Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
+For interventions, we expect the strength-based framework to be clearer for evaluating intervention strategies that seek to reduce the overall transmission at the population level, while the speed-based framework may be clearer for interventions that work at the individual level, like test-and-treat for HIV, or contact-tracing and quarantine for COVID-19.
+In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage.
 
-The contrast between speed- and strength-based perspectives can be found in other literature as well.
-For example, whether the instantaneous per capita rate of increase $r$ (speed) or expected life time reproductive success $R_0$ (strength) is the correct measures of fitness depends on the mechanism of density dependence \citep{pasztor1996r0}.
-When population is regulated by affecting all individuals identically, $r$ is a valid measure \citep{pasztor1996r0}.
-When population is regulated by affecting density-dependent juvenile mortality, $R_0$ is a valid measure \citep{mylius1995evolutionarily}.
-Even though the contrast between the strength- and speed-based perspectives was formalized in theoretical ecology more than 20 years ago, it is still rare in theoretical epidemiology.
+Some practical questions lend themselves more naturally to one approach or another.
+For example, in the classic case of vaccination to eliminate a previously established childhood disease, both disease spread and intervention can be clearly characterized using strength.
+In our HIV example, both the HIV epidemic and the test-and-treat intervention can be clearly characterized using speed.
+Other cases, such as social distancing (strength-like) in the early stages of COVID-19 (clear speed) may not fit so neatly into either paradigm, however.
 
-Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak clearly demonstrate the lack of emphasis on the speed of an epidemic.
-During the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
+There is an analogy here with measures of fitness in theoretical ecology.
+For example, when a population is regulated by density dependence that affects all individuals identically, $r$ may be the best measure of fitness \citep{pasztor1996r0}, 
+but when regulation primarily affects juveniles mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily}.
+The importance of speed-based perspectives are still rarely recognized in the case of infectious disease, however.
+
+Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak show an over-emphasis on strength at the expense of speed:   
+during the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
 For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
-We suggest that infectious disease modelers should be aware of the advantages (and complementarity) of these two frameworks when analyzing disease outbreaks.
+We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.
 

### 2020-02-27 doc/speedstrength.tex
edits

@@
 
-Epidemics can be characterized by speed (i.e., the exponential growth rate $r$) or strength (i.e., the reproductive number \RR).
+An epidemic can be characterized by speed (i.e., the exponential growth rate $r$) or strength (i.e., the reproductive number \RR).
 Disease modelers have historically placed much more emphasis in strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
@@
 We argue that, just as the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
-As an example, we show that evaluating the prospects of test-and-treat interventions against HIV can be done more clearly on the strength than speed scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
+\swp{We argue that the speed-based paradigm provides the clearest insight into certain questionsm just as the strength-based paradigm provides view in other cases.}
+As an example, we show that evaluating the prospects of ``test-and-treat'' interventions \swp{(i.e., identifying infected individuals and putting them to clinical care, such as antiretroviral therapy)} against the human immunodeficiency virus (HIV) can be done more clearly on the strength than speed scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
+\swp{you mean speed than strength, right?}
 We suggest that disease modelers should refrain from putting too much emphasis on the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary approaches.
@@
 
-On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy (ART) \citep{garnett2009treating, nah2017test}.
-We assume that the rate of identification and removal is initially constant.
-Thus, the proportion continuing to transmit declines exponentially, and the effect of intervention $\tsub{L}{test}$ increases exponentially.
+On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy (ART) \citep{garnett2009treating, granich2009universal, nah2017test}.
+We assume that the rate of identification and removal is initially constant \swp{is approximately constant in the beginning}. 
+Thus, the proportion continuing to transmit declines \swp{approximately} exponentially, and the effect of intervention $\tsub{L}{test}$ increases \swp{approximately} exponentially.
 However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
 (\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
-Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength.
+Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength (\figref{test}B).
 In this scenario, we predict that the intervention remains effective over the range of considered parameters.
@@
 The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because sexually active individuals are the most likely to seek testing);
-the assumed hazard rate goes down after a few months to reflect the saturation assumptions discussed above. 
+the assumed hazard rate goes down after a few months to reflect the saturation assumptions discussed above (\figref{test}C). 
 The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
 The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
-The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important (\figref{test}D).
 This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
@@
 
+\swp{can we combine the first two paragraphs into one paragraph? The second paragraph is just 2 sentences long...}
 The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \RR, or outbreak ``strength'' -- below 1.
@@
 Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
-For interventions, we expect the strength-based framework to be clearer for evaluating intervention strategies that seek to reduce the overall transmission at the population level, while the speed-based framework may be clearer for interventions that work at the individual level, like test-and-treat for HIV, or contact-tracing and quarantine for COVID-19.
-In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage.
+For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universa}, or contact-tracing and quarantine for COVID-19; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level.
+\swp{I switched the orders between speed and strength to remain consistent with the way previous sentences are organized.}
+In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.
 
 Some practical questions lend themselves more naturally to one approach or another.
-For example, in the classic case of vaccination to eliminate a previously established childhood disease, both disease spread and intervention can be clearly characterized using strength.
+For example, in the classic case of vaccination to eliminate a previously established childhood disease, both disease spread and intervention can be clearly characterized using strength \citep{anderson1985vaccination}.
 In our HIV example, both the HIV epidemic and the test-and-treat intervention can be clearly characterized using speed.

### 2020-02-27 doc/speedstrength.tex
format

@@
 
-\maketitle
+\begin{flushleft}{
+	\Large
+	\textbf\newline{
+		Speed and strength of an epidemic intervention
+	}
+}
+\newline
+\\
+Jonathan Dushoff\textsuperscript{1,2,3,*}
+Sang Woo Park\textsuperscript{4}
+\\
+\bigskip
+\textbf{1} Department of Biology, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{2} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{3} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{4} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
+\\
+\bigskip
+
+*Corresponding author: dushoff@mcmaster.ca
+\end{flushleft}
 
@@
 We suggest that disease modelers should refrain from putting too much emphasis on the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary approaches.
+
+\pagebreak
 

### 2020-02-28 doc/speedstrength.tex
Need to sleep

@@
 \swp{We argue that the speed-based paradigm provides the clearest insight into certain questionsm just as the strength-based paradigm provides view in other cases.}
-As an example, we show that evaluating the prospects of ``test-and-treat'' interventions \swp{(i.e., identifying infected individuals and putting them to clinical care, such as antiretroviral therapy)} against the human immunodeficiency virus (HIV) can be done more clearly on the strength than speed scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
-\swp{you mean speed than strength, right?}
+As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
 We suggest that disease modelers should refrain from putting too much emphasis on the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary approaches.
@@
 On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy (ART) \citep{garnett2009treating, granich2009universal, nah2017test}.
-We assume that the rate of identification and removal is initially constant \swp{is approximately constant in the beginning}. 
-Thus, the proportion continuing to transmit declines \swp{approximately} exponentially, and the effect of intervention $\tsub{L}{test}$ increases \swp{approximately} exponentially.
+We assume that the rate of identification and removal is approximately constant for newly infected individuals; thus, the proportion continuing to transmit declines approximately exponentially.
 However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment

### 2020-03-02 doc/speedstrength.tex
Still working on HIV explanation, will try to finish tomorrow morning

@@
 We argue that, just as the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
-\swp{We argue that the speed-based paradigm provides the clearest insight into certain questionsm just as the strength-based paradigm provides view in other cases.}
 As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
-We suggest that disease modelers should refrain from putting too much emphasis on the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary approaches.
+We suggest that disease modelers should avoid over-emphasizing the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary measures.
 
@@
 
+%% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.
+
 \begin{figure}[!t]
@@
 
-On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy (ART) \citep{garnett2009treating, granich2009universal, nah2017test}.
-We assume that the rate of identification and removal is approximately constant for newly infected individuals; thus, the proportion continuing to transmit declines approximately exponentially.
+Next, we consider a ``test-and-treat'' strategy in which infected individuals are identified, linked to care and receive antiretroviral therapy (ART) with the goal of both preserving health and preventing transmission through viral suppression.
+\citep{garnett2009treating, granich2009universal, nah2017test}.
+Our assumptions for this scenario are shown in \figref{test}.
+We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but increases very quickly (because sexually active individuals are the most likely to seek testing);
+after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment \figref{test}(C).
+The corresponding strength of intervention $L$ is shown in \figref{test}(A) and details of the assumption are given in the caption.
+
+In this example, we see that the
+
+The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
+The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
+The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important (\figref{test}D).
+This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
+
 However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
@@
 Using the speed-based decomposition provides a valuable perspective on this situation.
-The hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but peaks very fast (because sexually active individuals are the most likely to seek testing);
 the assumed hazard rate goes down after a few months to reflect the saturation assumptions discussed above (\figref{test}C). 

### 2020-03-02 doc/speedstrength.tex
Add the emailed comment
Ready for medR?

@@
 
-\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
+\newcommand{\comment}{}
+\renewcommand{\comment}[3]{}
+\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
+
 \newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
@@
 
-%% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.
+\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
 
@@
 Our assumptions for this scenario are shown in \figref{test}.
-We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but increases very quickly (because sexually active individuals are the most likely to seek testing);
+We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because there is no way for newly infected individuals to know that they have HIV) but increases very quickly (because sexually active individuals are the most likely to seek testing);
 after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment \figref{test}(C).
@@
 
-In this example, we see that the
+In this example, we see that, as \pEarly\ goes down and our estimate of epidemic strength increases, the estimate of intervention strength increases roughly in parallel.
+The increase in intervention strength makes sense, as we have more time to reach people on average before they transmit: this is the core of the result of \cite{eaton2014proportion}.
+In our scenario, we predict that the intervention remains effective over the range of considered parameters.
 
+\jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
+
+Though there is a clear intuition for why both strengths increase as early transmission goes down, the speed paradigm provides insight into why these two increases are so close to parallel.
 The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
-The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
-The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important (\figref{test}D).
-This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
+For the test-and-treat intervention, the effective epidemic speed also stays relatively constant (\figref{test}D), in part because we have (plausibly) assumed that hazard stays relatively constant for a few key months, and in part because the backward generation-interval distributions for different  scenarios are relatively similar.
+The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+Thus, the speed paradigm provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission.
 
@@
 Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength (\figref{test}B).
-In this scenario, we predict that the intervention remains effective over the range of considered parameters.
-
-Using the speed-based decomposition provides a valuable perspective on this situation.
-the assumed hazard rate goes down after a few months to reflect the saturation assumptions discussed above (\figref{test}C). 
-The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
-The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
-The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important (\figref{test}D).
-This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
 
@@
 
-\swp{can we combine the first two paragraphs into one paragraph? The second paragraph is just 2 sentences long...}
 The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \RR, or outbreak ``strength'' -- below 1.
@@
 Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
-For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universa}, or contact-tracing and quarantine for COVID-19; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level.
-\swp{I switched the orders between speed and strength to remain consistent with the way previous sentences are organized.}
+For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universa}, or contact-tracing and quarantine for COVID-19; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level, like condom use.
 In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.

### 2020-03-02 doc/speedstrength.tex
very minor tweaks

@@
 
-An epidemic can be characterized by speed (i.e., the exponential growth rate $r$) or strength (i.e., the reproductive number \RR).
+An epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproductive number \RR).
 Disease modelers have historically placed much more emphasis in strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
-Here we develop a mathematical framework for this classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
+Here, we develop a mathematical framework for this classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
 In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
@@
 
-\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
+% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
 
@@
 We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because there is no way for newly infected individuals to know that they have HIV) but increases very quickly (because sexually active individuals are the most likely to seek testing);
-after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment \figref{test}(C).
-The corresponding strength of intervention $L$ is shown in \figref{test}(A) and details of the assumption are given in the caption.
+after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment (\figref{test}C).
+The corresponding strength of intervention $L$ is shown in \figref{test}A and details of the assumption are given in the caption.
 
@@
 
-\jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
+% \jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
 
@@
 Thus, the speed paradigm provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission.
-
-However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
-(\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
-Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength (\figref{test}B).
 
@@
 While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of the outcome than the other.
-For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling},
+For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
 Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
-For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universa}, or contact-tracing and quarantine for COVID-19; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level, like condom use.
+For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level, like condom use.
 In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.

### 2020-03-02 doc/speedstrength.tex
Clarify a bit; not necessarily sharp.

@@
 \newcommand{\comment}{}
+\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
 \renewcommand{\comment}[3]{}
-\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
 
@@
 
-% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
+\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
 
@@
 
-% \jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
+\jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
 
@@
 we can control an outbreak if we can identify infected individuals and enroll them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the estimates of the amount of early transmission.
-The original explanation of the result relied on a strength-based arugment: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.
+The original explanation of the result relied on a strength-based argument: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.
 
-While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of the outcome than the other.
+While both speed- and strength-based frameworks can give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of a given measure.
 For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
@@
 
-Some practical questions lend themselves more naturally to one approach or another.
+When comparing interventions with epidemic parameters to evaluate strategies, the situation is similar.
+Some scenarios lend themselves naturally to a single approach.
 For example, in the classic case of vaccination to eliminate a previously established childhood disease, both disease spread and intervention can be clearly characterized using strength \citep{anderson1985vaccination}.
-In our HIV example, both the HIV epidemic and the test-and-treat intervention can be clearly characterized using speed.
-Other cases, such as social distancing (strength-like) in the early stages of COVID-19 (clear speed) may not fit so neatly into either paradigm, however.
+In our HIV example, both the HIV epidemic and the test-and-treat intervention can be best characterized using speed.
+Other cases, such as using social distancing (a strength-like intervention) in the early stages of COVID-19 (epidemic speed is observed) may not fit so neatly into either paradigm, however.
 

### 2020-03-03 doc/speedstrength.tex
weird spacing issues here and there
minor typos

@@
 An epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproductive number \RR).
-Disease modelers have historically placed much more emphasis in strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
+Disease modelers have historically placed much more emphasis on strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
 Here, we develop a mathematical framework for this classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
@@
 Here, we show that $r$ can also serve as a threshold, and also provide a useful metric for difficulty of elimination.
-We first generalize the idea that \RR\ measures the difficulty of elmination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number.
+We first generalize the idea that \RR\ measures the difficulty of elimination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number.
 We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed.
@@
 
-We compare two different possible intervention strategies to shed light on the speed and strenght decompositions.
+We compare two different possible intervention strategies to shed light on the speed and strength decompositions.
 First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.

### 2020-06-08 doc/speedstrength.tex
Autocommit (doc)
One more issue

@@
 Disease modelers have historically placed much more emphasis on strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
-Here, we develop a mathematical framework for this classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
+Here, we develop a mathematical framework for the classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
 In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
-We argue that, just as the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
+We argue that, while the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
 As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
@@
 
-Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population, called the \emph{basic} reproductive number \Ro, because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
+Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population -- called the \emph{basic} reproductive number \Ro, because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
 The insight that a case must on average cause at least one new case under good conditions for a disease to persist goes back $>100$ years \citep{ross1911prevention};
@@
 Thus, we will assume $K(\tau, t) \equiv K(\tau)$.
-Importantly, this means we are neglecting changes in susceptible proportion through time: $S(t) \approx S(0)$.
+Importantly, this means we are neglecting changes in susceptible proportion through time: that is, assuming that $S(t) \approx S(0)$.
 Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
@@
 \end{equation}
-Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease) is exactly $\theta=\RR$.
+Since $g$ integrates to 1, the reduction needed to prevent invasion (or to eliminate disease) is exactly $\theta=\RR$.
 We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread $\RR$.
@@
 The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
-For the test-and-treat intervention, the effective epidemic speed also stays relatively constant (\figref{test}D), in part because we have (plausibly) assumed that hazard stays relatively constant for a few key months, and in part because the backward generation-interval distributions for different scenarios are relatively similar.
-The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+For the test-and-treat intervention, the effective intervention speed also stays relatively constant (\figref{test}D), in part because we have (plausibly) assumed that hazard stays relatively constant for a few key months, and in part because the backward generation-interval distributions for different scenarios are relatively similar.
+The effective intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
+\jd{Above s.~does not match the picture. Is this something we can figure out? Or should we just drop the s.?}
 Thus, the speed paradigm provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission.
@@
 To illustrate this idea, we used simple assumptions to explore the effects of two HIV intervention strategies (condoms and test-and-treat), using both strength- and speed-based frameworks.
-In particular, we provided an alternative explanation for the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission does not affect the effectiveness of the ART:
+In particular, we provided an alternative explanation for the result of \cite{eaton2014proportion} who used detailed mathematical modeling of HIV transmission to show that the amount of early transmission has little effect on predicted effectiveness of a test and treat intervention:
 we can control an outbreak if we can identify infected individuals and enroll them on ART faster than the \emph{observed} rate at which new cases are generated, which does not depend on the estimates of the amount of early transmission.

### 2020-06-08 doc/speedstrength.tex
All the way through, pass to Park

@@
 The effective intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
-\jd{Above s.~does not match the picture. Is this something we can figure out? Or should we just drop the s.?}
 Thus, the speed paradigm provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission.
@@
 The original explanation of the result relied on a strength-based argument: increasing the amount of early transmission decreases the basic reproductive number, which negatively correlates with the outcome of the ART intervention \citep{eaton2014proportion}.
+The speed paradigm provides an additional insight: since we expect more early transmission to make our estimate of intervention speed (a little) faster, higher amounts of early transmission (when controlling for the observed initial growth rate) are expected to make control via test-and-treat (a little) easier.
 

### 2020-06-08 figure/condom.R
Try to clean; condom.pdf out of repo

@@ -7,19 +7,20 @@ source("../R/HIVfuns.R")
 source("../R/kernel.R")
 
 h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-yintercept <- 4 ## arbitrary value
+yintercept <- 4 ## Assumption about condom effectiveness
 
 lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
 
 g1 <- ggplot() +
 	geom_hline(yintercept=yintercept, col=2) +
-	scale_x_continuous("Time (years)") +
+	scale_x_continuous("Time since infection") +
 	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
 	ggtitle("A") +
 	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line.y = element_line()
+		panel.grid = element_blank()
+		, panel.border = element_blank()
+		, axis.line.y = element_line()
+		## , axis.line.x = element_line()
 	)
 
 early <- seq(from=0.1, to=0.4, by=0.01)

### 2020-06-15 doc/speedstrength.tex
minor edits

@@
 
-Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population -- called the \emph{basic} reproductive number \Ro, because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
+Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population -- called the \emph{basic} reproductive number \Ro\, --- because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
 The insight that a case must on average cause at least one new case under good conditions for a disease to persist goes back $>100$ years \citep{ross1911prevention};
@@
 
-\section{Methods}
+\section{Mathematical theory}
 
@@
 Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
-In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible, $S(t)$.
+In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ --- notably, the proportion of the population susceptible, $S(t)$.
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time.
@@
 therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
-Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\RR$ is thus the integral of $K$.
+Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\Ro$ is thus the integral of $K$.
 
@@
 \begin{equation}
-	\Rhat = \int \hat K(\tau) d\tau
+	\Rhat = \int \hat K(\tau) d\tau.
 \end{equation}
@@
 We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$.
-It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
+It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by the intrinsic generation-interval distribution:
 \begin{equation}
@@
 The infection kernel is shown in (\figref{example}A) for our baseline value of  
-$\tsub{p}{early} = 0.23$. We assume that the initial speed of the epidemic is $r=0.452\,\mathrm{year}^{-1}$(\figref{example}B), and ask what value of $\Ro$ would produce this rate of growth.
-When transmission is fast, (i.e., when \pEarly\ is large), individuals don't need to transmit as much to achieve this speed, so the estimated value of \Ro\ decreases (\figref{example}C).
+$\tsub{p}{early} = 0.23$. We assume that the initial speed of the epidemic is $r=0.452\,\mathrm{year}^{-1}$(\figref{example}B), and ask what value of $\RR$ would produce this rate of growth.
+When transmission is fast, (i.e., when \pEarly\ is large), individuals don't need to transmit as much to achieve this speed, so the estimated value of \RR\ decreases (\figref{example}C).
 Therefore, as \pEarly\ gets smaller, we expect stronger intervention to be required in order to control the disease.
@@
 Assuming that condoms act as a physical barrier, and that condom use will, on average, remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$ (\figref{condom}A).
-The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$, i.e., $\theta=4$, whereas the estimated strength of the epidemic decreases as the proportion of early transmission increases (\figref{condom}B).
-Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
+The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$, i.e., $\theta=4$, whereas the estimated strength of the epidemic \RR\ decreases as the proportion of early transmission \pEarly\ increases (\figref{condom}B).
+Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if the amount of early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
 
@@
 after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment (\figref{test}C).
-The corresponding strength of intervention $L$ is shown in \figref{test}A and details of the assumption are given in the caption.
+The corresponding strength of intervention $\tsub{L}{test}$ is shown in \figref{test}A and details of the assumption are given in the caption.
 
@@
 
-The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \RR, or outbreak ``strength'' -- below 1.
-The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \RR\, or even overlooked entirely \citep{park2020reconciling}.
+The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number --- \RR, or outbreak ``strength'' --- below 1.
+The exponential growth rate --- $r$, or outbreak ``speed'' --- is often seen just as a stepping stone to \RR\, or even overlooked entirely \citep{park2020reconciling}.
 We argue that \RR\ and $r$ provide equally valid, complementary perspectives on epidemic control, and that there are situations where each provides a clearer picture than the other.

### 2020-06-17 doc/speedstrength.tex
good night

@@
 The speed of an epidemic is characterized by the exponential growth rate $r$, which measures how \emph{fast} an epidemic grows at the population level.
-The strength of an epidemic is characterized by the reproductive number \RR, which measures how many new cases are caused by a typical \emph{individual} case.
+The strength of an epidemic is characterized by the reproductive number \RR, which measures \emph{how many} new cases are caused by a typical individual case.
 Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

### 2020-06-19 doc/speedstrength.tex
added one citation
speed

@@
 
+\section*{Significance statement}
+
+The exponential growth rate $r$ (speed) and the reproductive number (strength) of an epidemic $\RR$ represent the speed and strength of an epidemic.
+Even though both quantities have equally valid threshold values ($r=0$ and $\RR=1$) that determine whether a disease can spread in a population, disease modelers often measure the effectiveness of intervention by its ability to reduce $\RR$ below 1.
+We generalize the classic, strength-based paradigm by defining the speed and strength of an intervention, which must be greater than their corresponding values of an epidemic to prevent its spread.
+Both speed- and strength-based perspective can yield important, but different, biological insights into understanding the disease spread.
+
 \pagebreak
@@
 
-Here, we show that $r$ can also serve as a threshold, and also provide a useful metric for difficulty of elimination.
+Here, we show that $r$ can also serve as a threshold, and also provide a useful metric for difficulty of elimination (cf.
+\cite{ferretti2020quantifying}).
 We first generalize the idea that \RR\ measures the difficulty of elimination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number.

### 2020-06-20 doc/speedstrength.tex
Edit significance

--- 2020-06-17 doc/speedstrength.tex
@@
 \documentclass[12pt]{article}
+\usepackage[utf8]{inputenc}
 \usepackage[]{graphicx}
@@
 
+\section*{Significance statement}
+
+Researchers studying infectious-disease outbreaks often focus primarily on the reproductive number, which measures the “strength” of transmission – the number of new infections caused by an infectious individual.
+Much less theoretical and modeling attention is paid to the exponential growth rate, which measures epidemic “speed”. 
+We argue that this difference in emphasis is largely artificial: both strength and speed have thresholds for spread, and both provide frameworks to compare potential for spread to effectiveness of intervention.
+We generalize the classic strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
+We provide specific examples from important diseases, including rabies and HIV, to demonstrate the value of both perspectives.
+
 \pagebreak

### 2020-06-20 doc/abstract.pl
Automatically produce head material text for submission

@@ -1,0 +1,14 @@
+use strict;
+use 5.10.0;
+
+while(<>){
+	if (/section.*Abstract/ .. /pagebreak/){
+		next if/^\\/;
+		s/\$//g;
+		s/\\//g;
+		s/RR/R/g;
+		s/``/“/g;
+		s/''/”/g;
+		print;
+	}
+}

### 2020-06-20 doc/speedstrength.tex
done?

--- 2020-06-17 doc/speedstrength.tex
@@
 
+\section*{Significance statement}
+
+Researchers studying infectious-disease outbreaks often focus primarily on the reproductive number, which measures the ``strength'' of transmission --- the number of new infections caused by an infected individual.
+Less theoretical and modeling attention is paid to the exponential growth rate, which measures epidemic ``speed''. 
+We argue that this difference in emphasis is largely artificial: both strength and speed have thresholds for spread and provide frameworks to compare effectiveness of intervention.
+We generalize the classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
+We provide specific examples from Human immunodeficiency virus (HIV) to demonstrate the value of both perspectives.
+
 \pagebreak

### 2020-06-22 doc/abstract.pl
good nght

@@ -1,14 +1,18 @@
 use strict;
 use 5.10.0;
 
+$/ = "";
+
 while(<>){
 	if (/section.*Abstract/ .. /pagebreak/){
 		next if/^\\/;
+		s/\n/ /g;
 		s/\$//g;
 		s/\\//g;
 		s/RR/R/g;
 		s/``/“/g;
 		s/''/”/g;
-		print;
+		s/---?/–/g;
+		say;
 	}
 }

### 2020-06-27 doc/speedstrength.tex
merge with Daniel significance
good nght
move out significance

--- 2020-06-19 doc/speedstrength.tex
@@
 \usepackage{hyperref}
-\bibliographystyle{chicago}
+\bibliographystyle{plain}
 \date{\today}
@@
 
-\newcommand{\comment}{}
+\newcommand{\comment}{RENEW}
 \renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
@@
 We suggest that disease modelers should avoid over-emphasizing the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary measures.
-
-\section*{Significance statement}
-
-The exponential growth rate $r$ (speed) and the reproductive number (strength) of an epidemic $\RR$ represent the speed and strength of an epidemic.
-Even though both quantities have equally valid threshold values ($r=0$ and $\RR=1$) that determine whether a disease can spread in a population, disease modelers often measure the effectiveness of intervention by its ability to reduce $\RR$ below 1.
-We generalize the classic, strength-based paradigm by defining the speed and strength of an intervention, which must be greater than their corresponding values of an epidemic to prevent its spread.
-Both speed- and strength-based perspective can yield important, but different, biological insights into understanding the disease spread.
 
@@
 
-\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
+%% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
 
@@
 
-\jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
+%% \jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
 

### 2020-09-21 doc/speedstrength.tex
End of weekend ☹

@@
 \newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
+
+\newcommand{\rr}{\ensuremath{{r}}}
+\newcommand{\rhat}{\ensuremath{{\hat\rr}}}
+
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
@@
 
-Here, we show that $r$ can also serve as a threshold, and also provide a useful metric for difficulty of elimination (cf.
+Here, we argue that the dominance of \RR\ over $r$ in the disease-dynamics literature is excessive: $r$ has been under-rated as a metric.
+Like \RR\,  $r$ can also serve as a threshold, and we show that it can also provide a useful metric for difficulty of elimination (cf.
 \cite{ferretti2020quantifying}).
@@
 We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed.
-We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
+We argue that the primacy of $\RR$ over $r$ is partly due to history, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
 
@@
 \end{equation}
-This framework generalizes the work of \cite{fraser2004factors} who made parametric assumptions about the shape of $L(\tau)$.
+This framework generalizes the work of \cite{fraser2004factors}, who made parametric assumptions about the shape of $L(\tau)$.
 
@@
 \end{equation}
-In this more general case, we have again that the disease cannot spread when $\theta \geq \RR$.
+
+Again, we have that the disease cannot spread when $\theta \geq \RR$.
+In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, and the control measure must be at least as ``strong'' as the disease (as measured by \RR) to achieve elimination.
 
 \subsection{Speed-based decomposition}
+
+The above decomposition generalizes the argument that \RR\ is the key parameter in evaluating whether a disease can be controlled -- one of the main foundations of historical primacy of \RR.
+But we can in fact do an analogous decomposition based on speed, and place \rr\ in a similar role.
 
@@
 K(\tau) = b(\tau)\exp(r\tau)
+\eqlab{Kback}
 \end{equation}
@@
 
-Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation (\figref{constant}B), for example, by identifying and isolating infectious individuals.
+Now imagine an idealized intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation (\figref{constant}B), for example, by identifying and isolating infectious individuals.
 We then have:
+\begin{equation}
+	\hat K(\tau) = K(\tau)\exp(-\phi\tau)
+\end{equation}
+The interpretation is that average infectiousness for under this control regime is the product of the original infectiousness $K(\tau)$ (at age of infection $\tau$) and the probability $\exp(-\phi\tau)$ of escaping the hazard of control up to that time.
+
+Substituting \eqref{Kback}:
 \begin{equation}
@@
 
-%% \jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}
-
 Though there is a clear intuition for why both strengths increase as early transmission goes down, the speed paradigm provides insight into why these two increases are so close to parallel.
@@
 Thus, the speed paradigm provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission.
+
+\section{Other examples}
 

### 2020-09-22 doc/speedstrength.tex
some progress?

@@
 Here, we develop a mathematical framework for the classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
-In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
+In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$ \cite{dhmKermack}, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
 We argue that, while the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.

### 2020-10-01 doc/speedstrength.tex
good night

@@
 \end{equation}
-where $L(\tau)$ is the average proportional reduction for an individual infected time $\tau$ ago.
+where $L(\tau)$ is the average multiplicative reduction for an individual infected time $\tau$ ago.
 The post-intervention reproductive number is thus:
@@
 First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
-Assuming that condoms act as a physical barrier, and that condom use will, on average, remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$ (\figref{condom}A).
+Assuming that condoms act as a physical barrier, and that condom use will, on average, remain roughly constant through time, it is reasonable to model the multiplicative reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$ (\figref{condom}A).
 The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$, i.e., $\theta=4$, whereas the estimated strength of the epidemic \RR\ decreases as the proportion of early transmission \pEarly\ increases (\figref{condom}B).
@@
 \textbf{Evaluating a condom intervention using strength-based decomposition.}
-(A) Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
+(A) Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the multiplicative reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
 (B) The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
@@
 
+We reiterate that a complete calculation using the same assumptions under either paradigm will necessarily provide consistent answers.
+But in this particular case, the speed paradigm provides an answer whose causes are easier to understand.
+We argue that it is therefore easier to assess the assumptions underlying the necessarily incomplete assumptions that are made in reality.
+
 \section{Other examples}
+
+\subsection{COVID-19}
+
+There has been a great deal of discussion of the importance of pre-symptomatic  transmission of COVID-19.
+Pre-symptomatic transmission is likely to be hard to detect, and therefore hard to prevent.
+Thus, it might be supposed that an increase in the estimated importance of pre-symptomatic transmission would lead to an increase in the estimated difficulty
 
@@
 For example, when a population is regulated by density dependence that affects all individuals identically, $r$ may be the best measure of fitness \citep{pasztor1996r0}, 
-but when regulation primarily affects juveniles mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily}.
+but when regulation primarily affects juvenile mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily}.
+
 The importance of speed-based perspectives are still rarely recognized in the case of infectious disease, however. 
-
 Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak show an over-emphasis on strength at the expense of speed:   

### 2020-10-01 doc/speedstrength.tex
Working on revisions

@@
 Pre-symptomatic transmission is likely to be hard to detect, and therefore hard to prevent.
-Thus, it might be supposed that an increase in the estimated importance of pre-symptomatic transmission would lead to an increase in the estimated difficulty
+Thus, it might be supposed that an increase in the estimated importance of pre-symptomatic transmission would lead to an increase in the estimated difficulty.
+
+A generation-interval perspective can be used to challenge this view.
+More early transmission, for a given value of \rr, means less transmission per individual -- that is, a lower value of \RR.
+Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary.
+
+Given sufficient assumptions, a strength-based calculation could provide an answer for any particular scenario.
+But using the speed-based paradigm substantially clarifies the situation.
+As with the HIV example, our estimate of the speed of the epidemic is not changed by information about timing of transmission.
+Thus, the question is reduced to whether (and how) we expect pre-symptomatic transmission to affect the effective speed of intervention.
+
+Case detection is based on contact tracing, testing and symptoms.
+Thus, we don't expect pre-symptomatic infection to change the intervention hazard $h$.
+The effective speed, however, is a (sort-of) mean taken over the (backwards) generation interval.
+If the
 

### 2020-10-01 figure/HIVexample.R
good night

@@ -1,0 +1,73 @@
+
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+sourceFiles()
+makeGraphics(width=8, height=3)
+
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+g1 <- ggplot(h_base) +
+	geom_line(aes(time/12, density)) +
+	scale_x_continuous("Generation time (years)") +
+	scale_y_continuous("Density", expand=c(0, 0)) +
+	ggtitle("A") +
+	theme(
+		legend.position=c(0.9, 0.9),
+		legend.title=element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+coef(lfit)
+
+pp <- predict(lfit, se=TRUE)
+
+g2 <- ggplot(hiv_ts) +
+	geom_point(aes(year, prevalence)) +
+	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
+	scale_x_continuous("Year") +
+	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
+	ggtitle("B") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+r2R(h_base, coef(lfit)[2]/12)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <-  sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+g3 <- ggplot(earlydata) +
+	geom_line(aes(early, R0)) +
+	geom_point(aes(0.23, 3.17), size=5) +
+	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
+	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
+	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
[…]

### 2020-10-01 figure/gencontrol.R

@@ -1,0 +1,62 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+sourceFiles()
+makeGraphics(width=8, height=3)
+
+gbar <- 16.2
+gshape <- 1/0.58^2
+
+tvec <- seq(0, 50, by=0.01)
+
+g1 <- data.frame(
+  time=tvec,
+  density=1.5*dgamma(tvec, shape=gshape, rate=gshape/gbar)
+)
+
+g2 <- data.frame(
+  time=tvec,
+  density=dgamma(tvec, shape=gshape, rate=gshape/gbar)
+)
+
+p1 <- ggplot(data=g1) +
+  geom_line(aes(time, density, col="Baseline")) +
+  geom_line(data=g2, aes(time, density, col="Controlled")) +
+  geom_segment(aes(x=10.75, xend=10.75, y=0.07, yend=0.053),
+               arrow = arrow(length = unit(0.3, "cm"))) +
+  ggtitle("A. Constant strength intervention") +
+  scale_x_continuous("Time (days)") +
+  scale_y_continuous("Density") +
+  scale_color_manual(values=c(1, 2)) +
+  theme(
+    panel.grid = element_blank(),
+    panel.border = element_blank(),
+    axis.line=element_line(),
+    legend.position = c(0.85, 0.9),
+    legend.title = element_blank()
+  )
+
+b0 <- b0fun(g2, R2r(g2, 1.5))
+
+p2 <- ggplot(data=g1) +
+  geom_line(aes(time, density, col="Baseline")) +
+  geom_line(data=b0, aes(time, density, col="Controlled")) +
+  geom_segment(aes(x=10.75, xend=9.39, y=0.07, yend=0.06),
+               arrow = arrow(length = unit(0.3, "cm"))) +
+  ggtitle("B. Constant speed intervention") +
+  scale_x_continuous("Time (days)") +
+  scale_y_continuous("Density") +
+  scale_color_manual(values=c(1, 2)) +
+  theme(
+    panel.grid = element_blank(),
+    panel.border = element_blank(),
+    axis.line=element_line(),
+    legend.position = c(0.85, 0.9),
+    legend.title = element_blank()
+  )
+
+grid.arrange(arrangeGrob(p1, p2, nrow=1))
+

### 2020-10-01 figure/hivstrength.R

@@ -1,0 +1,60 @@
+
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R") 
+
+commandEnvironments() ## Read in any environments specified as dependencies
+sourceFiles()
+makeGraphics(width=8, height=3)
+
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+yintercept <- 4 ## Assumption about condom effectiveness
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+g1 <- ggplot() +
+	geom_hline(yintercept=yintercept, col=2) +
+	scale_x_continuous("Time since infection") +
+	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
+	ggtitle("A") +
+	theme(
+		panel.grid = element_blank()
+		, panel.border = element_blank()
+		, axis.line.y = element_line()
+		## , axis.line.x = element_line()
+	)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+g2 <- ggplot(earlydata) +
+	geom_line(aes(early, R0, col="Epidemic")) +
+	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
+  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
+  geom_point(aes(x=0.23, y=yintercept), size=5, col="red") +
+	scale_color_manual(values=c("black", "red")) +
+	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
+	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +
+	ggtitle("B") +
+	theme(
+		legend.position=c(0.85, 0.9),
+		legend.title = element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+grid.arrange(arrangeGrob(g1, g2, nrow=1))
+

### 2020-10-01 figure/tt.R

@@ -1,0 +1,111 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+sourceFiles()
+makeGraphics(width=8, height=6)
+
+maxRate <- 1/6
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+b_base <- b0fun(h_base, coef(lfit)[2]/12)
+
+tt <- testingFun(maxRate)
+
+g1 <- ggplot(tt) +
+	geom_line(aes(time/12, strength), col=2) +
+	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
+	scale_x_continuous("Time (years)", expand=c(0,0), limits=c(0, 16.5)) +
+	scale_y_log10(expression(L[test])) +
+	ggtitle("A") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+interstrength <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	theFun(hh, tt)
+})
+
+strengthdata <- data.frame(
+	early=early,
+	strength=interstrength
+)
+
+g2 <- ggplot(earlydata) +
+	geom_line(aes(early, R0, col="Epidemic")) +
+	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
+	scale_color_manual(values=c("black", "red")) +
+  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
+  geom_point(data=strengthdata[strengthdata$early==0.23,], aes(early, strength), size=5, col="red") +
+	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
+	scale_y_log10("Strength", limits=c(1, 16), expand=c(0, 0), breaks=c(1, 2, 4, 8, 16)) +
+	ggtitle("B") +
+	theme(
[…]

### 2020-10-01 figure/HIVexample.R
pre-dinner

@@ -70,4 +70,4 @@ g3 <- ggplot(earlydata) +
 		axis.line = element_line()
 	)
 
-grid.arrange(arrangeGrob(g1, g2, g3, nrow=1))
+grid.arrange(g1, g2, g3, nrow=1)

### 2020-10-01 figure/gencontrol.R

@@ -58,5 +58,5 @@ p2 <- ggplot(data=g1) +
     legend.title = element_blank()
   )
 
-grid.arrange(arrangeGrob(p1, p2, nrow=1))
+grid.arrange(p1, p2, nrow=1)
 

### 2020-10-01 figure/hivstrength.R

@@ -56,5 +56,5 @@ g2 <- ggplot(earlydata) +
 		axis.line = element_line()
 	)
 
-grid.arrange(arrangeGrob(g1, g2, nrow=1))
+grid.arrange(g1, g2, nrow=1)
 

### 2020-10-01 figure/tt.R

@@ -107,5 +107,5 @@ g4 <- ggplot(speeddata) +
 		axis.line = element_line()
 	)
 
-grid.arrange(arrangeGrob(g1, g2, g3, g4, nrow=2))
+grid.arrange(g1, g2, g3, g4, nrow=2)
 

### 2020-10-02 doc/speedstrength.tex
pre-dinner
Not really ready to meet Daniel ☺

@@
 \usepackage[utf8]{inputenc}
-\usepackage[]{graphicx}
-\usepackage[]{color}
+\usepackage{graphicx}
+\usepackage{grffile}
+\usepackage{color}
 \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
@@
 Disease modelers have historically placed much more emphasis on strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
-Here, we develop a mathematical framework for the classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
+Here, we develop a mathematical framework for the classic, strength-based paradigm and show that there is a dual speed-based paradigm which can provide complementary insights.
 In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$ \cite{dhmKermack}, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
@@
 As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
+We also discuss evaluating the effects of the importance of pre-symptomatic transmission of the SARS-Cov-2 virus.
 We suggest that disease modelers should avoid over-emphasizing the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary measures.
@@
 Here, we argue that the dominance of \RR\ over $r$ in the disease-dynamics literature is excessive: $r$ has been under-rated as a metric.
-Like \RR\,  $r$ can also serve as a threshold, and we show that it can also provide a useful metric for difficulty of elimination (cf.
-\cite{ferretti2020quantifying}).
+Like \RR,  $r$ can serve as a threshold, and we show that it can also provide a useful metric for difficulty of elimination (cf.~\cite{ferretti2020quantifying}).
 We first generalize the idea that \RR\ measures the difficulty of elimination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number. 
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/constant_intervention.pdf}
+\includegraphics[width=\textwidth]{figure/gencontrol.Rout.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure/HIV.pdf}
+\includegraphics[width=\textwidth]{figure/HIVexample.Rout.pdf}
 \caption{
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/condom.pdf}
+\includegraphics[width=\textwidth]{figure/hivstrength.Rout.pdf}
 \caption{
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/test_and_treat.pdf}
+\includegraphics[width=\textwidth]{figure/tt.Rout.pdf}
 \caption{
@@
 
-\section{Other examples}
-
-\subsection{COVID-19}
+%% \section{Other examples}\subsection{COVID-19}
+\section{Example: COVID-19}
 
@@
 Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary.
+
+For interventions which are SOMETHING, the strength-based perspective gives a clear answer.
+
+Conversely,
 

### 2020-10-04 code/HIVexample.R
Code directory merge

@@ -1,0 +1,73 @@
+
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+sourceFiles()
+makeGraphics(width=8, height=3)
+
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+g1 <- ggplot(h_base) +
+	geom_line(aes(time/12, density)) +
+	scale_x_continuous("Generation time (years)") +
+	scale_y_continuous("Density", expand=c(0, 0)) +
+	ggtitle("A") +
+	theme(
+		legend.position=c(0.9, 0.9),
+		legend.title=element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+coef(lfit)
+
+pp <- predict(lfit, se=TRUE)
+
+g2 <- ggplot(hiv_ts) +
+	geom_point(aes(year, prevalence)) +
+	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
+	scale_x_continuous("Year") +
+	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
+	ggtitle("B") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+r2R(h_base, coef(lfit)[2]/12)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <-  sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+g3 <- ggplot(earlydata) +
+	geom_line(aes(early, R0)) +
+	geom_point(aes(0.23, 3.17), size=5) +
+	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
+	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
+	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
[…]

### 2020-10-04 code/HIVfuns.R

@@ -1,0 +1,32 @@
+# Units are months
+# Units of g are 1/month, so Sum{step*g) should be 1
+
+testingFun <- function(maxRate
+					   , hMean=12, hShape=2, hheight=4
+					   , step = 0.1, window=200
+){
+	time <- seq(step, window, by=step)
+	tHaz0 <- dgamma(time, shape=hShape, scale=hMean/hShape)
+	tHaz <- hheight*tHaz0/max(tHaz0)
+	hazRate <- maxRate*(1-exp(-tHaz))
+	Haz <- step*cumsum(hazRate)
+	return(data.frame(time
+					  , hazRate
+					  , strength = exp(Haz)
+	))
+}
+
+HIVgen <- function(earlyProp 
+				   , earlyMean=3, earlyShape=3
+				   , lateMean=120, lateShape=2
+				   , step = 0.1, window=200
+){
+	time <- seq(step, window, by=step)
+	d0 <- (
+		earlyProp*dgamma(time, shape=earlyShape, scale=earlyMean/earlyShape)
+		+ (1-earlyProp)*dgamma(time, shape=lateShape, scale=lateMean/lateShape)
+	)
+	return(data.frame(time
+					  , density=d0/sum(step*d0)
+	))
+}

### 2020-10-04 code/condom.R

@@ -1,0 +1,60 @@
+
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R") 
+
+commandEnvironments() ## Read in any environments specified as dependencies
+sourceFiles()
+makeGraphics(width=8, height=3)
+
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+yintercept <- 4 ## Assumption about condom effectiveness
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+g1 <- ggplot() +
+	geom_hline(yintercept=yintercept, col=2) +
+	scale_x_continuous("Time since infection") +
+	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
+	ggtitle("A") +
+	theme(
+		panel.grid = element_blank()
+		, panel.border = element_blank()
+		, axis.line.y = element_line()
+		## , axis.line.x = element_line()
+	)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+g2 <- ggplot(earlydata) +
+	geom_line(aes(early, R0, col="Epidemic")) +
+	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
+  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
+  geom_point(aes(x=0.23, y=yintercept), size=5, col="red") +
+	scale_color_manual(values=c("black", "red")) +
+	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
+	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +
+	ggtitle("B") +
+	theme(
+		legend.position=c(0.85, 0.9),
+		legend.title = element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+grid.arrange(g1, g2, nrow=1)
+

### 2020-10-04 code/data.R

@@ -1,0 +1,9 @@
+## https://www.who.int/bulletin/volumes/91/1/12-106807/en/
+
+
+hiv_ts <- data.frame(
+	year=1990:2010,
+	prevalence=c(0.8, 1.4, 2.4, 4.3, 7.6, 10.4, 14.2, 17, 22.8,
+				 22.4, 24.5, 24.8, 26.5, 27.9, 29.5, 30.2, 29.1, 29.4,
+				 29.3, 29.4, 30.2)/100
+)

### 2020-10-04 code/tt.R

@@ -1,0 +1,111 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+sourceFiles()
+makeGraphics(width=8, height=6)
+
+maxRate <- 1/6
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+b_base <- b0fun(h_base, coef(lfit)[2]/12)
+
+tt <- testingFun(maxRate)
+
+g1 <- ggplot(tt) +
+	geom_line(aes(time/12, strength), col=2) +
+	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
+	scale_x_continuous("Time (years)", expand=c(0,0), limits=c(0, 16.5)) +
+	scale_y_log10(expression(L[test])) +
+	ggtitle("A") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+interstrength <- sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	theFun(hh, tt)
+})
+
+strengthdata <- data.frame(
+	early=early,
+	strength=interstrength
+)
+
+g2 <- ggplot(earlydata) +
+	geom_line(aes(early, R0, col="Epidemic")) +
+	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
+	scale_color_manual(values=c("black", "red")) +
+  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
+  geom_point(data=strengthdata[strengthdata$early==0.23,], aes(early, strength), size=5, col="red") +
+	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
+	scale_y_log10("Strength", limits=c(1, 16), expand=c(0, 0), breaks=c(1, 2, 4, 8, 16)) +
+	ggtitle("B") +
+	theme(
[…]

### 2020-10-04 figure/coronaexample.R
Refactoring

@@ -1,0 +1,72 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+sourceFiles()
+makeGraphics(width=8, height=3)
+
+h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+
+g1 <- ggplot(h_base) +
+	geom_line(aes(time/12, density)) +
+	scale_x_continuous("Generation time (years)") +
+	scale_y_continuous("Density", expand=c(0, 0)) +
+	ggtitle("A") +
+	theme(
+		legend.position=c(0.9, 0.9),
+		legend.title=element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+
+coef(lfit)
+
+pp <- predict(lfit, se=TRUE)
+
+g2 <- ggplot(hiv_ts) +
+	geom_point(aes(year, prevalence)) +
+	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
+	scale_x_continuous("Year") +
+	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
+	ggtitle("B") +
+	theme(
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+r2R(h_base, coef(lfit)[2]/12)
+
+early <- seq(from=0.1, to=0.4, by=0.01)
+
+R0 <-  sapply(early, function(x) {
+	hh <- HIVgen(earlyProp = x, step=0.1)
+	r2R(hh, coef(lfit)[2]/12)
+})
+
+earlydata <- data.frame(
+	early=early,
+	R0=R0
+)
+
+g3 <- ggplot(earlydata) +
+	geom_line(aes(early, R0)) +
+	geom_point(aes(0.23, 3.17), size=5) +
+	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
+	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
+	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
[…]

### 2020-10-04 R/HIVfuns.R
amsync

@@ -16,17 +16,6 @@ testingFun <- function(maxRate
 	))
 }
 
-flatFun <- function(L
-					, step = 0.1, window=200
-){
-	time <- seq(step, window, by=step)
-	hazRate <- 0*time
-	return(data.frame(time
-					  , hazRate
-					  , strength = hazRate+L
-	))
-}
-
 HIVgen <- function(earlyProp 
 				   , earlyMean=3, earlyShape=3
 				   , lateMean=120, lateShape=2

### 2020-10-04 R/HIVfuns.R
Simplify directory structure

@@ -1,32 +1,0 @@
-# Units are months
-# Units of g are 1/month, so Sum{step*g) should be 1
-
-testingFun <- function(maxRate
-					   , hMean=12, hShape=2, hheight=4
-					   , step = 0.1, window=200
-){
-	time <- seq(step, window, by=step)
-	tHaz0 <- dgamma(time, shape=hShape, scale=hMean/hShape)
-	tHaz <- hheight*tHaz0/max(tHaz0)
-	hazRate <- maxRate*(1-exp(-tHaz))
-	Haz <- step*cumsum(hazRate)
-	return(data.frame(time
-					  , hazRate
-					  , strength = exp(Haz)
-	))
-}
-
-HIVgen <- function(earlyProp 
-				   , earlyMean=3, earlyShape=3
-				   , lateMean=120, lateShape=2
-				   , step = 0.1, window=200
-){
-	time <- seq(step, window, by=step)
-	d0 <- (
-		earlyProp*dgamma(time, shape=earlyShape, scale=earlyMean/earlyShape)
-		+ (1-earlyProp)*dgamma(time, shape=lateShape, scale=lateMean/lateShape)
-	)
-	return(data.frame(time
-					  , density=d0/sum(step*d0)
-	))
-}

### 2020-10-04 R/data.R

@@ -1,8 +1,0 @@
-## https://www.who.int/bulletin/volumes/91/1/12-106807/en/
-
-hiv_ts <- data.frame(
-	year=1990:2010,
-	prevalence=c(0.8, 1.4, 2.4, 4.3, 7.6, 10.4, 14.2, 17, 22.8,
-				 22.4, 24.5, 24.8, 26.5, 27.9, 29.5, 30.2, 29.1, 29.4,
-				 29.3, 29.4, 30.2)/100
-)

### 2020-10-04 R/kernel.R

@@ -1,121 +1,0 @@
-# g, b are data frames with time and density
-# L is a data frame with time, strength and something about hazard
-
-## Calculate the initial backward distribution
-b0fun <- function(g, r){
-	step <- with(g, time[[2]]-time[[1]])
-	d0 <- with(g, density*exp(-r*time))
-	return(with(g, data.frame(time
-							  , density=d0/sum(step*d0)
-	)))
-}
-
-# How much does a given discount (r) _reduce_ R for distribution g?
-r2R <- function(g, r){
-	step <- with(g, time[[2]]-time[[1]])
-	return(with(g, 1/(step*sum(density*exp(-r*time)))))
-}
-
-R2r <- function(g, R){
-	step <- with(g, time[[2]]-time[[1]])
-	return(uniroot(
-		function(r, g){r2R(g, r)-R}
-		, interval=c(0, 1/step)
-		, g=g
-	)$root)
-}
-
-# Calculate the arithmetic mean inside the implicit equation for phi
-phiBar <- function(b, L, phi){
-	step <- with(b, time[[2]]-time[[1]])
-	return(with(cbind(b,L), {
-		step*sum(density*exp(phi*time)/strength)
-	}))
-}
-
-phiFun <- function(b, L){
-	step <- with(b, time[[2]]-time[[1]])
-	return(uniroot(
-		function(phi, b, L){phiBar(b, L, phi)-1}
-		, interval=c(0, 1/step)
-		, b=b, L=L
-	)$root)
-}
-
-theFun <- function(g, L){
-	step <- with(g, time[[2]]-time[[1]])
-	return(with(cbind(g, L), {
-		1/(step*sum(density/strength))
-	}))
-}
-
-densPlot <- function(time, k, scen, color="black", ymax=NULL, title=NULL){
-	if(is.null(ymax)) ymax <- max(k)
-	with(c(scen), {
-		if(is.null(title)) title <- disease
-		if(is.numeric(title)) title <-
-				paste("Reproductive number", sprintf("%5.2f", title))
-		# if(is.numeric(title)) title <- expression(R=title)
-		plot(time, k
-			 , main = title
-			 , xlab = paste0("time (", unit, ")")
-			 , ylab = paste0("Density (1/", unit, ")")
-			 , type = "l", col=color
-			 , ylim = c(0, ymax)
-		)
-	})
-}
-
[…]

### 2020-10-04 figure/HIV.R

@@ -1,73 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-source("../R/data.R")
-source("../R/HIVfuns.R")
-source("../R/kernel.R")
-
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-
-g1 <- ggplot(h_base) +
-	geom_line(aes(time/12, density)) +
-	scale_x_continuous("Generation time (years)") +
-	scale_y_continuous("Density", expand=c(0, 0)) +
-	ggtitle("A") +
-	theme(
-		legend.position=c(0.9, 0.9),
-		legend.title=element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-coef(lfit)
-
-pp <- predict(lfit, se=TRUE)
-
-g2 <- ggplot(hiv_ts) +
-	geom_point(aes(year, prevalence)) +
-	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
-	scale_x_continuous("Year") +
-	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
-	ggtitle("B") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-r2R(h_base, coef(lfit)[2]/12)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <-  sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-g3 <- ggplot(earlydata) +
-	geom_line(aes(early, R0)) +
-	geom_point(aes(0.23, 3.17), size=5) +
-	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
-	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
-	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
[…]

### 2020-10-04 figure/HIVexample.R

@@ -1,73 +1,0 @@
-
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R")
-sourceFiles()
-makeGraphics(width=8, height=3)
-
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-
-g1 <- ggplot(h_base) +
-	geom_line(aes(time/12, density)) +
-	scale_x_continuous("Generation time (years)") +
-	scale_y_continuous("Density", expand=c(0, 0)) +
-	ggtitle("A") +
-	theme(
-		legend.position=c(0.9, 0.9),
-		legend.title=element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-coef(lfit)
-
-pp <- predict(lfit, se=TRUE)
-
-g2 <- ggplot(hiv_ts) +
-	geom_point(aes(year, prevalence)) +
-	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
-	scale_x_continuous("Year") +
-	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
-	ggtitle("B") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-r2R(h_base, coef(lfit)[2]/12)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <-  sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-g3 <- ggplot(earlydata) +
-	geom_line(aes(early, R0)) +
-	geom_point(aes(0.23, 3.17), size=5) +
-	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
-	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
-	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
[…]

### 2020-10-04 figure/condom.R

@@ -1,57 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-source("../R/data.R")
-source("../R/HIVfuns.R")
-source("../R/kernel.R")
-
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-yintercept <- 4 ## Assumption about condom effectiveness
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-g1 <- ggplot() +
-	geom_hline(yintercept=yintercept, col=2) +
-	scale_x_continuous("Time since infection") +
-	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
-	ggtitle("A") +
-	theme(
-		panel.grid = element_blank()
-		, panel.border = element_blank()
-		, axis.line.y = element_line()
-		## , axis.line.x = element_line()
-	)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-g2 <- ggplot(earlydata) +
-	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
-  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
-  geom_point(aes(x=0.23, y=yintercept), size=5, col="red") +
-	scale_color_manual(values=c("black", "red")) +
-	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
-	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +
-	ggtitle("B") +
-	theme(
-		legend.position=c(0.85, 0.9),
-		legend.title = element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-gtot <- arrangeGrob(g1, g2, nrow=1)
-
-ggsave("condom.pdf", gtot, width=8, height=3)

### 2020-10-04 figure/constant_intervention.R

@@ -1,60 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-source("../R/kernel.R")
-
-gbar <- 16.2
-gshape <- 1/0.58^2
-
-tvec <- seq(0, 50, by=0.01)
-
-g1 <- data.frame(
-  time=tvec,
-  density=1.5*dgamma(tvec, shape=gshape, rate=gshape/gbar)
-)
-
-g2 <- data.frame(
-  time=tvec,
-  density=dgamma(tvec, shape=gshape, rate=gshape/gbar)
-)
-
-p1 <- ggplot(data=g1) +
-  geom_line(aes(time, density, col="Baseline")) +
-  geom_line(data=g2, aes(time, density, col="Controlled")) +
-  geom_segment(aes(x=10.75, xend=10.75, y=0.07, yend=0.053),
-               arrow = arrow(length = unit(0.3, "cm"))) +
-  ggtitle("A. Constant strength intervention") +
-  scale_x_continuous("Time (days)") +
-  scale_y_continuous("Density") +
-  scale_color_manual(values=c(1, 2)) +
-  theme(
-    panel.grid = element_blank(),
-    panel.border = element_blank(),
-    axis.line=element_line(),
-    legend.position = c(0.85, 0.9),
-    legend.title = element_blank()
-  )
-
-b0 <- b0fun(g2, R2r(g2, 1.5))
-
-p2 <- ggplot(data=g1) +
-  geom_line(aes(time, density, col="Baseline")) +
-  geom_line(data=b0, aes(time, density, col="Controlled")) +
-  geom_segment(aes(x=10.75, xend=9.39, y=0.07, yend=0.06),
-               arrow = arrow(length = unit(0.3, "cm"))) +
-  ggtitle("B. Constant speed intervention") +
-  scale_x_continuous("Time (days)") +
-  scale_y_continuous("Density") +
-  scale_color_manual(values=c(1, 2)) +
-  theme(
-    panel.grid = element_blank(),
-    panel.border = element_blank(),
-    axis.line=element_line(),
-    legend.position = c(0.85, 0.9),
-    legend.title = element_blank()
-  )
-
-ptot <- arrangeGrob(p1, p2, nrow=1)
-
-ggsave("constant_intervention.pdf", ptot, width=8, height=3)

### 2020-10-04 figure/coronaexample.R

@@ -1,72 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R")
-sourceFiles()
-makeGraphics(width=8, height=3)
-
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-
-g1 <- ggplot(h_base) +
-	geom_line(aes(time/12, density)) +
-	scale_x_continuous("Generation time (years)") +
-	scale_y_continuous("Density", expand=c(0, 0)) +
-	ggtitle("A") +
-	theme(
-		legend.position=c(0.9, 0.9),
-		legend.title=element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-coef(lfit)
-
-pp <- predict(lfit, se=TRUE)
-
-g2 <- ggplot(hiv_ts) +
-	geom_point(aes(year, prevalence)) +
-	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
-	scale_x_continuous("Year") +
-	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
-	ggtitle("B") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-r2R(h_base, coef(lfit)[2]/12)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <-  sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-g3 <- ggplot(earlydata) +
-	geom_line(aes(early, R0)) +
-	geom_point(aes(0.23, 3.17), size=5) +
-	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
-	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
-	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
[…]

### 2020-10-04 figure/gencontrol.R

@@ -1,62 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R")
-sourceFiles()
-makeGraphics(width=8, height=3)
-
-gbar <- 16.2
-gshape <- 1/0.58^2
-
-tvec <- seq(0, 50, by=0.01)
-
-g1 <- data.frame(
-  time=tvec,
-  density=1.5*dgamma(tvec, shape=gshape, rate=gshape/gbar)
-)
-
-g2 <- data.frame(
-  time=tvec,
-  density=dgamma(tvec, shape=gshape, rate=gshape/gbar)
-)
-
-p1 <- ggplot(data=g1) +
-  geom_line(aes(time, density, col="Baseline")) +
-  geom_line(data=g2, aes(time, density, col="Controlled")) +
-  geom_segment(aes(x=10.75, xend=10.75, y=0.07, yend=0.053),
-               arrow = arrow(length = unit(0.3, "cm"))) +
-  ggtitle("A. Constant strength intervention") +
-  scale_x_continuous("Time (days)") +
-  scale_y_continuous("Density") +
-  scale_color_manual(values=c(1, 2)) +
-  theme(
-    panel.grid = element_blank(),
-    panel.border = element_blank(),
-    axis.line=element_line(),
-    legend.position = c(0.85, 0.9),
-    legend.title = element_blank()
-  )
-
-b0 <- b0fun(g2, R2r(g2, 1.5))
-
-p2 <- ggplot(data=g1) +
-  geom_line(aes(time, density, col="Baseline")) +
-  geom_line(data=b0, aes(time, density, col="Controlled")) +
-  geom_segment(aes(x=10.75, xend=9.39, y=0.07, yend=0.06),
-               arrow = arrow(length = unit(0.3, "cm"))) +
-  ggtitle("B. Constant speed intervention") +
-  scale_x_continuous("Time (days)") +
-  scale_y_continuous("Density") +
-  scale_color_manual(values=c(1, 2)) +
-  theme(
-    panel.grid = element_blank(),
-    panel.border = element_blank(),
-    axis.line=element_line(),
-    legend.position = c(0.85, 0.9),
-    legend.title = element_blank()
-  )
-
-grid.arrange(p1, p2, nrow=1)
-

### 2020-10-04 figure/hivstrength.R

@@ -1,60 +1,0 @@
-
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R") 
-
-commandEnvironments() ## Read in any environments specified as dependencies
-sourceFiles()
-makeGraphics(width=8, height=3)
-
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-yintercept <- 4 ## Assumption about condom effectiveness
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-g1 <- ggplot() +
-	geom_hline(yintercept=yintercept, col=2) +
-	scale_x_continuous("Time since infection") +
-	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
-	ggtitle("A") +
-	theme(
-		panel.grid = element_blank()
-		, panel.border = element_blank()
-		, axis.line.y = element_line()
-		## , axis.line.x = element_line()
-	)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-g2 <- ggplot(earlydata) +
-	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
-  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
-  geom_point(aes(x=0.23, y=yintercept), size=5, col="red") +
-	scale_color_manual(values=c("black", "red")) +
-	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
-	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +
-	ggtitle("B") +
-	theme(
-		legend.position=c(0.85, 0.9),
-		legend.title = element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-grid.arrange(g1, g2, nrow=1)
-

### 2020-10-04 figure/test_and_treat.R

@@ -1,111 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-source("../R/data.R")
-source("../R/HIVfuns.R")
-source("../R/kernel.R")
-
-maxRate <- 1/6
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-b_base <- b0fun(h_base, coef(lfit)[2]/12)
-
-tt <- testingFun(maxRate)
-
-g1 <- ggplot(tt) +
-	geom_line(aes(time/12, strength), col=2) +
-	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
-	scale_x_continuous("Time (years)", expand=c(0,0), limits=c(0, 16.5)) +
-	scale_y_log10(expression(L[test])) +
-	ggtitle("A") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-interstrength <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	theFun(hh, tt)
-})
-
-strengthdata <- data.frame(
-	early=early,
-	strength=interstrength
-)
-
-g2 <- ggplot(earlydata) +
-	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
-	scale_color_manual(values=c("black", "red")) +
-  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
-  geom_point(data=strengthdata[strengthdata$early==0.23,], aes(early, strength), size=5, col="red") +
-	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
-	scale_y_log10("Strength", limits=c(1, 16), expand=c(0, 0), breaks=c(1, 2, 4, 8, 16)) +
-	ggtitle("B") +
-	theme(
[…]

### 2020-10-04 figure/tt.R

@@ -1,111 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R")
-sourceFiles()
-makeGraphics(width=8, height=6)
-
-maxRate <- 1/6
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-b_base <- b0fun(h_base, coef(lfit)[2]/12)
-
-tt <- testingFun(maxRate)
-
-g1 <- ggplot(tt) +
-	geom_line(aes(time/12, strength), col=2) +
-	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
-	scale_x_continuous("Time (years)", expand=c(0,0), limits=c(0, 16.5)) +
-	scale_y_log10(expression(L[test])) +
-	ggtitle("A") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-interstrength <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	theFun(hh, tt)
-})
-
-strengthdata <- data.frame(
-	early=early,
-	strength=interstrength
-)
-
-g2 <- ggplot(earlydata) +
-	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
-	scale_color_manual(values=c("black", "red")) +
-  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
-  geom_point(data=strengthdata[strengthdata$early==0.23,], aes(early, strength), size=5, col="red") +
-	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
-	scale_y_log10("Strength", limits=c(1, 16), expand=c(0, 0), breaks=c(1, 2, 4, 8, 16)) +
-	ggtitle("B") +
-	theme(
[…]

### 2020-10-04 code/HIVexample.R
This is why I don't work!!

@@ -1,4 +1,3 @@
-
 library(tidyr)
 library(dplyr)
 library(ggplot2); theme_set(theme_bw())
@@ -10,10 +9,12 @@ makeGraphics(width=8, height=3)
 
 h_base <- HIVgen(earlyProp = 0.23, step=0.1)
 
+scale <- 12 ## Intervals are months, plots and time series are in years
+
 g1 <- ggplot(h_base) +
-	geom_line(aes(time/12, density)) +
+	geom_line(aes(time/scale, density*scale)) +
 	scale_x_continuous("Generation time (years)") +
-	scale_y_continuous("Density", expand=c(0, 0)) +
+	scale_y_continuous("Density (per year)", expand=c(0, 0)) +
 	ggtitle("A") +
 	theme(
 		legend.position=c(0.9, 0.9),
@@ -23,15 +24,18 @@ g1 <- ggplot(h_base) +
 		axis.line = element_line()
 	)
 
+## Parameters from fit to prevalence ts
 lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
+p0 <- coef(lfit)[[1]]
+rfity <- coef(lfit)[[2]]
+rfitm <- coef(lfit)[[2]]/scale
 
 coef(lfit)
 
-pp <- predict(lfit, se=TRUE)
-
+## Prevalence time plot
 g2 <- ggplot(hiv_ts) +
 	geom_point(aes(year, prevalence)) +
-	stat_function(fun=function(x) exp(coef(lfit)[1] + coef(lfit)[2] * x)) +
+	stat_function(fun=function(x) exp(p0 + rfity*x)) +
 	scale_x_continuous("Year") +
 	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
 	ggtitle("B") +
@@ -41,13 +45,13 @@ g2 <- ggplot(hiv_ts) +
 		axis.line = element_line()
 	)
 
-r2R(h_base, coef(lfit)[2]/12)
+r2R(h_base, rfitm)
 
 early <- seq(from=0.1, to=0.4, by=0.01)
 
 R0 <-  sapply(early, function(x) {
 	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
+	r2R(hh, rfitm)
 })
 
 earlydata <- data.frame(
[…]

### 2020-10-04 code/coronaexample.R

@@ -1,0 +1,28 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+commandEnvironments()
+## makeGraphics(width=8, height=3)
+makeGraphics()
+
+g_pre <- gen(sat=0)
+g_post <- gen(sat=8)
+
+print(
+	ggplot(g_pre) +
+	geom_line(aes(time, density)) +
+	scale_x_continuous("Generation time") +
+	scale_y_continuous("Density", expand=c(0, 0)) +
+	theme(
+		legend.position=c(0.9, 0.9),
+		legend.title=element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	) +
+	NULL
+)
+

### 2020-10-04 code/coronafuns.R

@@ -1,0 +1,14 @@
+# Units are days
+
+gen <- function(sat, sMean=5, sShape=3
+	, step = 0.1, window=200
+){
+	time <- seq(step, window, by=step)
+	d0 <- (
+		dgamma(time, shape=sShape, scale=sMean/sShape)
+		* time/(time+sat)
+	)
+	return(data.frame(
+		time, density=d0/sum(step*d0)
+	))
+}

### 2020-10-04 code/HIVexample.R
Meet with Daniel

@@ -4,7 +4,7 @@ library(ggplot2); theme_set(theme_bw())
 library(gridExtra)
 
 source("makestuff/makeRfuns.R")
-sourceFiles()
+commandEnvironments()
 makeGraphics(width=8, height=3)
 
 h_base <- HIVgen(earlyProp = 0.23, step=0.1)

### 2020-10-04 code/coronaexample.R

@@ -14,8 +14,8 @@ g_post <- gen(sat=8)
 print(
 	ggplot(g_pre) +
 	geom_line(aes(time, density)) +
-	scale_x_continuous("Generation time") +
-	scale_y_continuous("Density", expand=c(0, 0)) +
+	scale_x_continuous("Generation time (days)") +
+	scale_y_continuous("Density (per day)", expand=c(0, 0)) +
 	theme(
 		legend.position=c(0.9, 0.9),
 		legend.title=element_blank(),
@@ -23,6 +23,7 @@ print(
 		panel.border = element_blank(),
 		axis.line = element_line()
 	) +
+	geom_line(aes(time, density), data=g_post) +
 	NULL
 )
 

### 2020-10-04 code/coronafuns.R

@@ -1,7 +1,6 @@
 # Units are days
-
 gen <- function(sat, sMean=5, sShape=3
-	, step = 0.1, window=200
+	, step = 0.1, window=20
 ){
 	time <- seq(step, window, by=step)
 	d0 <- (

### 2020-10-04 code/kernel.R

@@ -119,3 +119,6 @@ intPlot <- function(time, density, strength, R){
 	lines(time, R*density/strength, type="l")
 	par(op)
 }
+
+intervalMoments <- function(g){
+}

### 2020-10-04 significance.tex
Moving doc to main directory

\section*{Significance statement}

Researchers studying infectious-disease outbreaks often focus primarily on the reproductive number, which measures the ``strength'' of transmission --- the number of new infections caused by an infected individual.
Less theoretical and modeling attention is paid to the exponential growth rate, which measures epidemic ``speed''. 
We argue that this difference in emphasis is largely artificial: both strength and speed have thresholds for spread and provide frameworks to compare effectiveness of intervention.
We generalize the classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
We provide specific examples from HIV and other diseases to demonstrate the value of both perspectives.

### 2020-10-04 abstract.pl

@@ -1,0 +1,18 @@
+use strict;
+use 5.10.0;
+
+$/ = "";
+
+while(<>){
+	if (/section.*Abstract/ .. /pagebreak/){
+		next if/^\\/;
+		s/\n/ /g;
+		s/\$//g;
+		s/\\//g;
+		s/RR/R/g;
+		s/``/“/g;
+		s/''/”/g;
+		s/---?/–/g;
+		say;
+	}
+}

### 2020-10-04 code/kernel.R
trivial

@@ -120,5 +120,4 @@ intPlot <- function(time, density, strength, R){
 	par(op)
 }
 
-intervalMoments <- function(g){
-}
+intervalMoments <- function(g){}

### 2020-10-04 doc/abstract.pl
Good-bye, doc!

@@ -1,18 +1,0 @@
-use strict;
-use 5.10.0;
-
-$/ = "";
-
-while(<>){
-	if (/section.*Abstract/ .. /pagebreak/){
-		next if/^\\/;
-		s/\n/ /g;
-		s/\$//g;
-		s/\\//g;
-		s/RR/R/g;
-		s/``/“/g;
-		s/''/”/g;
-		s/---?/–/g;
-		say;
-	}
-}

### 2020-10-06 code/coronaexample.R
Good night!

@@ -9,7 +9,10 @@ commandEnvironments()
 makeGraphics()
 
 g_pre <- gen(sat=0)
-g_post <- gen(sat=8)
+g_post <- gen(sat=10)
+
+print(intervalMoments(g_pre))
+print(intervalMoments(g_post))
 
 print(
 	ggplot(g_pre) +

### 2020-10-06 code/coronafuns.R

@@ -1,13 +1,38 @@
 # Units are days
+
+## A phenomenological generation interval
+## A gamma function multiplied by a saturation step
+## Slow saturation corresponds to less pre-symptomatic transmission
 gen <- function(sat, sMean=5, sShape=3
-	, step = 0.1, window=20
+	, step = 0.05, window=20
 ){
 	time <- seq(step, window, by=step)
 	d0 <- (
 		dgamma(time, shape=sShape, scale=sMean/sShape)
-		* time/(time+sat)
+		* (time/(time+sat))
 	)
 	return(data.frame(
 		time, density=d0/sum(step*d0)
 	))
 }
+
+## How quickly is COVID-19 found and isolated?
+## • decay rate (units 1/time) to represent contract tracing
+## • Saturation to correspond to symptom-based isolation
+## 	Not directly analogous to the saturation above!
+## 	We are assuming isolation is not changing across our range of 
+## 	pre-symptomatic scenarios
+
+controlFun <- function(baseRate
+	, sat = 0, decay=0
+	, step = 0.05, window=20
+){
+	time <- seq(step, window, by=step)
+	hazRate <- baseRate*exp(-decay*time)*time/(time+sat)
+	Haz <- step*cumsum(hazRate)
+	return(data.frame(time
+		, hazRate
+		, strength = exp(Haz)
+	))
+}
+

### 2020-10-06 code/kernel.R

@@ -120,4 +120,15 @@ intPlot <- function(time, density, strength, R){
 	par(op)
 }
 
-intervalMoments <- function(g){}
+intervalMoments <- function(g){
+	attach(g)
+	s0 <- sum(density)
+	s1 <- sum(time*density)
+	s2 <- sum(time*time*density)
+	detach(g)
+
+	return(c(weight=s0
+		, mean = s1/s0
+		, CV2 = s2*s0/s1^2-1
+	))
+}

### 2020-10-08 code/HIVexample.R
Lots of refactoring; some units

@@ -9,10 +9,8 @@ makeGraphics(width=8, height=3)
 
 h_base <- HIVgen(earlyProp = 0.23, step=0.1)
 
-scale <- 12 ## Intervals are months, plots and time series are in years
-
 g1 <- ggplot(h_base) +
-	geom_line(aes(time/scale, density*scale)) +
+	geom_line(aes(time*month/year, density*year/month)) +
 	scale_x_continuous("Generation time (years)") +
 	scale_y_continuous("Density (per year)", expand=c(0, 0)) +
 	ggtitle("A") +
@@ -24,18 +22,10 @@ g1 <- ggplot(h_base) +
 		axis.line = element_line()
 	)
 
-## Parameters from fit to prevalence ts
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-p0 <- coef(lfit)[[1]]
-rfity <- coef(lfit)[[2]]
-rfitm <- coef(lfit)[[2]]/scale
-
-coef(lfit)
-
 ## Prevalence time plot
 g2 <- ggplot(hiv_ts) +
 	geom_point(aes(year, prevalence)) +
-	stat_function(fun=function(x) exp(p0 + rfity*x)) +
+	stat_function(fun=function(x) exp(p0 + rfityear*x)) +
 	scale_x_continuous("Year") +
 	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
 	ggtitle("B") +
@@ -45,13 +35,13 @@ g2 <- ggplot(hiv_ts) +
 		axis.line = element_line()
 	)
 
-r2R(h_base, rfitm)
+r2R(h_base, rfitmonth)
 
 early <- seq(from=0.1, to=0.4, by=0.01)
 
 R0 <-  sapply(early, function(x) {
 	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, rfitm)
+	r2R(hh, rfitmonth)
 })
 
 earlydata <- data.frame(

### 2020-10-08 code/condom.R

@@ -13,8 +13,6 @@ makeGraphics(width=8, height=3)
 h_base <- HIVgen(earlyProp = 0.23, step=0.1)
 yintercept <- 4 ## Assumption about condom effectiveness
 
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
 g1 <- ggplot() +
 	geom_hline(yintercept=yintercept, col=2) +
 	scale_x_continuous("Time since infection") +
@@ -30,8 +28,8 @@ g1 <- ggplot() +
 early <- seq(from=0.1, to=0.4, by=0.01)
 
 R0 <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
+	hh <- HIVgen(earlyProp = x)
+	r2R(hh, rfitmonth)
 })
 
 earlydata <- data.frame(

### 2020-10-08 code/data.R

@@ -1,6 +1,5 @@
 ## https://www.who.int/bulletin/volumes/91/1/12-106807/en/
 
-
 hiv_ts <- data.frame(
 	year=1990:2010,
 	prevalence=c(0.8, 1.4, 2.4, 4.3, 7.6, 10.4, 14.2, 17, 22.8,

### 2020-10-08 code/tt.R

@@ -4,15 +4,13 @@ library(ggplot2); theme_set(theme_bw())
 library(gridExtra)
 
 source("makestuff/makeRfuns.R")
-sourceFiles()
+commandEnvironments()
 makeGraphics(width=8, height=6)
 
 maxRate <- 1/6
 h_base <- HIVgen(earlyProp = 0.23, step=0.1)
 
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:8,])
-
-b_base <- b0fun(h_base, coef(lfit)[2]/12)
+b_base <- b0fun(h_base, rfitmonth)
 
 tt <- testingFun(maxRate)
 
@@ -32,7 +30,7 @@ early <- seq(from=0.1, to=0.4, by=0.01)
 
 R0 <- sapply(early, function(x) {
 	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, coef(lfit)[2]/12)
+	r2R(hh, rfitmonth)
 })
 
 earlydata <- data.frame(
@@ -81,7 +79,7 @@ g3 <- ggplot(tt) +
 
 interspeed <- sapply(early, function(x) {
 	hh <- HIVgen(earlyProp = x, step=0.1)
-	bb <- b0fun(hh, coef(lfit)[2]/12)
+	bb <- b0fun(hh, rfitmonth)
 	
 	phiFun(bb, tt)
 })
@@ -93,9 +91,9 @@ speeddata <- data.frame(
 
 g4 <- ggplot(speeddata) +
 	geom_line(aes(early, speed, col="Intervention")) +
-	geom_hline(aes(yintercept=coef(lfit)[[2]], col="Epidemic" )) +
+	geom_hline(aes(yintercept=rfityear, col="Epidemic" )) +
   geom_point(data=speeddata[speeddata$early==0.23,], aes(early, speed), size=5, col="red") +
-  geom_point(x=0.23, y=coef(lfit)[[2]], size=5) +
+  geom_point(x=0.23, y=rfityear, size=5) +
 	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
 	scale_y_continuous(expression(Speed~(year^{-1})), limits=c(0, 0.8), expand=c(0, 0)) +
 	scale_color_manual(values=c("black", "red")) +

### 2020-10-08 code/units.R

@@ -1,0 +1,11 @@
+## Units are retrofitted, and not consistently implemented yet!
+## 2020 Oct 08 (Thu)
+
+source("makestuff/makeRfuns.R") 
+
+year <- 1
+day <- year/365.25
+month <- year/12
+
+saveVars(day, year, month)
+

### 2020-10-08 code/coronaexample.R
Doodling with coronafuns

@@ -8,15 +8,16 @@ commandEnvironments()
 ## makeGraphics(width=8, height=3)
 makeGraphics()
 
-g_pre <- gen(sat=0)
-g_post <- gen(sat=10)
+## Generation intervals
+g_pre <- gen(sat=0*day)
+g_post <- gen(sat=10*day)
 
 print(intervalMoments(g_pre))
 print(intervalMoments(g_post))
 
 print(
 	ggplot(g_pre) +
-	geom_line(aes(time, density)) +
+	geom_line(aes(time/day, density*day)) +
 	scale_x_continuous("Generation time (days)") +
 	scale_y_continuous("Density (per day)", expand=c(0, 0)) +
 	theme(
@@ -26,7 +27,30 @@ print(
 		panel.border = element_blank(),
 		axis.line = element_line()
 	) +
-	geom_line(aes(time, density), data=g_post) +
+	geom_line(aes(time/day, density*day), data=g_post) +
 	NULL
 )
 
+## Interventions
+
+base <- 0.1/day
+
+flat <- controlFun(base/2)
+trace <- controlFun(base, decay=1/4/day)
+respond <- controlFun(base, sat=4*day)
+
+print(ggplot()
+	+ scale_x_continuous("Time since infection (days)")
+	+ scale_y_continuous("Control hazard (per day)", expand=c(0, 0))
+	+ geom_line(data=flat, aes(time/day, hazRate*day))
+	+ geom_line(data=trace, aes(time/day, hazRate*day))
+	+ geom_line(data=respond, aes(time/day, hazRate*day))
+)
+
+print(ggplot()
+	+ scale_x_continuous("Time since infection (days)")
+	+ scale_y_continuous("Control strength", expand=c(0, 0))
+	+ geom_line(data=flat, aes(time/day, strength))
+	+ geom_line(data=trace, aes(time/day, strength))
+	+ geom_line(data=respond, aes(time/day, strength))
+)

### 2020-10-08 code/coronafuns.R

@@ -3,8 +3,8 @@
 ## A phenomenological generation interval
 ## A gamma function multiplied by a saturation step
 ## Slow saturation corresponds to less pre-symptomatic transmission
-gen <- function(sat, sMean=5, sShape=3
-	, step = 0.05, window=20
+gen <- function(sat, sMean=5*day, sShape=3
+	, step = 0.05*day, window=20*day
 ){
 	time <- seq(step, window, by=step)
 	d0 <- (
@@ -24,8 +24,8 @@ gen <- function(sat, sMean=5, sShape=3
 ## 	pre-symptomatic scenarios
 
 controlFun <- function(baseRate
-	, sat = 0, decay=0
-	, step = 0.05, window=20
+	, sat = 0*day, decay=0/day
+	, step = 0.05*day, window=20*day
 ){
 	time <- seq(step, window, by=step)
 	hazRate <- baseRate*exp(-decay*time)*time/(time+sat)

### 2020-10-08 code/coronaexample.R
Did all the coding I want to satisfy myself about structure and qualitative result. Not yet tried to make nice pictures.

@@ -54,3 +54,14 @@ print(ggplot()
 	+ geom_line(data=trace, aes(time/day, strength))
 	+ geom_line(data=respond, aes(time/day, strength))
 )
+
+## Intervention speed
+
+r0 <- 0.23/day
+
+phiFun(b0fun(g_pre, r0), flat)*day
+phiFun(b0fun(g_post, r0), flat)*day
+phiFun(b0fun(g_pre, r0), trace)*day
+phiFun(b0fun(g_post, r0), trace)*day
+phiFun(b0fun(g_pre, r0), respond)*day
+phiFun(b0fun(g_post, r0), respond)*day

### 2020-10-08 code/kernel.R

@@ -25,6 +25,8 @@ R2r <- function(g, R){
 	)$root)
 }
 
+### Speed of intervention
+
 # Calculate the arithmetic mean inside the implicit equation for phi
 phiBar <- function(b, L, phi){
 	step <- with(b, time[[2]]-time[[1]])
@@ -33,6 +35,7 @@ phiBar <- function(b, L, phi){
 	}))
 }
 
+### Find the phi that satisfies the equation
 phiFun <- function(b, L){
 	step <- with(b, time[[2]]-time[[1]])
 	return(uniroot(

### 2020-10-08 code/tt.R

@@ -15,7 +15,7 @@ b_base <- b0fun(h_base, rfitmonth)
 tt <- testingFun(maxRate)
 
 g1 <- ggplot(tt) +
-	geom_line(aes(time/12, strength), col=2) +
+	geom_line(aes(time*month/year, strength), col=2) +
 	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
 	scale_x_continuous("Time (years)", expand=c(0,0), limits=c(0, 16.5)) +
 	scale_y_log10(expression(L[test])) +
@@ -66,8 +66,8 @@ g2 <- ggplot(earlydata) +
 	)
 
 g3 <- ggplot(tt) +
-	geom_line(aes(time/12, hazRate*12), col=2) +
-	geom_hline(aes(yintercept=phiFun(b_base, tt)*12), lty=2, col=2) +
+	geom_line(aes(time*month/year, hazRate*year/month), col=2) +
+	geom_hline(aes(yintercept=phiFun(b_base, tt)*year/month), lty=2, col=2) +
 	scale_x_continuous("Time (years)", limits=c(0, 16.5), expand=c(0, 0)) +
 	scale_y_continuous(expression(h[test]~(year^{-1}))) +
 	ggtitle("C") +
@@ -86,7 +86,7 @@ interspeed <- sapply(early, function(x) {
 
 speeddata <- data.frame(
 	early=early,
-	speed=interspeed*12
+	speed=interspeed*year/month
 )
 
 g4 <- ggplot(speeddata) +

### 2020-10-09 speedstrength.tex
Moving doc to main directory
chaos and confusion

--- 2020-10-02 doc/speedstrength.tex
@@
 We first generalize the idea that \RR\ measures the difficulty of elimination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number. 
-We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength required for elimination, and the threshold $r=0$ and a corresponding minimal intervention speed.
+We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the fact that the threshold $\RR=1$ determines the effective strength of intervention required for disease elimination, and the fact that the threshold $r=0$ determines the effective speed of intervention required for elimination.
 We argue that the primacy of $\RR$ over $r$ is partly due to history, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{figure/gencontrol.Rout.pdf}
+\includegraphics[width=\textwidth]{code/gencontrol.Rout.pdf}
 \caption{
@@
 Again, we have that the disease cannot spread when $\theta \geq \RR$.
-In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, and the control measure must be at least as ``strong'' as the disease (as measured by \RR) to achieve elimination.
+In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, this must exceed the strength of the disease (\RR) to achieve elimination.
 
@@
 
+The disease cannot spread when $\phi \geq r$.
+In other words, \rr, like \RR, is a metric for how hard a disease is to eliminate: we can measure the ``speed'' of a control measure, this must exceed the speed of the disease (\rr) to achieve elimination.
+
 \section{Example: Human immunodeficiency virus (HIV)}
 
-In this section, we use both strength- and speed-based decompositions to compare different intervention strategies for the human immunodeficiency virus (HIV). 
+In this section, we use both strength- and speed-based decompositions to evaluate intervention strategies for the human immunodeficiency virus (HIV). 
 In particular, we study how the amount of early HIV transmission affects estimates of intervention effectiveness. 
 These examples are not detailed estimates for specific scenarios; 
-instead, they are meant to demonstrate how strength- and speed-based decompositions can help evaluate control strategies.
+instead, they are meant to demonstrate how strength- and speed-based perspectives provide different qualitative insights, while yielding the same quantitative answers.
 
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/HIVexample.Rout.pdf}
+\includegraphics[width=\textwidth]{code/HIVexample.Rout.pdf}
 \caption{
@@
 We compare two different possible intervention strategies to shed light on the speed and strength decompositions.
-First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
+First, we consider a condom intervention that reduces HIV transmission by an average of approximately 75\%.
 Assuming that condoms act as a physical barrier, and that condom use will, on average, remain roughly constant through time, it is reasonable to model the multiplicative reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$ (\figref{condom}A).
 The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$, i.e., $\theta=4$, whereas the estimated strength of the epidemic \RR\ decreases as the proportion of early transmission \pEarly\ increases (\figref{condom}B).
-Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if the amount of early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.
+Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if the amount of early transmission is low, we infer that disease spread is too strong to be controlled completely by our intervention.
 
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{figure/hivstrength.Rout.pdf}
+\includegraphics[width=\textwidth]{code/condom.Rout.pdf}
 \caption{
@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{figure/tt.Rout.pdf}
+\includegraphics[width=\textwidth]{code/tt.Rout.pdf}
 \caption{
@@
 In this example, we see that, as \pEarly\ goes down and our estimate of epidemic strength increases, the estimate of intervention strength increases roughly in parallel.
-The increase in intervention strength makes sense, as we have more time to reach people on average before they transmit: this is the core of the result of \cite{eaton2014proportion}.
+The increase in intervention strength makes sense: less early transmission means more time to reach people before they transmit and higher strength of control.
+This is the core of the result of \cite{eaton2014proportion}.
 In our scenario, we predict that the intervention remains effective over the range of considered parameters.
@@
 Pre-symptomatic transmission is likely to be hard to detect, and therefore hard to prevent.
-Thus, it might be supposed that an increase in the estimated importance of pre-symptomatic transmission would lead to an increase in the estimated difficulty.
+Thus, it might be supposed that an increase in the estimated importance of pre-symptomatic transmission would lead to an increase in the estimated difficulty of control.
 
-A generation-interval perspective can be used to challenge this view.
+A generation-interval perspective \cite{park2019practical} can be used to challenge this view.
 More early transmission, for a given value of \rr, means less transmission per individual -- that is, a lower value of \RR.
 Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary.
+For interventions like lockdowns, which are generally assumed affect everyone roughly equally, the strength-based perspective thus gives a clear (and somewhat surprising) answer: more early transmission means we will conclude that non-targeted interventions are \emph{more} effective.
 
-For interventions which are SOMETHING, the strength-based perspective gives a clear answer.
+For interventions which focus on infected individuals, like contact tracing or test-based isolation, the speed-based paradigm provides clearer insight into the likely effects of early transmission. 
+More early transmission does not change our estimate of the speed of the epidemic, but it does change our estimate of the \emph{effective} speed of a given control curve.
 
-Conversely,
+Early transmission gives more weight in the calculation of effective speed ($\phi$) to the early period of infection.
+Thus, if the hazard of control is increasing through time, as might be expected with isolation of symptomatic cases, more early transmission means that the effective speed of control is lower than originally thought, and control will be harder.
+Conversely, if hazard of control decreases through time, as might be expected with isolation of returning travellers, or event-based contact tracing, more early transmission means that control will be easier than originally though.
 
-Given sufficient assumptions, a strength-based calculation could provide an answer for any particular scenario.
-But using the speed-based paradigm substantially clarifies the situation.
-As with the HIV example, our estimate of the speed of the epidemic is not changed by information about timing of transmission.
-Thus, the question is reduced to whether (and how) we expect pre-symptomatic transmission to affect the effective speed of intervention.
-
-Case detection is based on contact tracing, testing and symptoms.
-Thus, we don't expect pre-symptomatic infection to change the intervention hazard $h$.
-The effective speed, however, is a (sort-of) mean taken over the (backwards) generation interval.
-If the
+With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
+But in these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result.
 
@@
 Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
-For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level, like condom use.
+For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that target infected people, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}; we expect the strength-based framework to be clearer for intervention strategies that target the general population, like condom use, or susceptible people, like prophylaxis.
 In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.

### 2020-10-09 speedstrength.tex
post-meeting edits

@@
 (A) Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the multiplicative reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
-(B) The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
+(B) The estimated amount of early transmission affects estimated epidemic strength \RR, but not intervention strength \Rhat of a condom-based intervention.
 The black and red circles indicate the baseline scenario.
@@
 
-%% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
+\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
+\jd{The above is an old comment. I'm not currently loving the idea of including that figure, but I guess we should make it and see?}
 
@@
 The dashed line shows the corresponding effective strength of the intervention (from \eqref{strengthMean}) assuming 23\% early transmission.
-(B) Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
+(B) Increase in the estimated amount of early transmission decreases the estimated epidemic strength \RR\ as well as the estimated strength of a test-and-treat intervention \Rhat.
 (C) The assumed hazard for the test-and-treat intervention. 
 The dashed line shows the corresponding effective speed of the intervention (from \eqref{speedMean}) assuming 23\% early transmission.
-(D) The estimated amount of early transmission has little effect on the effective speed of intervention, and none on the speed of the epidemic estimated from incidence data.
+(D) The estimated amount of early transmission has little effect on the effective speed of intervention \rhat, and none on the speed of the epidemic estimated from incidence data \rr.
 Circles indicate the baseline scenario.
@@
 More early transmission does not change our estimate of the speed of the epidemic, but it does change our estimate of the \emph{effective} speed of a given control curve. 
-
-Early transmission gives more weight in the calculation of effective speed ($\phi$) to the early period of infection.
+This is because more early transmission gives more weight in the calculation of effective speed ($\phi$) to the early period of infection.
 Thus, if the hazard of control is increasing through time, as might be expected with isolation of symptomatic cases, more early transmission means that the effective speed of control is lower than originally thought, and control will be harder.
 Conversely, if hazard of control decreases through time, as might be expected with isolation of returning travellers, or event-based contact tracing, more early transmission means that control will be easier than originally though.
-
-With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
-But in these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result.
 
@@
 In this study, we:
-first extended the standard paradigm of \RR\ as critical parameter for control, by defining the strength of an intervention on the same scale as \RR, the strength of the epidemic; 
+first extended the standard paradigm of \RR\ as a critical parameter for control, by defining the strength of an intervention on the same scale as \RR, the strength of the epidemic; 
 then constructed a parallel interpretation which measures the speed of an intervention on the same scale as $r$, the speed of an epidemic.
@@
 The speed paradigm provides an additional insight: since we expect more early transmission to make our estimate of intervention speed (a little) faster, higher amounts of early transmission (when controlling for the observed initial growth rate) are expected to make control via test-and-treat (a little) easier.
+
+We also discussed the question of uncertainties introduced by the unknown magnitude of pre-symptomatic transmission of COVID-19.
+We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while speed-based paradigm is a clearer way to think about interventions targeted to infected people.
 
@@
 Other cases, such as using social distancing (a strength-like intervention) in the early stages of COVID-19 (epidemic speed is observed) may not fit so neatly into either paradigm, however.
+With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
+But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result.
 

### 2020-10-09 code/HIVexample.R
Hand over

@@ -7,7 +7,7 @@ source("makestuff/makeRfuns.R")
 commandEnvironments()
 makeGraphics(width=8, height=3)
 
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+h_base <- HIVgen(earlyProp = earlyBase, step=0.1)
 
 g1 <- ggplot(h_base) +
 	geom_line(aes(time*month/year, density*year/month)) +
@@ -52,9 +52,9 @@ earlydata <- data.frame(
 ## R vs. early transmission
 g3 <- ggplot(earlydata) +
 	geom_line(aes(early, R0)) +
-	geom_point(aes(0.23, 3.17), size=5) +
-	geom_segment(aes(x=0.23, y=0, xend=0.23, yend=3.17), lty=2) +
-	geom_segment(aes(x=0, y=3.17, xend=0.23, yend=3.17), lty=2) +
+	geom_point(aes(earlyBase, 3.17), size=5) +
+	geom_segment(aes(x=earlyBase, y=0, xend=earlyBase, yend=3.17), lty=2) +
+	geom_segment(aes(x=0, y=3.17, xend=earlyBase, yend=3.17), lty=2) +
 	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
 	scale_y_continuous("Reproduction number", limits=c(0, 5), expand=c(0, 0)) +
 	ggtitle("C") +

### 2020-10-09 code/HIVfuns.R

@@ -1,6 +1,10 @@
 # Units are months
 # Units of g are 1/month, so Sum{step*g) should be 1
 
+earlyFast <- 0.4
+earlyBase <- 0.23
+earlySlow <- 0.1
+
 testingFun <- function(maxRate
 					   , hMean=12, hShape=2, hheight=4
 					   , step = 0.1, window=200

### 2020-10-09 code/condom.R

@@ -10,11 +10,11 @@ commandEnvironments() ## Read in any environments specified as dependencies
 sourceFiles()
 makeGraphics(width=8, height=3)
 
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
-yintercept <- 4 ## Assumption about condom effectiveness
+Lcondom <- 4
+h_base <- HIVgen(earlyProp = earlyBase, step=0.1)
 
 g1 <- ggplot() +
-	geom_hline(yintercept=yintercept, col=2) +
+	geom_hline(yintercept=Lcondom, col=2) +
 	scale_x_continuous("Time since infection") +
 	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
 	ggtitle("A") +
@@ -39,9 +39,9 @@ earlydata <- data.frame(
 
 g2 <- ggplot(earlydata) +
 	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_hline(aes(yintercept=yintercept, col="Intervention")) +
-  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
-  geom_point(aes(x=0.23, y=yintercept), size=5, col="red") +
+	geom_hline(aes(yintercept=Lcondom, col="Intervention")) +
+  geom_point(data=earlydata[earlydata$early==earlyBase,], aes(early, R0), size=5) +
+  geom_point(aes(x=earlyBase, y=Lcondom), size=5, col="red") +
 	scale_color_manual(values=c("black", "red")) +
 	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
 	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +

### 2020-10-09 code/hivGens.R

@@ -1,0 +1,36 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("makestuff/makeRfuns.R")
+commandEnvironments()
+makeGraphics(width=8, height=3)
+
+base = HIVgen(earlyProp = earlyBase)
+fast = HIVgen(earlyProp = earlyFast)
+slow = HIVgen(earlyProp = earlySlow)
+
+gens <- (
+	list(
+		base = base
+		, fast = fast
+		, slow = slow
+	)
+	%>% bind_rows(.id = "type")
+)
+
+g1 <- (ggplot(gens)
+	+ geom_line(aes(time*month/year, density*year/month, color=type))
+	+ scale_x_continuous("Generation time (years)")
+	+ scale_y_continuous("Density (per year)", expand=c(0, 0))
+	+ ggtitle("A")
+	+ theme(
+		legend.position=c(0.9, 0.9),
+		legend.title=element_blank(),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+)
+grid.arrange(g1, g1, nrow=1)

### 2020-10-09 code/tt.R

@@ -8,7 +8,7 @@ commandEnvironments()
 makeGraphics(width=8, height=6)
 
 maxRate <- 1/6
-h_base <- HIVgen(earlyProp = 0.23, step=0.1)
+h_base <- HIVgen(earlyProp = earlyBase, step=0.1)
 
 b_base <- b0fun(h_base, rfitmonth)
 
@@ -26,7 +26,7 @@ g1 <- ggplot(tt) +
 		axis.line = element_line()
 	)
 
-early <- seq(from=0.1, to=0.4, by=0.01)
+early <- seq(from=earlySlow, to=earlyFast, by=0.01)
 
 R0 <- sapply(early, function(x) {
 	hh <- HIVgen(earlyProp = x, step=0.1)
@@ -52,8 +52,8 @@ g2 <- ggplot(earlydata) +
 	geom_line(aes(early, R0, col="Epidemic")) +
 	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
 	scale_color_manual(values=c("black", "red")) +
-  geom_point(data=earlydata[earlydata$early==0.23,], aes(early, R0), size=5) +
-  geom_point(data=strengthdata[strengthdata$early==0.23,], aes(early, strength), size=5, col="red") +
+  geom_point(data=earlydata[earlydata$early==earlyBase,], aes(early, R0), size=5) +
+  geom_point(data=strengthdata[strengthdata$early==earlyBase,], aes(early, strength), size=5, col="red") +
 	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
 	scale_y_log10("Strength", limits=c(1, 16), expand=c(0, 0), breaks=c(1, 2, 4, 8, 16)) +
 	ggtitle("B") +
@@ -92,8 +92,8 @@ speeddata <- data.frame(
 g4 <- ggplot(speeddata) +
 	geom_line(aes(early, speed, col="Intervention")) +
 	geom_hline(aes(yintercept=rfityear, col="Epidemic" )) +
-  geom_point(data=speeddata[speeddata$early==0.23,], aes(early, speed), size=5, col="red") +
-  geom_point(x=0.23, y=rfityear, size=5) +
+  geom_point(data=speeddata[speeddata$early==earlyBase,], aes(early, speed), size=5, col="red") +
+  geom_point(x=earlyBase, y=rfityear, size=5) +
[…]

### 2020-10-10 speedstrength.tex
done for now

@@
 
+\swp{Suggestion for the entire manuscript: I feel like it could be easier to read if we keep the ordering of (speed and strength) or (strength and speed).
+I wonder if we should says strength and speed, given that we almost always introduce strength first and then there's a corresponding speed?
+Not necessary but just a thought...
+If you think this is helpful, I can do this in no time.}
 An epidemic can be described by its \emph{speed} and \emph{strength}.
@@
 
-Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population -- called the \emph{basic} reproductive number \Ro\, --- because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
+Much research has prioritized estimates of \RR, and particularly its value in a fully susceptible population --- called the \emph{basic} reproductive number \Ro\, --- because \RR\ has a threshold value (i.e., $\RR=1$) that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts \citep{anderson1991infectious, diekmann1990definition}.
 The insight that a case must on average cause at least one new case under good conditions for a disease to persist goes back $>100$ years \citep{ross1911prevention};
@@
 We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the fact that the threshold $\RR=1$ determines the effective strength of intervention required for disease elimination, and the fact that the threshold $r=0$ determines the effective speed of intervention required for elimination.
+\swp{This sentence is a bit long. How about we break it into two sentences?}
+\swp{
+We then show that we can likewise measure an intervention's ``speed'' on the same scale as the growth rate.
+There is a duality between the thresholds $\RR=1$ and $r=0$, which determine the effective strength and speed of intervention required for disease elimination.
+}
 We argue that the primacy of $\RR$ over $r$ is partly due to history, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
@@
 
-Imagine a control measure that proportionally reduces $K$, for example, by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
+Imagine a control measure that proportionally reduces $K$ \swp{by a factor of $\theta$}, for example, by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A).
+We then have:
 \begin{equation}
@@
 \end{equation}
-
 Again, we have that the disease cannot spread when $\theta \geq \RR$. 
 In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, this must exceed the strength of the disease (\RR) to achieve elimination.
+
+\swp{
+We note that intervention function $L$ and the strength of intervention $\theta$ do not need to be calculated explicitly to determine elimination in many contexts---rather, they are abstractions of the current epidemic modeling practices.
+Modelers typically rely on mechanistic models (often using ordinary differential equations models) to model disease spread and try to determine intervention parameters, such as vaccination proportion, to prevent an outbreak.
+By doing so, they make implicit assumptions about the shape of $L$ and therefore $\theta$.
+In these cases, the strength of intervention corresponds to the ratio of the reproduction numbers for the null model without intervention and the model with intervention.
+}
+\swp{Need a concluding sentence?}
 
@@
 \end{equation}
-We define the speed of a general intervention to be $\phi = r - \hat r$. 
+We define the speed of a general intervention to be $\phi = r - \hat r$
 We can then show that $\phi$ is a (sort of) mean satisfying the implicit equation:
@@
 Specifically, the speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.
+\swp{
+Like intervention strength $\theta$, intervention speed $\phi$ is also an abstraction---
+that is, the mechanistic models of interventions make implicit assumptions about the shape of the hazard rate $h$ and therefore $\phi$.
+The speed of intervention then corresponds to the difference between the observed growth rate and the growth rate of a simulated epidemic under intervention.
+}
 
@@
 A generation-interval perspective \cite{park2019practical} can be used to challenge this view. 
-More early transmission, for a given value of \rr, means less transmission per individual -- that is, a lower value of \RR.
+More early transmission, for a given value of \rr, means \swp{shorter generation intervals and therefore} less transmission per individual -- that is, a lower value of \RR.
 Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary.
@@
 
-For interventions which focus on infected individuals, like contact tracing or test-based isolation, the speed-based paradigm provides clearer insight into the likely effects of early transmission. 
+For interventions which \swp{target} infected individuals, like contact tracing or test-based isolation, the speed-based paradigm provides clearer insight into the likely effects of early transmission.
 More early transmission does not change our estimate of the speed of the epidemic, but it does change our estimate of the \emph{effective} speed of a given control curve. 
@@
 Thus, if the hazard of control is increasing through time, as might be expected with isolation of symptomatic cases, more early transmission means that the effective speed of control is lower than originally thought, and control will be harder.
-Conversely, if hazard of control decreases through time, as might be expected with isolation of returning travellers, or event-based contact tracing, more early transmission means that control will be easier than originally though.
+Conversely, if the hazard of control decreases through time, as might be expected with isolation of returning travellers, or event-based contact tracing, more early transmission means that control will be easier than originally thought.
+\swp{I think this part needs to be explained more clearly.
+If hazard decreases through time, then the intervention focuses on reducing early transmission (whereas before, we try to reduce late transmission).
+So if there is more early transmission, than easier to control.
+I also wonder if it's better to frame it in terms of symptom-based intervention vs non-symptom-based?
+Increasing/decreasing hazard may be a bit vague.}
 
@@
 We also discussed the question of uncertainties introduced by the unknown magnitude of pre-symptomatic transmission of COVID-19.
-We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while speed-based paradigm is a clearer way to think about interventions targeted to infected people.
+We showed that the strength-based paradigm provides clear insight into how this uncertainty affects \swp{non-targeted} interventions for the general population, while speed-based paradigm is a clearer way to think about interventions targeted to infected people.
 
 While both speed- and strength-based frameworks can give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of a given measure.
-For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
+For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, \swp{\rr can be directly observed from case data but} the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
 Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
+
 For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that target infected people, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}; we expect the strength-based framework to be clearer for intervention strategies that target the general population, like condom use, or susceptible people, like prophylaxis.
@@
 
-When comparing interventions with epidemic parameters to evaluate strategies, the situation is similar.
+When comparing interventions with \swp{uncertainties in} epidemic parameters to evaluate strategies, the situation is similar.
 Some scenarios lend themselves naturally to a single approach.
@@
 With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
-But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result.
+But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result\swp{---the overall conclusion can change whether we calibrate our models to \rr or \RR.}
+\swp{I feel like this is important but I'm not sure if this is a good place to put it. I think we need to say it explicitly.}
 
@@
 but when regulation primarily affects juvenile mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily}.
+\swp{Do we need to explain more? feels a bit abrupt.}
 

### 2020-10-12 code/hivGens.R
downing

@@ -11,16 +11,18 @@ base = HIVgen(earlyProp = earlyBase)
 fast = HIVgen(earlyProp = earlyFast)
 slow = HIVgen(earlyProp = earlySlow)
 
-gens <- (
+factor_order <- c("base", "slow", "fast")
+g <- (
 	list(
 		base = base
 		, fast = fast
 		, slow = slow
 	)
 	%>% bind_rows(.id = "type")
+	%>% mutate(type = factor(type, levels=factor_order))
 )
 
-g1 <- (ggplot(gens)
+g1 <- (ggplot(g)
 	+ geom_line(aes(time*month/year, density*year/month, color=type))
 	+ scale_x_continuous("Generation time (years)")
 	+ scale_y_continuous("Density (per year)", expand=c(0, 0))

### 2020-10-12 code/HIVfuns.R
generation pictures

@@ -6,8 +6,8 @@ earlyBase <- 0.23
 earlySlow <- 0.1
 
 testingFun <- function(maxRate
-					   , hMean=12, hShape=2, hheight=4
-					   , step = 0.1, window=200
+	, hMean=12, hShape=2, hheight=4
+	, step = 0.1, window=200
 ){
 	time <- seq(step, window, by=step)
 	tHaz0 <- dgamma(time, shape=hShape, scale=hMean/hShape)

### 2020-10-12 code/hivGens.R

@@ -7,11 +7,13 @@ source("makestuff/makeRfuns.R")
 commandEnvironments()
 makeGraphics(width=8, height=3)
 
-base = HIVgen(earlyProp = earlyBase)
-fast = HIVgen(earlyProp = earlyFast)
-slow = HIVgen(earlyProp = earlySlow)
+legend_order <- c("base", "slow", "fast")
+months = 120
 
-factor_order <- c("base", "slow", "fast")
+base = HIVgen(earlyProp = earlyBase, window=months)
+fast = HIVgen(earlyProp = earlyFast, window=months)
+slow = HIVgen(earlyProp = earlySlow, window=months)
+
 g <- (
 	list(
 		base = base
@@ -19,14 +21,34 @@ g <- (
 		, slow = slow
 	)
 	%>% bind_rows(.id = "type")
-	%>% mutate(type = factor(type, levels=factor_order))
+	%>% mutate(type = factor(type, levels=legend_order))
 )
 
-g1 <- (ggplot(g)
+g <- (
+	list(
+		base = base
+		, fast = fast
+		, slow = slow
+	)
+	%>% bind_rows(.id = "type")
+	%>% mutate(type = factor(type, levels=legend_order))
+)
+
+b <- (
+	list(
+		base = b0fun(base, rfitmonth)
+		, fast = b0fun(fast, rfitmonth)
+		, slow = b0fun(slow, rfitmonth)
+	)
+	%>% bind_rows(.id = "type")
+	%>% mutate(type = factor(type, levels=legend_order))
+)
+
+gplot <- (ggplot(g)
 	+ geom_line(aes(time*month/year, density*year/month, color=type))
 	+ scale_x_continuous("Generation time (years)")
 	+ scale_y_continuous("Density (per year)", expand=c(0, 0))
-	+ ggtitle("A")
+	+ ggtitle("Intrinsic generation interval")
 	+ theme(
 		legend.position=c(0.9, 0.9),
 		legend.title=element_blank(),
@@ -35,4 +57,19 @@ g1 <- (ggplot(g)
 		axis.line = element_line()
 	)
 )
-grid.arrange(g1, g1, nrow=1)
+
+bplot <- (ggplot(b)
+	+ geom_line(aes(time*month/year, density*year/month, color=type))
+	+ scale_x_continuous("Generation time (years)")
+	+ scale_y_continuous("Density (per year)", expand=c(0, 0))
+	+ ggtitle("Backward generation interval")
[…]

### 2020-10-12 speedstrength.tex
Going through swps

@@
 
-\newcommand{\comment}{RENEW}
+\newcommand{\comment}{RENEW the comment command}
+\renewcommand{\comment}[3]{}
 \renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
-\renewcommand{\comment}[3]{}
 
@@
 We first generalize the idea that \RR\ measures the difficulty of elimination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number. 
-We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the fact that the threshold $\RR=1$ determines the effective strength of intervention required for disease elimination, and the fact that the threshold $r=0$ determines the effective speed of intervention required for elimination.
-\swp{This sentence is a bit long. How about we break it into two sentences?}
-\swp{
 We then show that we can likewise measure an intervention's ``speed'' on the same scale as the growth rate.
-There is a duality between the thresholds $\RR=1$ and $r=0$, which determine the effective strength and speed of intervention required for disease elimination.
-}
+Thus, there is a dual sense in which \rr\ also measures difficulty of elimination. 
 We argue that the primacy of $\RR$ over $r$ is partly due to history, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
@@
 
-Imagine a control measure that proportionally reduces $K$ \swp{by a factor of $\theta$}, for example, by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A).
-We then have:
+Imagine a control measure that proportionally reduces $K$ by a factor of $\theta$, for example, by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
 \begin{equation}
@@
 We note that intervention function $L$ and the strength of intervention $\theta$ do not need to be calculated explicitly to determine elimination in many contexts---rather, they are abstractions of the current epidemic modeling practices.
-Modelers typically rely on mechanistic models (often using ordinary differential equations models) to model disease spread and try to determine intervention parameters, such as vaccination proportion, to prevent an outbreak.
+Modelers typically rely on mechanistic models (often based on ordinary differential equations) to model disease spread and try to determine intervention parameters, such as vaccination proportion, to prevent an outbreak.
 By doing so, they make implicit assumptions about the shape of $L$ and therefore $\theta$.

### 2020-10-12 speedstrength.tex
first time through comments

@@
 
-An epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproductive number \RR).
+An epidemic can be characterized by its  strength (i.e., the reproductive number \RR) and speed (i.e., the exponential growth rate $r$).
 Disease modelers have historically placed much more emphasis on strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
 Here, we develop a mathematical framework for the classic, strength-based paradigm and show that there is a dual speed-based paradigm which can provide complementary insights.
-In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$ \cite{dhmKermack}, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
+In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$ \cite{dhmKermack}, and show that we can measure the strength and speed of an intervention on the same scale as the strength and speed of an epidemic, respectively.
 We argue that, while the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
-As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
+As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the strength than speed scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
 We also discuss evaluating the effects of the importance of pre-symptomatic transmission of the SARS-Cov-2 virus.
@@
 
-\swp{Suggestion for the entire manuscript: I feel like it could be easier to read if we keep the ordering of (speed and strength) or (strength and speed).
-I wonder if we should says strength and speed, given that we almost always introduce strength first and then there's a corresponding speed?
-Not necessary but just a thought...
-If you think this is helpful, I can do this in no time.}
-An epidemic can be described by its \emph{speed} and \emph{strength}.
+An epidemic can be described by its \emph{strength} and \emph{speed}.
+The strength of an epidemic is characterized by the reproductive number \RR, which measures \emph{how many} new cases are caused by a typical individual case.
 The speed of an epidemic is characterized by the exponential growth rate $r$, which measures how \emph{fast} an epidemic grows at the population level.
-The strength of an epidemic is characterized by the reproductive number \RR, which measures \emph{how many} new cases are caused by a typical individual case.
-Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.
+Knowing the strength and speed of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.
 
@@
 }
-\swp{Need a concluding sentence?}
 
@@
 
-We compare two different possible intervention strategies to shed light on the speed and strength decompositions.
+We compare two different possible intervention strategies to shed light on the strength and speed decompositions.
 First, we consider a condom intervention that reduces HIV transmission by an average of approximately 75\%.
@@
 A generation-interval perspective \cite{park2019practical} can be used to challenge this view. 
-More early transmission, for a given value of \rr, means \swp{shorter generation intervals and therefore} less transmission per individual -- that is, a lower value of \RR.
+More early transmission, for a given value of \rr, means {shorter generation intervals and therefore} less transmission per individual -- that is, a lower value of \RR.
 Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary.
@@
 
-For interventions which \swp{target} infected individuals, like contact tracing or test-based isolation, the speed-based paradigm provides clearer insight into the likely effects of early transmission.
+For interventions which {target} infected individuals, like contact tracing or test-based isolation, the speed-based paradigm provides clearer insight into the likely effects of early transmission. 
 More early transmission does not change our estimate of the speed of the epidemic, but it does change our estimate of the \emph{effective} speed of a given control curve. 
@@
 
-While both speed- and strength-based frameworks can give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of a given measure.
-For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, \swp{\rr can be directly observed from case data but} the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
+While both strength- and speed-based frameworks can give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of a given measure.
+For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, {\rr\ can be directly observed from case data but} the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
 Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
@@
 For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that target infected people, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}; we expect the strength-based framework to be clearer for intervention strategies that target the general population, like condom use, or susceptible people, like prophylaxis.
-In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.
+In other cases, such as real-time rollout of vaccines during an outbreak, both strength and speed approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.
 
-When comparing interventions with \swp{uncertainties in} epidemic parameters to evaluate strategies, the situation is similar.
+When comparing proposed interventions with estimated epidemic parameters to evaluate strategies, the situation is similar.
 Some scenarios lend themselves naturally to a single approach.
@@
 With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
-But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result\swp{---the overall conclusion can change whether we calibrate our models to \rr or \RR.}
+But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result
+%% \jd{What does this comment mean?}
+\swp{---the overall conclusion can change whether we calibrate our models to \rr or \RR.}
 \swp{I feel like this is important but I'm not sure if this is a good place to put it. I think we need to say it explicitly.}

### 2020-10-13 speedstrength.tex
comments on the review & new title
power

@@
 	\textbf\newline{
-		Speed and strength of an epidemic intervention
+		Strength and speed of an epidemic intervention
 	}
@@
 \section*{Abstract}
+
+\swp{I also changed the title to strength and speed}
 
@@
 
+\swp{A new suggested paragraph.
+I think some version of this paragraph may help deal with above sentence (the image you sent me saying that you don't understand).
+I think it can also help deal with the reviewer's comments about the growth phase}
+\swp{
+Here, we focused on the initial growth phase an epidemic, assuming that the observed growth rate \rr is fixed throughout our analysis.
+For endemic diseases, as discussed previously, strength-based framework can provide a clearer perspective---therefore, it makes more sense to rely on fixed \RR when evaluating the difficulty of elimination.
+In these cases, parameter uncertainty will generally not affect the efficacy of strength-like interventions but can affect conclusions about speed-lik interventions:
+given \RR, shorter generation intervals imply faster epidemic \citep{eaton2014proportion, powers2014impact, park2019practical}, requiring faster intervention to control .
+While both strength- and speed-based frameworks will give the same answers for a given set of assumptions about an epidemic,
+these conclusions can depend on whether we calibrate our models to the strength or speed of an epidemic.
+}
+
 There is an analogy here with measures of fitness in theoretical ecology.
@@
 For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
+\swp{does this feel weak now?}
 We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.

### 2020-10-13 speedstrength.tex
one omre

@@
 For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
-\swp{does this feel weak now?}
+\swp{does this feel weak now?
+Maybe need something like ``Many subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing, with a few exceptions'' with citations?}
 We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.

### 2020-10-13 prsb_responses.tex
archiving prsb_responses

SWP: I prefer to use Thank you to be extra polite.

\hypertarget{referee-1}{%
\subsection{Referee: 1}\label{referee-1}}

\begin{quote}\sl
One additional reference that might be worth mentioning is O. Diekmann,
H. Heesterbeek, and H. Metz,The legacy of Kermack and McKendrick, in
Epidemic Models, Their Structure and Relation to Data (D. Mollinson,
ed.) (1995), pp.95-115.
\end{quote}

Thank you. We now cite this article as an early mentioner of the
little-r threshold.

\hypertarget{referee-2}{%
\subsection{Referee: 2}\label{referee-2}}

\begin{quote}\sl
In this work the authors propose that ``strength'' and ``speed'' are
complementary frameworks for understanding interventions in infectious
disease models. This is an interesting and appealing idea based on
simple re-arrangements of well-known equations for exponentially growing
epidemics. As the authors themselves note, comparing growth rates r
rather than reproductive numbers R is already well known in ecology.
\end{quote}

\begin{quote}\sl
In my view this paper could bring these ideas to light in infectious
disease more thoroughly and compellingly with an expanded application
section, a section on the relationship to data and/or estimation from
data, improved and expanded section on interventions and the distinct
views provided by (and usefulness of) the two distinct frameworks,
and/or with some extension of the results to population dynamics that
differ from constant rate exponential growth or decay. As it stands it
feels quite minimal, though I appreciate the appeal of the conceptual
point.
\end{quote}

Thank you for this feedback. We agree that these ideas are similar to
ideas that have been discussed in the ecological context, but also feel
that this is a perspective sorely lacking in the disease-modeling world.

We have expanded the application section with examples from COVID-19. To
our mind, these examples further underline the absence of r-based
thinking in the disease-modeling world. We also address more explicitly
the link between the parameters of the initial phase and the prospects
for long-term control. We feel that issues dealing directly with
non-exponential disease dynamics are complicated enough for another
paper.

SWP: I feel like we need a better defense for why we are being
qualitative, rather than quantitative. We should also be more concrete
about our argument. Here's a suggestion. SWP: We have expanded the
application section with examples from COVID-19, providing a qualitative
understanding on how uncertainties in the amount of presymptomatic
transmission can affect conclusions about the efficacy of intervention.
In the beginning of the current pandemic, there was much discussion on
how presymptomatic transmission could make intervention more
difficult---however, some of these conclusions were based on the
assumption that R is constant, rather than r (e.g., Hellewell et
al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
of cases and contacts.'' The Lancet Global Health. In reality, given
observed r, more presymptomatic transmission reduces R, meaning that
strength-based interventions will be easier than previously thought. To
our mind, these examples further underline the absence of r-based
thinking in the disease-modeling world. While we appreciate the
reviewer's concern that a more data-driven approach could provide
further insight, we are worried that details may take away general ideas
behind speed- and strength-based paradigms. We feel that it is important
to provide a conceptual framework at this stage to a broader audience.
We also address more explicitly the link between the parameters of the
initial phase and the prospects for long-term control. We feel that
issues dealing directly with non-exponential disease dynamics are
complicated enough for another paper.

\begin{quote}\sl
Near equation (9) the authors need to make a more explicit link with
survival analysis so that we are clear on why we multiply K by survival
functions. Survival functions multiply in the sense that if there are
two hazards at rates lambda1 and lambda2 then the survival function is
S(t) = S1(t) S2(t) . But K(tau) is proportional to the density g(tau),
and is not a survival function. What is the interpretation of the
density times the survival function? Why would the interventions not
change the density directly, so that g becomes ghat (and then this would
impact the hazard rate and survival function via the usual relationship
S(t) = \_t \^{}( ghat (x) dx ) ) ?
\end{quote}

We have expanded (9) into two steps and attempted to explain the logic
clearly. SWP: Can we change this into a tex file so we can actually add
equations?? Or is it OK to leave it like this??

\begin{quote}\sl
In practice what is the need for equations (6) and (12)? Both
essentially re-write the other equations; are these averages that can be
linked to estimates from data, for example?
\end{quote}

We think that the value of (6) and (12 -- now 13) is conceptual. We have
added explanations following these derivations. In retrospect, we
believe that Ss 2.2 and 2.3 were written in too ``mathematical'' a
style: we got a bit carried away with the formal parallelism. The new
version is still short, but engages more with the biological concepts.

SWP: We should note that in practice 6 and 12 are done implicitly.
People use ODEs to model interventions, which make implicit assumptions
about the shape of L and h, and therefore are already implicitly doing 6
and 12.

\begin{quote}\sl
Presumably both frameworks give a decline when the growth rate is
negative and growth when it is positive. Therefore they would never
contradict each other as to whether an intervention was sufficient or
not - is this correct? This should be made explicit.
\end{quote}

This is correct, and we have worked to be more explicit. SWP: Where? Can
we actually cite what we changed?

\begin{quote}\sl
It would also be good to more explicitly state and discuss what the
differences are. I realize that this is in the discussion and in some
sense in Figure 4, but I found the discussion section vague,
particularly since the analysis is only relevant for the case of
exponential growth. In particular, where would the two frameworks give
different estimates of the uncertainty surrounding the benefits of an
intervention, or the comparison between two interventions?
\end{quote}

We never expect to get a different final answer from the two formalisms;
they are two ways of doing essentially the same calculation. We have
tried to be more clear about this as mentioned above. We do expect in
some cases to be able to use one formalism or another to more clearly
illuminate the key issues and quantities determining the detailed
answer, as we try to explain in our HIV example and in our new COVID
example.

SWP: However, we do expect to get different answers whether we assume r
or R is fixed for a given set of scenarios. For endemic diseases, such
as measles, it makes more sense to assume that R is fixed. For emerging
diseases, such COVID-19, it makes more sense to assume that r is
fixed---this is often overlooked approach in the modeling community. We
now discuss this in the paper.

\begin{quote}\sl
On a minor note, since condoms are ultimately an individual choice
applied to individual (potential) transmission events, why are they a
population-level intervention whereas test and treat is deemed
individual? More details on why the two frameworks give different
intuition or analytical capacity would be good.
\end{quote}

Thank you for this. We were using individual to mean targeted at
infectious individuals, and population to mean targeted at everyone. We
have now dropped these shortcuts, and we simply explain this directly.

SWP: I like this better for individual vs population-based: The efficacy
of individual-based intervention depends on time since infection (e.g.,
more likely to be isolated during symptomatic stage). The efficacy of
population-based intervention doesn't (e.g., the relative efficacy of
condom doesn't really change). What do you think?

\begin{quote}\sl
In Figures 3 and 4, the black lines are not clear to me -- the strength
is defined as R/Rhat and the speed as rhat -r . So what is the strength
(or speed) of ``the epidemic'' (black lines) - what's Rhat and what's R?
It might be good to have some prevalence curves
\end{quote}

We define strength and speed of the epidemic in the introduction, and
compare these to strength and speed of intervention efforts. We now
reiterate these points (including in the figure captions).

\begin{quote}\sl
Minor quibble: equation (4) has L described in words as the ``average
proportional reduction'' but where the reduction is 75\%, in fact the
denominator is 4 (1 / (1-0.75) ) -- rephrase.
\end{quote}

Done, thank you.

### 2020-10-13 responses.tex
archiving prsb_responses

--- 2020-10-13 prsb_responses.tex
@@
-SWP: I prefer to use Thank you to be extra polite.
-
 \hypertarget{referee-1}{%
@@
 paper.
-
-SWP: I feel like we need a better defense for why we are being
-qualitative, rather than quantitative. We should also be more concrete
-about our argument. Here's a suggestion. SWP: We have expanded the
+We have expanded the
 application section with examples from COVID-19, providing a qualitative
@@
 impact the hazard rate and survival function via the usual relationship
-S(t) = \_t \^{}( ghat (x) dx ) ) ?
+S(t) = \int_t ^\infty ( ghat (x) dx )  ) ?
 \end{quote}
@@
 We have expanded (9) into two steps and attempted to explain the logic
-clearly. SWP: Can we change this into a tex file so we can actually add
+clearly. 
 equations?? Or is it OK to leave it like this??
@@
 
-We never expect to get a different final answer from the two formalisms;
-they are two ways of doing essentially the same calculation. We have
-tried to be more clear about this as mentioned above. We do expect in
-some cases to be able to use one formalism or another to more clearly
-illuminate the key issues and quantities determining the detailed
-answer, as we try to explain in our HIV example and in our new COVID
-example.
+We never expect to get a different final answer from the two formalisms if they are both correctly applied;
+they are two ways of doing essentially the same calculation. 
+We have tried to be more clear about this as mentioned above. 
+We do expect in some cases to be able to use one formalism or another to more clearly illuminate the key issues and quantities determining the detailed answer, as we try to explain in our HIV example and in our new COVID example.
 
-SWP: However, we do expect to get different answers whether we assume r
-or R is fixed for a given set of scenarios. For endemic diseases, such
-as measles, it makes more sense to assume that R is fixed. For emerging
-diseases, such COVID-19, it makes more sense to assume that r is
-fixed---this is often overlooked approach in the modeling community. We
-now discuss this in the paper.
+Keeping the two paradigms in mind, however, can help to do the analysis correctly. 
+In particular, researchers should give careful thought to which quantities to fix when considering scenarios for disease spread. 
+In particular, in emerging epidemics, researchers often vary assumptions about parameters while holding R constant, even in cases where it would be more consistent with available information to hold \rr\ constant.
+We now discuss this in the paper.
 
@@
 condom doesn't really change). What do you think?
+JD: This is exactly the kind of detail I'm avoiding at this point. We can get back to at in the next paper.
 

### 2020-10-13 responses.tex
Meetings

@@
+\newcommand{\RR}{\ensuremath{{\mathcal R}}}
+\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
+\newcommand{\rr}{\ensuremath{{r}}}
+
 \hypertarget{referee-1}{%
@@
 impact the hazard rate and survival function via the usual relationship
-S(t) = \int_t ^\infty ( ghat (x) dx )  ) ?
+$S(t) = \int_t ^\infty ( ghat (x) dx )  )$ ?
 \end{quote}
@@
 In particular, researchers should give careful thought to which quantities to fix when considering scenarios for disease spread. 
-In particular, in emerging epidemics, researchers often vary assumptions about parameters while holding R constant, even in cases where it would be more consistent with available information to hold \rr\ constant.
+In particular, in emerging epidemics, researchers often vary assumptions about parameters while holding \RR\ constant, even in cases where it would be more consistent with available information to hold \rr\ constant.
 We now discuss this in the paper.
@@
 condom doesn't really change). What do you think?
-JD: This is exactly the kind of detail I'm avoiding at this point. We can get back to at in the next paper.
+JD: This is exactly the kind of detail I'm avoiding at this point. We can get back to it in the next paper?
 
@@
 proportional reduction'' but where the reduction is 75\%, in fact the
-denominator is 4 (1 / (1-0.75) ) -- rephrase.
+denominator is 4: $(1 / (1-0.75) )$ -- rephrase.
 \end{quote}

### 2020-10-13 speedstrength.tex
Meetings

@@
 \section*{Abstract}
-
-\swp{I also changed the title to strength and speed}
 
@@
 
-\swp{
-We note that intervention function $L$ and the strength of intervention $\theta$ do not need to be calculated explicitly to determine elimination in many contexts---rather, they are abstractions of the current epidemic modeling practices.
-Modelers typically rely on mechanistic models (often based on ordinary differential equations) to model disease spread and try to determine intervention parameters, such as vaccination proportion, to prevent an outbreak.
+We note that intervention function $L$ and the strength of intervention $\theta$ need not be calculated explicitly in many contexts: they can usefully be thought of as abstractions of existing modeling practices.
+Modelers typically rely on mechanistic models (often based on ordinary differential equations) to model disease spread and evaluate intervention effects.
 By doing so, they make implicit assumptions about the shape of $L$ and therefore $\theta$.
-In these cases, the strength of intervention corresponds to the ratio of the reproduction numbers for the null model without intervention and the model with intervention.
-}
 
@@
 Specifically, the speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.
-\swp{
-Like intervention strength $\theta$, intervention speed $\phi$ is also an abstraction---
-that is, the mechanistic models of interventions make implicit assumptions about the shape of the hazard rate $h$ and therefore $\phi$.
-The speed of intervention then corresponds to the difference between the observed growth rate and the growth rate of a simulated epidemic under intervention.
-}
+Like intervention strength $\theta$, intervention speed $\phi$ is also an abstraction -- that is, the mechanistic models of interventions make implicit assumptions about the shape of the hazard rate $h$ and therefore $\phi$.
 
@@
 This is because more early transmission gives more weight in the calculation of effective speed ($\phi$) to the early period of infection.
-Thus, if the hazard of control is increasing through time, as might be expected with isolation of symptomatic cases, more early transmission means that the effective speed of control is lower than originally thought, and control will be harder.
-Conversely, if the hazard of control decreases through time, as might be expected with isolation of returning travellers, or event-based contact tracing, more early transmission means that control will be easier than originally thought.
-\swp{I think this part needs to be explained more clearly.
-If hazard decreases through time, then the intervention focuses on reducing early transmission (whereas before, we try to reduce late transmission).
-So if there is more early transmission, than easier to control.
-I also wonder if it's better to frame it in terms of symptom-based intervention vs non-symptom-based?
-Increasing/decreasing hazard may be a bit vague.}
+
+If control depends on identifying cases from symptoms, we might expect the hazard of control to increase through time.
+In this case, more early transmission will give more weight to the low effectiveness of control early in infection: thus, the effective speed of control will be lower than originally thought, and control will be harder.
+Conversely, if control is mostly done by tracing travellers and contacts, cases might be most likely to be identified early on: in this case more weight on the early hazard of control will mean a faster effective speed.
+Thus, control will be easier than originally thought.
 

### 2020-10-14 responses.tex
good night

@@
 \subsection{Referee: 1}\label{referee-1}}
+
+JD: I moved this to tex, and then figured out that the only equation here is from the reviewer, which makes it awkward to edit.
 
@@
 
-SWP: We should note that in practice 6 and 12 are done implicitly.
-People use ODEs to model interventions, which make implicit assumptions
-about the shape of L and h, and therefore are already implicitly doing 6
-and 12.
+We note that in practice 6 and 12 are often implicit in mechanistic models. We now make this point in the MS near the end of both S2.2 and S2.3.
 
@@
 
-This is correct, and we have worked to be more explicit. SWP: Where? Can
-we actually cite what we changed?
+This is correct, and we have worked to be more explicit.
 
@@
 infectious individuals, and population to mean targeted at everyone. We
-have now dropped these shortcuts, and we simply explain this directly.
+have now dropped these shortcuts, and we simply explain this directly. We have also added more discussion of how frameworks differ, both in examples and in Discussion.
 

### 2020-10-14 speedstrength.tex
good night

@@
 We argue that, while the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
-As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the strength than speed scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
+As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection.
 We also discuss evaluating the effects of the importance of pre-symptomatic transmission of the SARS-Cov-2 virus.
@@
 In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ --- notably, the proportion of the population susceptible, $S(t)$.
+
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time.
@@
 Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
+Note that this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to increase under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
 
@@
 therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
-Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\Ro$ is thus the integral of $K$.
+Since $g$ is a distribution, it integrates to 1, and the basic reproductive number \RR\ is thus the integral of $K$.
 
@@
 
-\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}
-\jd{The above is an old comment. I'm not currently loving the idea of including that figure, but I guess we should make it and see?}
-
 \begin{figure}[!t]
@@
 The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
-For the test-and-treat intervention, the effective intervention speed also stays relatively constant (\figref{test}D), in part because we have (plausibly) assumed that hazard stays relatively constant for a few key months, and in part because the backward generation-interval distributions for different scenarios are relatively similar.
+For the test-and-treat intervention, the effective intervention speed also stays relatively constant (\figref{test}D), in part because we have (plausibly) assumed that hazard stays relatively constant for a few key months, and in part because the backward generation-interval distributions for different scenarios are relatively similar (Supplemental figure).
 The effective intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
@@
 But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result
-%% \jd{What does this comment mean?}
 \swp{---the overall conclusion can change whether we calibrate our models to \rr or \RR.}

### 2020-10-14 responses.tex
Hand back to Daniel

@@
 
+\newcommand{\comment}{RENEW the comment command}
+\renewcommand{\comment}[3]{}
+\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
+\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
+\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
+
 \hypertarget{referee-1}{%
@@
 
-JD: I moved this to tex, and then figured out that the only equation here is from the reviewer, which makes it awkward to edit.
+\jd{I moved this to tex, and then figured out that the only equation here is from the reviewer, which makes it awkward to edit.}
 
@@
 We have expanded the application section with examples from COVID-19. To
-our mind, these examples further underline the absence of r-based
+our mind, these examples further underline the absence of \rr-based
 thinking in the disease-modeling world. We also address more explicitly
@@
 difficult---however, some of these conclusions were based on the
-assumption that R is constant, rather than r (e.g., Hellewell et
+assumption that \RR\ is constant, rather than \rr\ (e.g., Hellewell et
 al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
-of cases and contacts.'' The Lancet Global Health. In reality, given
-observed r, more presymptomatic transmission reduces R, meaning that
+of cases and contacts.'' \emph{ Lancet Global Health}. In reality, given
+observed \rr, more presymptomatic transmission reduces \RR, meaning that
 strength-based interventions will be easier than previously thought. To
-our mind, these examples further underline the absence of r-based
+our mind, these examples further underline the absence of \rr-based
 thinking in the disease-modeling world. While we appreciate the
@@
 
-SWP: I like this better for individual vs population-based: The efficacy
+\swp{I like this better for individual vs population-based: The efficacy
 of individual-based intervention depends on time since infection (e.g.,
@@
 population-based intervention doesn't (e.g., the relative efficacy of
-condom doesn't really change). What do you think?
-JD: This is exactly the kind of detail I'm avoiding at this point. We can get back to it in the next paper?
+condom doesn't really change). What do you think?}
+\jd{This is exactly the kind of detail I'm avoiding at this point. We can get back to it in the next paper?}
 

### 2020-10-14 speedstrength.tex
Hand back to Daniel

@@
 \renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
-
 \newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
@@
 We also discussed the question of uncertainties introduced by the unknown magnitude of pre-symptomatic transmission of COVID-19.
-We showed that the strength-based paradigm provides clear insight into how this uncertainty affects \swp{non-targeted} interventions for the general population, while speed-based paradigm is a clearer way to think about interventions targeted to infected people.
+We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while the speed-based paradigm is a clearer way to think about interventions targeted to infected people.
+We concluded that increasing pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions.
+In hindsight, these conclusions are consistent with common sense, but in practice the speed-based framework provides a clear way to think about these questions.
 
 While both strength- and speed-based frameworks can give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of a given measure.
-For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, {\rr\ can be directly observed from case data but} the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
+For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, \rr\ can be directly observed from case data but the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
 Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
+
+Thinking explicitly about the two perspectives can also reduce confusion.
+Because of the dominance of the strength paradigm, researchers often explore different scenarios while holding \RR\ fixed.
+Fixed \RR\ is in fact a good default assumption for many endemic diseases.
+For invading diseases, however, \rr\ is likely to be better constrained by data than \RR.
+In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value) \citep{eaton2014proportion, powers2014impact, park2019practical}.
 
@@
 With sufficiently detailed assumptions, we could do a correct calculation for any scenario in any paradigm. 
-But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result
-\swp{---the overall conclusion can change whether we calibrate our models to \rr or \RR.}
-\swp{I feel like this is important but I'm not sure if this is a good place to put it. I think we need to say it explicitly.}
+But in many of these examples, using the more appropriate paradigm for each scenario lets us know what to expect, and may strengthen our intuition for \emph{how} the assumptions lead to the result.
 
-\swp{A new suggested paragraph.
-I think some version of this paragraph may help deal with above sentence (the image you sent me saying that you don't understand).
-I think it can also help deal with the reviewer's comments about the growth phase}
-\swp{
-Here, we focused on the initial growth phase an epidemic, assuming that the observed growth rate \rr is fixed throughout our analysis.
-For endemic diseases, as discussed previously, strength-based framework can provide a clearer perspective---therefore, it makes more sense to rely on fixed \RR when evaluating the difficulty of elimination.
-In these cases, parameter uncertainty will generally not affect the efficacy of strength-like interventions but can affect conclusions about speed-lik interventions:
-given \RR, shorter generation intervals imply faster epidemic \citep{eaton2014proportion, powers2014impact, park2019practical}, requiring faster intervention to control .
-While both strength- and speed-based frameworks will give the same answers for a given set of assumptions about an epidemic,
-these conclusions can depend on whether we calibrate our models to the strength or speed of an epidemic.
-}
-
-There is an analogy here with measures of fitness in theoretical ecology.
-For example, when a population is regulated by density dependence that affects all individuals identically, $r$ may be the best measure of fitness \citep{pasztor1996r0}, 
-but when regulation primarily affects juvenile mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily}.
-\swp{Do we need to explain more? feels a bit abrupt.}
+In population ecology, the duality between strength and speed is more widely recognized.
+For example, when a population is regulated by density dependence that affects all individuals identically, $r$ may be the best measure of fitness, 
+but when regulation primarily affects juvenile mortality, \RR\ is likely to be superior \citep{mylius1995evolutionarily,pasztor1996r0}.
+There is also a link between the duality of these perspectives and the evolutionary tradeoff between speed and strength, commonly theorized as a tradeoff between \rr\ and carrying capacity $K$ \cite{Pianka70}.
 
@@
 Maybe need something like ``Many subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing, with a few exceptions'' with citations?}
+\jd{It would be great if you could do this.}
 We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.

### 2020-10-15 response.tex
some more edits

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\rr}{\ensuremath{{r}}}
\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript. 
We have made major revisions to our manuscript to address the reviewers' comments.
We have also added a new section and discuss how presymptomatic transmission can affect the efficacy of COVID-19 interventions.
Below please find our detailed responses to reviewers.

\rev{Reviewer \#1}

\revtext{One additional reference that might be worth mentioning is O. Diekmann,
H. Heesterbeek, and H. Metz,The legacy of Kermack and McKendrick, in
Epidemic Models, Their Structure and Relation to Data (D. Mollinson,
ed.) (1995), pp.95-115.}

Thank you. We now cite this article as an early mentioner of the
little-r threshold.

\rev{Reviewer \#2}

\revtext{In this work the authors propose that ``strength'' and ``speed'' are
complementary frameworks for understanding interventions in infectious
disease models. This is an interesting and appealing idea based on
simple re-arrangements of well-known equations for exponentially growing
epidemics. As the authors themselves note, comparing growth rates r
rather than reproductive numbers R is already well known in ecology.}

\revtext{In my view this paper could bring these ideas to light in infectious
disease more thoroughly and compellingly with an expanded application
section, a section on the relationship to data and/or estimation from
data, improved and expanded section on interventions and the distinct
views provided by (and usefulness of) the two distinct frameworks,
and/or with some extension of the results to population dynamics that
differ from constant rate exponential growth or decay. As it stands it
feels quite minimal, though I appreciate the appeal of the conceptual
point.}

Thank you for this feedback. We agree that these ideas are similar to
ideas that have been discussed in the ecological context, but also feel
that this is a perspective sorely lacking in the disease-modeling world.

We have expanded the
application section with examples from COVID-19, providing a qualitative
understanding on how uncertainties in the amount of presymptomatic
transmission can affect conclusions about the efficacy of intervention.
In the beginning of the current pandemic, there was much discussion on
how presymptomatic transmission could make intervention more
difficult---however, some of these conclusions were based on the
assumption that \RR\ is constant, rather than \rr\ (e.g., Hellewell et
al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
of cases and contacts.'' \emph{ Lancet Global Health}). In reality, given
observed \rr, more presymptomatic transmission reduces \RR, meaning that
strength-based interventions will be easier than previously thought. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world. While we appreciate the
reviewer's concern that a more data-driven approach could provide
further insight, we are worried that details may take away general ideas
behind speed- and strength-based paradigms. We feel that it is important
to provide a conceptual framework at this stage to a broader audience.
We also address more explicitly the link between the parameters of the
initial phase and the prospects for long-term control. We feel that
issues dealing directly with non-exponential disease dynamics are
complicated enough for another paper.

\revtext{Near equation (9) the authors need to make a more explicit link with
survival analysis so that we are clear on why we multiply K by survival
functions. Survival functions multiply in the sense that if there are
two hazards at rates lambda1 and lambda2 then the survival function is
S(t) = S1(t) S2(t) . But K(tau) is proportional to the density g(tau),
and is not a survival function. What is the interpretation of the
density times the survival function? Why would the interventions not
change the density directly, so that g becomes ghat (and then this would
impact the hazard rate and survival function via the usual relationship
$S(t) = \int_t ^\infty ( \hat{g} (x) dx )  )$ ?}

We have expanded (9) into two steps and attempted to explain the logic clearly:

``Now imagine an idealized intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation (Fig. 1B), for example, by identifying and isolating infectious individuals.
We then have:
\begin{equation}
	\hat K(\tau) = K(\tau)\exp(-\phi\tau)
\end{equation}
The interpretation is that average infectiousness for under this control regime is the product of the original infectiousness $K(\tau)$ (at age of infection $\tau$) and the probability $\exp(-\phi\tau)$ of escaping the hazard of control up to that time.

Substituting (8):
\begin{equation}
	\hat K(\tau) = K(\tau)\exp(-\phi\tau) = b(\tau)\exp((r-\phi)\tau)
\end{equation}
Since $b$ is a distribution (which integrates to 1), the reduction needed to prevent invasion (or to eliminate disease) is exactly $\phi=r$. 
We call $\phi$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is faster than the speed of spread.''

\revtext{
In practice what is the need for equations (6) and (12)? Both
essentially re-write the other equations; are these averages that can be
linked to estimates from data, for example?
}

We think that the value of (6) and (12 -- now 13) is conceptual. We have
added explanations following these derivations. In retrospect, we
believe that S2.2 and S2.3 were written in too ``mathematical'' a
style: we got a bit carried away with the formal parallelism. The new
version is still short, but engages more with the biological concepts.

We note that in practice 6 and 12 are often implicit in mechanistic models. We now make this point in the MS near the end of both S2.2 and S2.3:

``
We note that intervention function $L$ and the strength of intervention $\theta$ need not be calculated explicitly in many contexts: they can usefully be thought of as abstractions of existing modeling practices.
Modelers typically rely on mechanistic models (often based on ordinary differential equations) to model disease spread and evaluate intervention effects.
By doing so, they make implicit assumptions about the shape of $L$ and therefore $\theta$.
''

``Like intervention strength $\theta$, intervention speed $\phi$ is also an abstraction --- that is, the mechanistic models of interventions make implicit assumptions about the shape of the hazard rate $h$ and therefore $\phi$.''

\revtext{
Presumably both frameworks give a decline when the growth rate is
negative and growth when it is positive. Therefore they would never
contradict each other as to whether an intervention was sufficient or
not - is this correct? This should be made explicit.
}

This is correct, and we have worked to be more explicit. 
\swp{Where? Can we get a quote?}

\revtext{
It would also be good to more explicitly state and discuss what the
differences are. I realize that this is in the discussion and in some
sense in Figure 4, but I found the discussion section vague,
particularly since the analysis is only relevant for the case of
exponential growth. In particular, where would the two frameworks give
different estimates of the uncertainty surrounding the benefits of an
intervention, or the comparison between two interventions?
}

We never expect to get a different final answer from the two formalisms if they are both correctly applied;
they are two ways of doing essentially the same calculation. 
We have tried to be more clear about this as mentioned above. 
We do expect in some cases to be able to use one formalism or another to more clearly illuminate the key issues and quantities determining the detailed answer, as we try to explain in our HIV example and in our new COVID example.

Keeping the two paradigms in mind, however, can help to do the analysis correctly.
In particular, in emerging epidemics, researchers often vary assumptions about parameters while holding \RR\ constant, even in cases where it would be more consistent with available information to hold \rr\ constant.
We now discuss this in the paper:

``Thinking explicitly about the two perspectives can also reduce confusion. Because of the dominance of the strength paradigm, researchers often explore different scenarios while holding \RR\ fixed. Fixed \RR\ is in fact a good default assumption for many endemic diseases. For invading diseases, however, \rr\ is likely to be better constrained by data than \RR. In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level \swp{(i.e., higher proportion of early transmission)} look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value). \swp{I think we can remove the parenthesis in the last sentence}''

\revtext{
On a minor note, since condoms are ultimately an individual choice
applied to individual (potential) transmission events, why are they a
population-level intervention whereas test and treat is deemed
individual? More details on why the two frameworks give different
intuition or analytical capacity would be good.
}

Thank you for this. We were using individual to mean targeted at
infectious individuals, and population to mean targeted at everyone. We
have now dropped these shortcuts, and we simply explain this directly:

``We expect the strength-based framework to be clearer for intervention strategies that target the general population, like condom use, or susceptible people, like prophylaxis.''

We have also added more discussion of how frameworks differ, both in examples and in Discussion.

\revtext{In Figures 3 and 4, the black lines are not clear to me -- the strength
is defined as R/Rhat and the speed as rhat -r . So what is the strength
(or speed) of ``the epidemic'' (black lines) - what's Rhat and what's R?
It might be good to have some prevalence curves}

We define strength and speed of the epidemic in the introduction, and
compare these to strength and speed of intervention efforts. We now
reiterate these points (including in the figure captions).

\revtext{
Minor quibble: equation (4) has L described in words as the ``average
proportional reduction'' but where the reduction is 75\%, in fact the
denominator is 4: $(1 / (1-0.75) )$ -- rephrase.
}

Done, thank you.

\end{document}

### 2020-10-15 speedstrength.tex
some more edits

@@
 We argue that the primacy of $\RR$ over $r$ is partly due to history, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
+\swp{maybe this is where it might be better to say "speed provides the better framing ..."?}
 
@@
 Note that this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to increase under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
+\swp{I feel like there's too much ``something, ...'' phrasing. How about the following?} 
+\swp{
+Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time and assume $K(\tau, t) \equiv K(\tau)$.
+Importantly, this means we are neglecting changes in susceptible proportion through time: $S(t) \approx S(0)$.
+Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
+However, this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to spread under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
+}
+\swp{``conditions characteristic of a naive population'' doesn't really make sense...}
 
@@
 In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, this must exceed the strength of the disease (\RR) to achieve elimination.
+\swp{The last two/three sentences feel repetitive.}
 
@@
 
-The above decomposition generalizes the argument that \RR\ is the key parameter in evaluating whether a disease can be controlled -- one of the main foundations of historical primacy of \RR.
+The above decomposition generalizes the argument that \RR\ is the key parameter in evaluating whether a disease can be controlled --- one of the main foundations of historical primacy of \RR.
 But we can in fact do an analogous decomposition based on speed, and place \rr\ in a similar role.
@@
 Specifically, the speed $\phi$ is a mean of the hazard $h$ in the sense that an increase (or decrease) in $h$ produces the same sign of change in $\phi$, and if $h$ is constant across the generation then $\phi=h$.
-Like intervention strength $\theta$, intervention speed $\phi$ is also an abstraction -- that is, the mechanistic models of interventions make implicit assumptions about the shape of the hazard rate $h$ and therefore $\phi$.
+Like intervention strength $\theta$, intervention speed $\phi$ is also an abstraction --- that is, the mechanistic models of interventions make implicit assumptions about the shape of the hazard rate $h$ and therefore $\phi$.
 
@@
 Though there is a clear intuition for why both strengths increase as early transmission goes down, the speed paradigm provides insight into why these two increases are so close to parallel.
-The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
+The estimated epidemic speed depends only on the observed growth rate --- it does not change if we change our assumption about the proportion of early transmission.
 For the test-and-treat intervention, the effective intervention speed also stays relatively constant (\figref{test}D), in part because we have (plausibly) assumed that hazard stays relatively constant for a few key months, and in part because the backward generation-interval distributions for different scenarios are relatively similar (Supplemental figure).
@@
 
-There has been a great deal of discussion of the importance of pre-symptomatic transmission of COVID-19. 
+There has been a great deal on discussion of the importance of pre-symptomatic transmission of COVID-19 \cite{ferretti2020quantifying, he2020temporal, hellewell2020feasibility}. 
 Pre-symptomatic transmission is likely to be hard to detect, and therefore hard to prevent. 
@@
 A generation-interval perspective \cite{park2019practical} can be used to challenge this view. 
-More early transmission, for a given value of \rr, means {shorter generation intervals and therefore} less transmission per individual -- that is, a lower value of \RR.
+More early transmission, for a given value of \swp{observed} \rr, means {shorter generation intervals and therefore} less transmission per individual --- that is, a lower value of \RR.
 Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary.
-For interventions like lockdowns, which are generally assumed affect everyone roughly equally, the strength-based perspective thus gives a clear (and somewhat surprising) answer: more early transmission means we will conclude that non-targeted interventions are \emph{more} effective.
+For interventions like lockdowns, which are generally assumed \swp{to} affect everyone roughly equally, the strength-based perspective thus gives a clear (and somewhat surprising) answer: more early transmission means we will conclude that non-targeted interventions are \emph{more} effective.
 
@@
 We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while the speed-based paradigm is a clearer way to think about interventions targeted to infected people.
-We concluded that increasing pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions.
+We concluded that \swp{``a higher estimate of'' instead of ``increasing''?}increasing pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions.
 In hindsight, these conclusions are consistent with common sense, but in practice the speed-based framework provides a clear way to think about these questions.
@@
 For invading diseases, however, \rr\ is likely to be better constrained by data than \RR.
-In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value) \citep{eaton2014proportion, powers2014impact, park2019practical}.
+In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level \swp{(i.e., higher proportion of early transmission)} look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value) \citep{eaton2014proportion, powers2014impact, park2019practical}.
+\swp{I think we can remove the parenthesis in the last sentence}
 
-For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that target infected people, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}; we expect the strength-based framework to be clearer for intervention strategies that target the general population, like condom use, or susceptible people, like prophylaxis.
+For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that target infected people, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}.
+We expect the strength-based framework to be clearer for intervention strategies that target the general population, like condom use, or susceptible people, like prophylaxis.
 In other cases, such as real-time rollout of vaccines during an outbreak, both strength and speed approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.
@@
 For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
-\swp{does this feel weak now?
-Maybe need something like ``Many subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing, with a few exceptions'' with citations?}
-\jd{It would be great if you could do this.}
+Subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing \cite{hellewell2020feasibility,kretzschmar2020impact,kucharski2020contact}, with a few exceptions \cite{ferretti2020quantifying}.
+\swp{How is this? Don't want to overwhelm with citations. Seems good enough for me.}
 We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.

### 2020-10-19 response.tex
ready for meeting?

@@
 application section with examples from COVID-19, providing a qualitative
-understanding on how uncertainties in the amount of presymptomatic
+examination of how uncertainties in the amount of presymptomatic
 transmission can affect conclusions about the efficacy of intervention.
-In the beginning of the current pandemic, there was much discussion on
+In the beginning of the current pandemic, there was much discussion of
 how presymptomatic transmission could make intervention more
@@
 our mind, these examples further underline the absence of \rr-based
-thinking in the disease-modeling world. While we appreciate the
-reviewer's concern that a more data-driven approach could provide
-further insight, we are worried that details may take away general ideas
-behind speed- and strength-based paradigms. We feel that it is important
-to provide a conceptual framework at this stage to a broader audience.
-We also address more explicitly the link between the parameters of the
-initial phase and the prospects for long-term control. We feel that
-issues dealing directly with non-exponential disease dynamics are
-complicated enough for another paper.
+thinking in the disease-modeling world.
+
+While we appreciate the reviewer's concern that a more data-driven approach could provide further insight, we wanted to keep the focus on the basic general principles without getting bogged down in the details of specific model or data-fitting assumptions.
+
+We now address more explicitly the link between the parameters of the
+initial phase and the prospects for long-term control. 
+We do not go more deeply here into the subject of non-exponential disease dynamics. 
+We are actively working on this topic, and the issues are quite complicated, so we feel it's best to present these in another paper.
 
@@
 
-This is correct, and we have worked to be more explicit. 
-\swp{Where? Can we get a quote?}
+This is correct, and we have worked to be more explicit, specifically at the end of S2 and the end of S3.
 

### 2020-10-19 responses.tex
ready for meeting?

@@
 \subsection{Referee: 1}\label{referee-1}}
-
-\jd{I moved this to tex, and then figured out that the only equation here is from the reviewer, which makes it awkward to edit.}
 
@@
 paper.
+\swp{Seems like they're repeated. Delete the first one?}
 We have expanded the
@@
 clearly.
-equations?? Or is it OK to leave it like this??
 
@@
 
-\swp{I like this better for individual vs population-based: The efficacy
-of individual-based intervention depends on time since infection (e.g.,
-more likely to be isolated during symptomatic stage). The efficacy of
-population-based intervention doesn't (e.g., the relative efficacy of
-condom doesn't really change). What do you think?}
-\jd{This is exactly the kind of detail I'm avoiding at this point. We can get back to it in the next paper?}
-
 \begin{quote}\sl

### 2020-10-19 speedstrength.tex
ready for meeting?

@@
 Thus, there is a dual sense in which \rr\ also measures difficulty of elimination. 
-We argue that the primacy of $\RR$ over $r$ is partly due to history, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
-\swp{maybe this is where it might be better to say "speed provides the better framing ..."?}
+We argue that the primacy of $\RR$ over $r$ is partly due to history, and that there are cases where speed provides the better framing for practical disease questions than strength (as well as the reverse).
+We provide examples of both situations, for both HIV and COVID-19.
 
@@
 \swp{I feel like there's too much ``something, ...'' phrasing. How about the following?} 
-\swp{
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time and assume $K(\tau, t) \equiv K(\tau)$.
@@
 However, this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to spread under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
-}
 \swp{``conditions characteristic of a naive population'' doesn't really make sense...}
@@
 \end{equation}
-Again, we have that the disease cannot spread when $\theta \geq \RR$. 
+As in the constant-$L$ case above, we have that the disease cannot spread when $\theta \geq \RR$. 
 In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, this must exceed the strength of the disease (\RR) to achieve elimination.
@@
 In other words, \rr, like \RR, is a metric for how hard a disease is to eliminate: we can measure the ``speed'' of a control measure, this must exceed the speed of the disease (\rr) to achieve elimination.
+Since both $\phi \geq r$ and $\theta \geq \RR$ are conditions for control, they are necessarily equivalent: the speed paradigm does not provide a different answer, it provides a different way of thinking about the steps to the correct answer.
 
@@
 
-There has been a great deal on discussion of the importance of pre-symptomatic transmission of COVID-19 \cite{ferretti2020quantifying, he2020temporal, hellewell2020feasibility}. 
+There has been a great deal of discussion of the importance of pre-symptomatic transmission of COVID-19 \cite{ferretti2020quantifying, he2020temporal, hellewell2020feasibility}. 
 Pre-symptomatic transmission is likely to be hard to detect, and therefore hard to prevent. 
@@
 A generation-interval perspective \cite{park2019practical} can be used to challenge this view. 
-More early transmission, for a given value of \swp{observed} \rr, means {shorter generation intervals and therefore} less transmission per individual --- that is, a lower value of \RR.
+More early transmission, for a given value of observed \rr, means {shorter generation intervals and therefore} less transmission per individual --- that is, a lower value of \RR.
 Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary.
-For interventions like lockdowns, which are generally assumed \swp{to} affect everyone roughly equally, the strength-based perspective thus gives a clear (and somewhat surprising) answer: more early transmission means we will conclude that non-targeted interventions are \emph{more} effective.
+For interventions like lockdowns, which are generally assumed to affect everyone roughly equally, the strength-based perspective thus gives a clear (and somewhat surprising) answer: more early transmission means we will conclude that non-targeted interventions are \emph{more} effective.
 
@@
 We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while the speed-based paradigm is a clearer way to think about interventions targeted to infected people.
-We concluded that \swp{``a higher estimate of'' instead of ``increasing''?}increasing pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions.
+We concluded that a higher estimate of pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions.
 In hindsight, these conclusions are consistent with common sense, but in practice the speed-based framework provides a clear way to think about these questions.
@@
 For invading diseases, however, \rr\ is likely to be better constrained by data than \RR.
-In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level \swp{(i.e., higher proportion of early transmission)} look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value) \citep{eaton2014proportion, powers2014impact, park2019practical}.
-\swp{I think we can remove the parenthesis in the last sentence}
+In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level (i.e., higher proportion of early transmission) look relatively more dangerous, because these scenarios will have \rr\ faster than the observed value \citep{eaton2014proportion, powers2014impact, park2019practical}.
 
@@
 Subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing \cite{hellewell2020feasibility,kretzschmar2020impact,kucharski2020contact}, with a few exceptions \cite{ferretti2020quantifying}.
-\swp{How is this? Don't want to overwhelm with citations. Seems good enough for me.}
 We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.

### 2020-10-19 speedstrength.tex
meeting with Daniel

@@
 
-Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time.
-Thus, we will assume $K(\tau, t) \equiv K(\tau)$.
-Importantly, this means we are neglecting changes in susceptible proportion through time: that is, assuming that $S(t) \approx S(0)$.
-Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
-Note that this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to increase under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
-\swp{I feel like there's too much ``something, ...'' phrasing. How about the following?} 
 Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time and assume $K(\tau, t) \equiv K(\tau)$.
@@
 However, this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to spread under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
-\swp{``conditions characteristic of a naive population'' doesn't really make sense...}
 
@@
 In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, this must exceed the strength of the disease (\RR) to achieve elimination.
-\swp{The last two/three sentences feel repetitive.}
 

### 2020-10-19 response.tex
with Daniel

@@
 
-``Thinking explicitly about the two perspectives can also reduce confusion.
+Thinking explicitly about the two perspectives can also reduce confusion.
 Because of the dominance of the strength paradigm, researchers often explore different scenarios while holding \RR\ fixed.
@@
 For invading diseases, however, \rr\ is likely to be better constrained by data than \RR.
-In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level \swp{(i.e., higher proportion of early transmission)} look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value).
-\swp{I think we can remove the parenthesis in the last sentence}''
+In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level (i.e., higher proportion of early transmission) look relatively more dangerous, because these scenarios will have \rr\ faster than the observed value \ldots
 

### 2020-10-20 abstract.pl
statements

@@ -8,6 +8,7 @@ while(<>){
 		next if/^\\/;
 		s/\n/ /g;
 		s/\$//g;
+		s/\s*\\\w+\{[^}]*}//;
 		s/\\//g;
 		s/RR/R/g;
 		s/``/“/g;

### 2020-10-20 abstract.pl
amsync

@@ -1,19 +1,0 @@
-use strict;
-use 5.10.0;
-
-$/ = "";
-
-while(<>){
-	if (/section.*Abstract/ .. /pagebreak/){
-		next if/^\\/;
-		s/\n/ /g;
-		s/\$//g;
-		s/\s*\\\w+\{[^}]*}//;
-		s/\\//g;
-		s/RR/R/g;
-		s/``/“/g;
-		s/''/”/g;
-		s/---?/–/g;
-		say;
-	}
-}

### 2020-10-20 code/HIVexample.R

@@ -1,68 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R")
-commandEnvironments()
-makeGraphics(width=8, height=3)
-
-h_base <- HIVgen(earlyProp = earlyBase, step=0.1)
-
-g1 <- ggplot(h_base) +
-	geom_line(aes(time*month/year, density*year/month)) +
-	scale_x_continuous("Generation time (years)") +
-	scale_y_continuous("Density (per year)", expand=c(0, 0)) +
-	ggtitle("A") +
-	theme(
-		legend.position=c(0.9, 0.9),
-		legend.title=element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-## Prevalence time plot
-g2 <- ggplot(hiv_ts) +
-	geom_point(aes(year, prevalence)) +
-	stat_function(fun=function(x) exp(p0 + rfityear*x)) +
-	scale_x_continuous("Year") +
-	scale_y_continuous("HIV prevalence", limits=c(0, 0.328), expand=c(0, 0)) +
-	ggtitle("B") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-r2R(h_base, rfitmonth)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <-  sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, rfitmonth)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-## R vs. early transmission
-g3 <- ggplot(earlydata) +
-	geom_line(aes(early, R0)) +
-	geom_point(aes(earlyBase, 3.17), size=5) +
-	geom_segment(aes(x=earlyBase, y=0, xend=earlyBase, yend=3.17), lty=2) +
-	geom_segment(aes(x=0, y=3.17, xend=earlyBase, yend=3.17), lty=2) +
-	scale_x_continuous("Proportion of early transmission", expand=c(0, 0)) +
-	scale_y_continuous("Reproduction number", limits=c(0, 5), expand=c(0, 0)) +
-	ggtitle("C") +
-	theme(
-		legend.position="none",
[…]

### 2020-10-20 code/HIVfit.R

@@ -1,14 +1,0 @@
-source("makestuff/makeRfuns.R") 
-
-commandEnvironments()
-
-lastyear <- 8
-hiv_ts <- csvRead(comment="#")
-
-lfit <- lm(log(prevalence)~year, data=hiv_ts[1:lastyear,])
-
-p0 <- coef(lfit)[[1]]
-rfityear <- coef(lfit)[[2]]
-rfitmonth <- coef(lfit)[[2]]*month/year
-
-saveVars(hiv_ts, p0, rfityear, rfitmonth)

### 2020-10-20 code/HIVfuns.R

@@ -1,36 +1,0 @@
-# Units are months
-# Units of g are 1/month, so Sum{step*g) should be 1
-
-earlyFast <- 0.4
-earlyBase <- 0.23
-earlySlow <- 0.1
-
-testingFun <- function(maxRate
-	, hMean=12, hShape=2, hheight=4
-	, step = 0.1, window=200
-){
-	time <- seq(step, window, by=step)
-	tHaz0 <- dgamma(time, shape=hShape, scale=hMean/hShape)
-	tHaz <- hheight*tHaz0/max(tHaz0)
-	hazRate <- maxRate*(1-exp(-tHaz))
-	Haz <- step*cumsum(hazRate)
-	return(data.frame(time
-					  , hazRate
-					  , strength = exp(Haz)
-	))
-}
-
-HIVgen <- function(earlyProp 
-				   , earlyMean=3, earlyShape=3
-				   , lateMean=120, lateShape=2
-				   , step = 0.1, window=200
-){
-	time <- seq(step, window, by=step)
-	d0 <- (
-		earlyProp*dgamma(time, shape=earlyShape, scale=earlyMean/earlyShape)
-		+ (1-earlyProp)*dgamma(time, shape=lateShape, scale=lateMean/lateShape)
-	)
-	return(data.frame(time
-					  , density=d0/sum(step*d0)
-	))
-}

### 2020-10-20 code/condom.R

@@ -1,58 +1,0 @@
-
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R") 
-
-commandEnvironments() ## Read in any environments specified as dependencies
-sourceFiles()
-makeGraphics(width=8, height=3)
-
-Lcondom <- 4
-h_base <- HIVgen(earlyProp = earlyBase, step=0.1)
-
-g1 <- ggplot() +
-	geom_hline(yintercept=Lcondom, col=2) +
-	scale_x_continuous("Time since infection") +
-	scale_y_continuous(expression(L[condom]), limits=c(3, 5)) +
-	ggtitle("A") +
-	theme(
-		panel.grid = element_blank()
-		, panel.border = element_blank()
-		, axis.line.y = element_line()
-		## , axis.line.x = element_line()
-	)
-
-early <- seq(from=0.1, to=0.4, by=0.01)
-
-R0 <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x)
-	r2R(hh, rfitmonth)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-g2 <- ggplot(earlydata) +
-	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_hline(aes(yintercept=Lcondom, col="Intervention")) +
-  geom_point(data=earlydata[earlydata$early==earlyBase,], aes(early, R0), size=5) +
-  geom_point(aes(x=earlyBase, y=Lcondom), size=5, col="red") +
-	scale_color_manual(values=c("black", "red")) +
-	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.4), expand=c(0, 0)) +
-	scale_y_log10("Strength", limits=c(1, 8), expand=c(0, 0), breaks=c(1, 2, 4, 8)) +
-	ggtitle("B") +
-	theme(
-		legend.position=c(0.85, 0.9),
-		legend.title = element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-grid.arrange(g1, g2, nrow=1)
-

### 2020-10-20 code/hivGens.R

@@ -1,75 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R")
-commandEnvironments()
-makeGraphics(width=8, height=3)
-
-legend_order <- c("base", "slow", "fast")
-months = 120
-
-base = HIVgen(earlyProp = earlyBase, window=months)
-fast = HIVgen(earlyProp = earlyFast, window=months)
-slow = HIVgen(earlyProp = earlySlow, window=months)
-
-g <- (
-	list(
-		base = base
-		, fast = fast
-		, slow = slow
-	)
-	%>% bind_rows(.id = "type")
-	%>% mutate(type = factor(type, levels=legend_order))
-)
-
-g <- (
-	list(
-		base = base
-		, fast = fast
-		, slow = slow
-	)
-	%>% bind_rows(.id = "type")
-	%>% mutate(type = factor(type, levels=legend_order))
-)
-
-b <- (
-	list(
-		base = b0fun(base, rfitmonth)
-		, fast = b0fun(fast, rfitmonth)
-		, slow = b0fun(slow, rfitmonth)
-	)
-	%>% bind_rows(.id = "type")
-	%>% mutate(type = factor(type, levels=legend_order))
-)
-
-gplot <- (ggplot(g)
-	+ geom_line(aes(time*month/year, density*year/month, color=type))
-	+ scale_x_continuous("Generation time (years)")
-	+ scale_y_continuous("Density (per year)", expand=c(0, 0))
-	+ ggtitle("Intrinsic generation interval")
-	+ theme(
-		legend.position=c(0.9, 0.9),
-		legend.title=element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-)
-
-bplot <- (ggplot(b)
-	+ geom_line(aes(time*month/year, density*year/month, color=type))
-	+ scale_x_continuous("Generation time (years)")
-	+ scale_y_continuous("Density (per year)", expand=c(0, 0))
-	+ ggtitle("Backward generation interval")
-	+ theme(
-		legend.position=c(0.9, 0.9),
-		legend.title=element_blank(),
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-)
-
-grid.arrange(gplot, bplot, nrow=1)

### 2020-10-20 code/kernel.R

@@ -1,137 +1,0 @@
-# g, b are data frames with time and density
-# L is a data frame with time, strength and something about hazard
-
-## Calculate the initial backward distribution
-b0fun <- function(g, r){
-	step <- with(g, time[[2]]-time[[1]])
-	d0 <- with(g, density*exp(-r*time))
-	return(with(g, data.frame(time
-							  , density=d0/sum(step*d0)
-	)))
-}
-
-# How much does a given discount (r) _reduce_ R for distribution g?
-r2R <- function(g, r){
-	step <- with(g, time[[2]]-time[[1]])
-	return(with(g, 1/(step*sum(density*exp(-r*time)))))
-}
-
-R2r <- function(g, R){
-	step <- with(g, time[[2]]-time[[1]])
-	return(uniroot(
-		function(r, g){r2R(g, r)-R}
-		, interval=c(0, 1/step)
-		, g=g
-	)$root)
-}
-
-### Speed of intervention
-
-# Calculate the arithmetic mean inside the implicit equation for phi
-phiBar <- function(b, L, phi){
-	step <- with(b, time[[2]]-time[[1]])
-	return(with(cbind(b,L), {
-		step*sum(density*exp(phi*time)/strength)
-	}))
-}
-
-### Find the phi that satisfies the equation
-phiFun <- function(b, L){
-	step <- with(b, time[[2]]-time[[1]])
-	return(uniroot(
-		function(phi, b, L){phiBar(b, L, phi)-1}
-		, interval=c(0, 1/step)
-		, b=b, L=L
-	)$root)
-}
-
-theFun <- function(g, L){
-	step <- with(g, time[[2]]-time[[1]])
-	return(with(cbind(g, L), {
-		1/(step*sum(density/strength))
-	}))
-}
-
-densPlot <- function(time, k, scen, color="black", ymax=NULL, title=NULL){
-	if(is.null(ymax)) ymax <- max(k)
-	with(c(scen), {
-		if(is.null(title)) title <- disease
-		if(is.numeric(title)) title <-
-				paste("Reproductive number", sprintf("%5.2f", title))
-		# if(is.numeric(title)) title <- expression(R=title)
-		plot(time, k
-			 , main = title
-			 , xlab = paste0("time (", unit, ")")
[…]

### 2020-10-20 code/tt.R

@@ -1,109 +1,0 @@
-library(tidyr)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("makestuff/makeRfuns.R")
-commandEnvironments()
-makeGraphics(width=8, height=6)
-
-maxRate <- 1/6
-h_base <- HIVgen(earlyProp = earlyBase, step=0.1)
-
-b_base <- b0fun(h_base, rfitmonth)
-
-tt <- testingFun(maxRate)
-
-g1 <- ggplot(tt) +
-	geom_line(aes(time*month/year, strength), col=2) +
-	geom_hline(yintercept=theFun(h_base, tt), lty=2, col=2) +
-	scale_x_continuous("Time (years)", expand=c(0,0), limits=c(0, 16.5)) +
-	scale_y_log10(expression(L[test])) +
-	ggtitle("A") +
-	theme(
-		panel.grid = element_blank(),
-		panel.border = element_blank(),
-		axis.line = element_line()
-	)
-
-early <- seq(from=earlySlow, to=earlyFast, by=0.01)
-
-R0 <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	r2R(hh, rfitmonth)
-})
-
-earlydata <- data.frame(
-	early=early,
-	R0=R0
-)
-
-interstrength <- sapply(early, function(x) {
-	hh <- HIVgen(earlyProp = x, step=0.1)
-	theFun(hh, tt)
-})
-
-strengthdata <- data.frame(
-	early=early,
-	strength=interstrength
-)
-
-g2 <- ggplot(earlydata) +
-	geom_line(aes(early, R0, col="Epidemic")) +
-	geom_line(data=strengthdata, aes(early, strength, col="Intervention")) +
-	scale_color_manual(values=c("black", "red")) +
-  geom_point(data=earlydata[earlydata$early==earlyBase,], aes(early, R0), size=5) +
-  geom_point(data=strengthdata[strengthdata$early==earlyBase,], aes(early, strength), size=5, col="red") +
-	scale_x_continuous("Proportion of early transmission", limits=c(0.1, 0.42), expand=c(0, 0)) +
-	scale_y_log10("Strength", limits=c(1, 16), expand=c(0, 0), breaks=c(1, 2, 4, 8, 16)) +
-	ggtitle("B") +
-	theme(
-		legend.position=c(0.85, 0.9),
-		legend.title = element_blank(),
[…]

### 2020-10-20 code/units.R

@@ -1,11 +1,0 @@
-## Units are retrofitted, and not consistently implemented yet!
-## 2020 Oct 08 (Thu)
-
-source("makestuff/makeRfuns.R") 
-
-year <- 1
-day <- year/365.25
-month <- year/12
-
-saveVars(day, year, month)
-
